### 2024-09-05 manuscript.Rmd
Test

---
title: "Manuscript title here"
output: html_document
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("manuscript.Rmd")` in
the console and you should only see headings -->

<!-- Structure and most comments are from the Strengthening the
reporting of observational studies in epidemiology (STROBE) statement,
see https://www.strobe-statement.org/. For more explanations see the
paper Strengthening the Reporting of Observational Studies in
Epidemiology (STROBE):explanation and elaboration.  -->

<!-- Indicate the study’s design with a commonly used term in the
title, for example: "Associations between resuscitation interventions
and opportunities for improvement in adult trauma patients: A cohort
study" -->

Abstract
========
hej

<!-- Provide in the abstract an informative and balanced summary of
what was done and what was found. Not more than 300 words. -->

Background
----------

Methods
-------

Results
-------

Conclusion
----------

Introduction
============

<!-- Explain the scientific background and rationale for the
investigation being reported. End by stating the aim or specific
objectives, including any prespecified hypotheses. The introduction
should not be longer than 1-1.5 pages, typically 4-5 paragraphs, each
6-8 lines long. Support every statement with a citation -->

You can cite document here like this [@exampleKey9999]. Open the file
bibliography.bib to learn more.

Methods 
=======

Study design
------------

<!-- Present key elements of study design -->

Setting
-------

<!-- Describe the setting, locations, and relevant dates, including
periods of recruitment, exposure, follow-up, and data collection -->

Participants
------------

<!-- Cohort study: Give the eligibility criteria, and the sources and
methods of selection of participants. Describe methods of
follow-up. For matched studies, give matching criteria and number of
exposed and unexposed -->

<!-- Case-control study: Give the eligibility criteria, and the
sources and + methods of case ascertainment and control
selection. Give the rationale for the choice of cases and
controls. For matched studies, give matching criteria and the number
of controls per case -->

<!-- Cross-sectional study: Give the eligibility criteria, and the
sources and methods of selection of participants -->

Variables and data sources/measurements
---------------------------------------

<!-- Clearly define all outcomes, exposures, predictors, potential
confounders, and effect modifiers. Give diagnostic criteria, if
applicable. For each variable of interest, give sources of data and
details of methods of assessment (measurement). Describe comparability
of assessment methods if there is more than one group -->

Bias 
----

<!-- Describe any efforts to address potential sources of bias -->

Study size 
----------

<!-- Explain how the study size was arrived at -->

Quantitative variables
----------------------

<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

Statistical methods
-------------------

<!--

(a) Describe all statistical methods, including those used to control
for confounding

(b) Describe any methods used to examine subgroups and interactions

(c) Explain how missing data were addressed

(d)

Cohort study: If applicable, explain how loss to follow-up was addressed

Case-control study:If applicable, explain how matching of cases and
controls was addressed

Cross-sectional study: If applicable, describe analytical methods
taking account of sampling strategy

(e) Describe any sensitivity analyses
-->

Results
=======

<!--

Participants
------------

(a) Report numbers of individuals at each stage of study—eg numbers
potentially eligible, examined for eligibility, confirmed eligible,
included in the study, completing follow-up, and analysed

(b) Give reasons for non-participation at each stage

(c) Consider use of a flow diagram

Descriptive data
----------------

(a) Give characteristics of study participants (eg demographic,
clinical, social) and information on exposures and potential
confounders

(b) Indicate number of participants with missing data for each
variable of interest

(c) Cohort study — Summarise follow-up time (eg, average and total
amount)

Outcome data
------------

Cohort study — Report numbers of outcome events or summary measures
over time

Case-control study — Report numbers in each exposure category, or
summary measures of exposure

Cross-sectional study — Report numbers of outcome events or summary
measures

Main results
------------

(a) Give unadjusted estimates and, if applicable, confounder-adjusted
estimates and their precision (eg, 95% confidence interval). Make
clear which confounders were adjusted for and why they were included

(b) Report category boundaries when continuous variables were
categorized

(c) If relevant, consider translating estimates of relative risk into
absolute risk for a meaningful time period

Other analyses 
--------------

Report other analyses done—eg analyses of subgroups and interactions,
and sensitivity analyses 
-->

You can include code in this document like this:

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

You can also embed plots:

```{r plot, echo=FALSE}
plot(pressure)
```

You can also mix text and code, so called inline code, like this: `r 2+5`.

Discussion
==========

<!--

Key results
-----------
Summarise key results with reference to study objectives

Limitations
-----------

Discuss limitations of the study, taking into account sources of
potential bias or imprecision.  Discuss both direction and magnitude
of any potential bias

Interpretation
--------------

Give a cautious overall interpretation of results considering
objectives, limitations, multiplicity of analyses, results from
similar studies, and other relevant evidence

Generalisability
----------------

Discuss the generalisability (external validity) of the study results

-->

Conclusion
==========

<!-- Give a short conclusion, corresponding to the aim -->

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2024-09-07 manuscript.Rmd
Test 2
punkt skiss på intro

@@
 ========
-hej
+
+Test
 
@@
 bibliography.bib to learn more.
+
+Trauma is a major global public health concern, especially impacting economically active populations. Effective management of trauma care is essential to improving outcomes for severe injuries.
+
+kolla upp/skriva om i intro
+  - Varför trauma är viktigt
+  - Varför taruma patienter i chock är viktig subdel av akut omhändertagande av taruma pat
+    - Kanske (infoga källa) att chock är stor del av mortalitet.
+  - (PR) Definiton av chock
+  - Klassifikationer av chock nivå
+  - Kliniskt för att förbättra omhändertagande används OFI
+    - Förklara vad OFI är för något
+  - Korrelation mellan chock nivåer och OFI
+    - Var kommer specifika typer av OFI in??
 

### 2024-09-07 manuscript.Rmd
Lagt till klassifikations förslag, ska läsa vidare

@@
   - Klassifikationer av chock nivå
+    - ATLS klassifikation för blödningschock.
+      - Obstruktiv som tamponad?
+      - Neurogen som trauma mot hlasrygg?
+    - RETTS som klassifikation?
+    - NEWS som klassifikation?
   - Kliniskt för att förbättra omhändertagande används OFI

### 2024-09-07 manuscript.Rmd
Förtydligat frågorna som behöver ställas

--- 2024-09-05 manuscript.Rmd
@@
 bibliography.bib to learn more.
+
+Trauma is a major global public health concern, especially impacting economically active populations. Effective management of trauma care is essential to improving outcomes for severe injuries.
+
+kolla upp/skriva om i intro
+  - Varför trauma är viktigt
+    - Mortalitet för unga
+  - Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
+    - Kanske (infoga källa) att chock är stor del av mortalitet.
+  - Definiton av chock
+    - olika typer av chock: hypovolem, obstruktiv, neurogen, kardiogen och distributiv.
+    - trauma oftas de 3 första??
+  - Klassifikationer av chock nivå
+    - ATLS klassifikation för blödningschock.
+      - Hur ska jag ta med:
+        - Obstruktiv som tamponad?
+        - Neurogen som trauma mot halsrygg?
+      - RETTS och NEWS istället?? de används vid triage men är inte chock specifik.
+    - Kanske borde använda ATLS 4 kategorier + obstruktiv + neurogen vid sidan??
+  - Kliniskt för att förbättra omhändertagande används OFI
+    - Förklara vad OFI är för något
+  - Korrelation mellan chock nivåer och OFI (Är det då antal OFI/chock nivåer som anses?)
+    - Var kommer specifika typer av OFI in?? Finns denna sorts data?
+     - Om det finns kan man då även lägga till i frågeställningen att man jämför hur OFI subkategorier skiljer sig mellan de olika chock nivåerna.
 

### 2024-09-08 manuscript.Rmd
Läst metod och fatta bättre, skrev av ska fundera och förstå

--- 2024-09-07 manuscript.Rmd
@@
 kolla upp/skriva om i intro
-  - Varför trauma är viktigt
-  - Varför taruma patienter i chock är viktig subdel av akut omhändertagande av taruma pat
-    - Kanske (infoga källa) att chock är stor del av mortalitet.
-  - (PR) Definiton av chock
-  - Klassifikationer av chock nivå
-  - Kliniskt för att förbättra omhändertagande används OFI
-    - Förklara vad OFI är för något
-  - Korrelation mellan chock nivåer och OFI
-    - Var kommer specifika typer av OFI in??
+  Varför trauma är viktigt
+    Mortalitet för unga
+  Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
+    Kanske (infoga källa) att chock är stor del av mortalitet.
+  Definiton av chock
+    olika typer av chock: hypovolem, obstruktiv, neurogen, kardiogen och distributiv.
+    trauma oftas de 3 första??
+  Klassifikationer av chock nivå
+    ATLS klassifikation för blödningschock.
+      Hur ska jag ta med:
+        Obstruktiv som tamponad?
+        Neurogen som trauma mot halsrygg?
+      RETTS och NEWS istället?? de används vid triage men är inte chock specifik.
+    Kanske borde använda ATLS 4 kategorier + obstruktiv + neurogen vid sidan??
+  Kliniskt för att förbättra omhändertagande används OFI
+    Förklara vad OFI är för något
+  Syfte: Korrelation mellan chock nivåer och OFI (Är det då antal OFI/chock nivåer som anses?)
+    Vilka typer av OFI finns hos shockpatienter?
+    Koppling mellan nivå av chock och sannolikhet till OFI.
+    \textit{Var kommer specifika typer av OFI in?? Finns denna sorts data?
+     Om det finns kan man då även lägga till i frågeställningen att man jämför hur OFI subkategorier skiljer sig mellan de olika chock nivåerna.}
 

### 2024-09-08 manuscript.Rmd
sortera

@@
 
-kolla upp/skriva om i intro
-  Varför trauma är viktigt
+1  Varför trauma är viktigt
     Mortalitet för unga
@@
       Kanske (infoga källa) att chock är stor del av mortalitet.
-  Definiton av chock
+2  Definiton av chock
     olika typer av chock: hypovolem, obstruktiv, neurogen, kardiogen och distributiv.
     trauma oftas de 3 första??
-  Klassifikationer av chock nivå
+3  Klassifikationer av chock
     ATLS klassifikation för blödningschock.
@@
     Kanske borde använda ATLS 4 kategorier + obstruktiv + neurogen vid sidan??
-  Kliniskt för att förbättra omhändertagande används OFI
+4  Kliniskt för att förbättra omhändertagande används OFI
     Förklara vad OFI är för något
-  Syfte: Korrelation mellan chock nivåer och OFI (Är det då antal OFI/chock nivåer som anses?)
+5  Syfte: Korrelation mellan chock nivåer och OFI (Är det då antal OFI/chock nivåer som anses?)
     Vilka typer av OFI finns hos shockpatienter?

### 2024-09-08 manuscript.Rmd
pico?

@@
     Förklara vad OFI är för något
-5  Syfte: Korrelation mellan chock nivåer och OFI (Är det då antal OFI/chock nivåer som anses?)
-    Vilka typer av OFI finns hos shockpatienter?
-    Koppling mellan nivå av chock och sannolikhet till OFI.
-    \textit{Var kommer specifika typer av OFI in?? Finns denna sorts data?
-     Om det finns kan man då även lägga till i frågeställningen att man jämför hur OFI subkategorier skiljer sig mellan de olika chock nivåerna.}
+5  Frågeställning:
+    a, Vilka typer av OFI finns hos chockpatienter?
+    b, Koppling mellan nivå av chock och sannolikhet till OFI.
+    Vilken av de två är huvudfrågeställningen?
+    PICO: Population: Chock pat, bland 8000 (2014-2023)
+          Intervention: OFI?
+          Comparison: mellan chocknivåer?
+          Outcome: Sannolikhet till OFI?
+      Det blir ju inte direkt någon ja/nej svar på frågeställningen...funkar det?
 

### 2024-09-08 manuscript.Rmd
spara

@@
 
-Trauma is a major global public health concern, especially impacting economically active populations. Effective management of trauma care is essential to improving outcomes for severe injuries.
-
 1  Varför trauma är viktigt
@@
       Kanske (infoga källa) att chock är stor del av mortalitet.
+
+Trauma is a major global public health concern in form of death and disability.
+It affects over four milion peole each year, among those younger and economicaly active(??) populations are the most affected.
+
 2  Definiton av chock

### 2024-09-08 manuscript.Rmd
Lagt till text under styckena

@@
 Trauma is a major global public health concern in form of death and disability.
-It affects over four milion peole each year, among those younger and economicaly active(??) populations are the most affected.
+It affects over four milion peole each year, among those younger and economicaly active(??) populations are the most affected(källa).
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone(källa), especially in patients arriving in shock(källa).
+Therefore, careful studying of this part is of importance.
 
@@
     trauma oftas de 3 första??
+
+Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[].
+There are five types of causes, which are hypovolemia, obstructive, neurogenic, cardiogenic and distributive.
+These causes can occur alone or in combination.
+The most usuall one during trauma are hypovolemia in form of a large bleeding.
+
 3  Klassifikationer av chock
@@
     Kanske borde använda ATLS 4 kategorier + obstruktiv + neurogen vid sidan??
+
+Clinically to be able to differentiate the degree of shock, usually the ATLS classification for major haemorrhage are used.
+The classification are divided into four classes, based on the patients: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output.
+However this only suits hypovolemic shocks and not the other types.
+
 4  Kliniskt för att förbättra omhändertagande används OFI
     Förklara vad OFI är för något
+
+To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
+OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the straucture and clinical process of trauma care(Hussein).
+
 5  Frågeställning:
@@
       Det blir ju inte direkt någon ja/nej svar på frågeställningen...funkar det?
+
+This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
 

### 2024-09-09 manuscript.Rmd
skiss metod 1

@@
     PICO: Population: Chock pat, bland 8000 (2014-2023)
-          Intervention: OFI?
+          Intervention: OFI efter vård?
           Comparison: mellan chocknivåer?
@@
 
+Registry based cohort study, using data from the trauma registry and trauma care database at the Karolinska University Hospital in Solna.
+
+This project will link the two databases and assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
+A 5% significance level and 95% confidence levels will be used.
+
 Setting
@@
 periods of recruitment, exposure, follow-up, and data collection -->
+
+The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2012 and 2023.
+Karolinska University Hospital in Solna acts as a level one trauma center according to critera set by American College of Surgeons(), and treats all major trauma in the greater metropolitan area of Stockholm.
+The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023.
+
+The outcome is opportunities for improvement, such as delayed management or technical errors, as identified by a multidisciplinary trauma quality improvement review board and recorded in the trauma care quality database.
 
@@
 sources and methods of selection of participants -->
+
+Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
+Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
+Examples of OFIs identified in this process may include the need for more senior trauma team members or additional emergency operating rooms.
+These issues are then categorized into broader areas, such as resource or logistical deficiencies.
+The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields.
+The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
 

### 2024-09-09 manuscript.Rmd
Sorterat till det, design och setting någorlunda nöjd. Participant blev kosntig delad i två.

@@
 To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
-OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the straucture and clinical process of trauma care(Hussein).
+OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the stracture and clinical process of trauma care(Hussein).
 
@@
 
-Registry based cohort study, using data from the trauma registry and trauma care database at the Karolinska University Hospital in Solna.
-
-This project will link the two databases and assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
-A 5% significance level and 95% confidence levels will be used.
+We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
 
@@
 The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2012 and 2023.
+The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023.
 Karolinska University Hospital in Solna acts as a level one trauma center according to critera set by American College of Surgeons(), and treats all major trauma in the greater metropolitan area of Stockholm.
-The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023.
-
-The outcome is opportunities for improvement, such as delayed management or technical errors, as identified by a multidisciplinary trauma quality improvement review board and recorded in the trauma care quality database.
 
@@
 Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
-Examples of OFIs identified in this process may include the need for more senior trauma team members or additional emergency operating rooms.
-These issues are then categorized into broader areas, such as resource or logistical deficiencies.
+The identified OFIs are then categorized into broader areas.
 The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields.
 The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
+
+To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
+We included all patients in ...
+between ...
+and ...
+We excluded patients younger than 18 and/or were dead on arrival.
 
@@
 of assessment methods if there is more than one group -->
+
+The outcome is OFI, such as delayed management or technical errors, as identified by a multidisciplinary trauma quality improvement review board and recorded in the trauma care quality database.
 

### 2024-09-09 manuscript.Rmd
Sparar inför natten

@@
 
-The outcome is OFI, such as delayed management or technical errors, as identified by a multidisciplinary trauma quality improvement review board and recorded in the trauma care quality database.
+The outcome is OFI recorded in the trauma care quality database and are binary (as in yes exist or no does not exist) when used to calculate the association with degree of shock. 
+The subgroups of OFI will be explored and reported in a list, and therefore non-binary.
+
+To classify the patients the following variables will be used: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output.
 
@@
 <!-- Explain how the study size was arrived at -->
+
+Size of the data.
 
@@
 -->
+
+(a) Logistic regression to determine the association between the binary results of OFI existence and non-binary degree of shock.
+
+(b) ?
+
+(c) patient case removed?
+
+(d) patient case removed?
+
+(e) ?
 

### 2024-09-10 manuscript.Rmd
råka fixa till referenserna
tog tillbaka det jag råka tabort

@@
 title: "Manuscript title here"
-output: html_document
 author: "Your name here"
+output:
+  word_document: default
+  html_document: default
 bibliography: bibliography.bib
@@
     Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
-      Kanske (infoga källa) att chock är stor del av mortalitet.
+      Kanske att chock är stor del av mortalitet.
 
 Trauma is a major global public health concern in form of death and disability.
-It affects over four milion peole each year, among those younger and economicaly active(??) populations are the most affected(källa).
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone(källa), especially in patients arriving in shock(källa).
-Therefore, careful studying of this part is of importance.
+It affects over four milion peole each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010].
+This combined with the fact that trauma patients in shock have a higher mortality rate, than those not[@vang_shock_2022], makes it important to study and improve the care for trauma patients in shock.
 
@@
 
-Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[].
+Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
 There are five types of causes, which are hypovolemia, obstructive, neurogenic, cardiogenic and distributive.
 These causes can occur alone or in combination.
-The most usuall one during trauma are hypovolemia in form of a large bleeding.
+The most usuall one during trauma are hypovolemia in form of a large bleeding[@teixeira_preventable_2007].
 
@@
 
-Clinically to be able to differentiate the degree of shock, usually the ATLS classification for major haemorrhage are used.
+Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used.
 The classification are divided into four classes, based on the patients: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output.
@@
 
-4  Kliniskt för att förbättra omhändertagande används OFI
+Skriva om stycke 4 och istället lägga in innehåll om använda retts som en fortsättnig på stycke 3.
+
+  (4  Kliniskt för att förbättra omhändertagande används OFI
     Förklara vad OFI är för något
@@
   To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
-OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the stracture and clinical process of trauma care(Hussein).
+OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the structure and clinical process of trauma care(Hussein).)
 
@@
 
-```{r main, echo=FALSE}
-source("main.R") ## This "imports" the main script file of your project and run any code in it
-```
-
 You can also embed plots:

### 2024-09-11 manuscript.Rmd
sparar för natten
Lagt till retts och news i stycke 3, men kasnke ska flytta delar till metod?

@@
 
-Test
-
 <!-- Provide in the abstract an informative and balanced summary of
@@
 
-Trauma is a major global public health concern in form of death and disability.
-It affects over four milion peole each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010].
-This combined with the fact that trauma patients in shock have a higher mortality rate, than those not[@vang_shock_2022], makes it important to study and improve the care for trauma patients in shock.
-
 2  Definiton av chock
@@
     trauma oftas de 3 första??
-
-Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
-There are five types of causes, which are hypovolemia, obstructive, neurogenic, cardiogenic and distributive.
-These causes can occur alone or in combination.
-The most usuall one during trauma are hypovolemia in form of a large bleeding[@teixeira_preventable_2007].
 
@@
 
-Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used.
-The classification are divided into four classes, based on the patients: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output.
-However this only suits hypovolemic shocks and not the other types.
-
-Skriva om stycke 4 och istället lägga in innehåll om använda retts som en fortsättnig på stycke 3.
-
   (4  Kliniskt för att förbättra omhändertagande används OFI
     Förklara vad OFI är för något
-
       To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
@@
       Det blir ju inte direkt någon ja/nej svar på frågeställningen...funkar det?
+
+Trauma is a major global public health concern in form of death and disability.
+It affects over four milion peole each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010].
+This combined with the fact that trauma patients in shock have a higher mortality rate, than those not[@vang_shock_2022], makes it important to study and improve the care for trauma patients in shock.
+
+Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
+There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive.
+These causes can occur alone or in combination.
+The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
+
+Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used.
+The classification are divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
+However this only suits hypovolemic shocks and not the other types.
+There are two other classifications, namely RETTS and NEWS2.
+These are used in emergency triaging and can be used on other types of shocks, however they are not shock specific classifications.
+If we look into all three classifications we will be able to find similarities in that all use HF, BP, RR and mental status.
+Therefore we decided to use these markers as the independent variables for classification of shock regardless of their nature.
+(sista två meningarna borde tillhöra metod...)
 
@@
 
-(a) Logistic regression to determine the association between the binary results of OFI existence and non-binary degree of shock.
+(a) Logistic regression to determine the association between the binary results of OFI existence and teh four degrees of shock.
 
@@
 
+```{r main, echo=FALSE} 
+source("main.R") ## This "imports" the main script file of your project and run any code in it 
+```
+
 You can also embed plots:

### 2024-09-11 manuscript.Rmd
test

@@
       To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
-OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the structure and clinical process of trauma care(Hussein).)
+OFI is multidiciplinary morbidity and mortality reviews, where opportunity for improvements are identified in the structure and clinical process of trauma care[@albaaj_patient_2023].)
 

### 2024-09-11 project-plan.Rmd
Ordbajs varianten av intro, saknar OFI.

---
title: "Manuscript title here"
subtitle: "Add a Subtitle if Needed"
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- The project plan should be 3-4 pages long and written in
Rmarkdown (like this document). Remove all comments in italic when you 
use this document as a template for your own project plan. -->

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
the console and you should only see headings -->

Introduction
============

*This is part A) in the project plan [sv: projektredogörelse]. *

*The background/introduction should summarise the scope of the problem,
what is known about the problem, what is not known, what specific
knowledge gap the study is supposed to fill and why this is
important. It should end be stating the aim of the study. The
background should be 4-5 paragraphs long and each paragraph should be
between 3-5 sentences. Strive to make the paragraphs about the same
length.*

*bibliography.bib includes an example reference. Add additional
references in that file as bibtex entries and then cite in text using:
"[@exampleKey9999]". One can convert DOIs' to bibtex entries using 
https://www.doi2bib.org/. If you use Zotero you can simply export 
entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*

Trauma is a significant global public health concern, responsible for causing death and disability in over four million people each year. Among those affected, younger populations are disproportionately impacted, making trauma management a critical issue in healthcare systems worldwide[@noauthor_injuries_nodate]. The initial management of trauma patients in the hospital setting is extremely time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010]. Trauma patients presenting in shock have a notably higher mortality rate compared to those who are not in shock, underscoring the need for improved care strategies for these high-risk patients[@vang_shock_2022].
Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023]. There are four primary causes of shock: hypovolemic, obstructive, cardiogenic, and distributive. These etiologies can occur in isolation or in combination. In the context of trauma, hypovolemic shock, usually caused by massive hemorrhage, is the most common type of shock[@teixeira_preventable_2007].

To classify shock in trauma 
The Advanced Trauma Life Support (ATLS) system provides a widely used classification for hypovolemic shock based on blood loss, heart rate, blood pressure, respiratory rate, mental status, and urine output. ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe, and helps guide resuscitative efforts. However, this classification is specific to hypovolemic shock and does not account for the other types of shock.

To address the broader spectrum of shock types encountered in trauma, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are commonly used in emergency care settings. While these systems are not shock-specific, they offer useful frameworks for assessing patients in various shock states by prioritizing vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR), and mental status.

RETTS is designed for rapid triage and treatment prioritization in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency. It incorporates physiological parameters similar to those used in ATLS, making it applicable to patients in various forms of shock, including hypovolemic, cardiogenic, and distributive shock.

NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of hospitalized patients. Like ATLS and RETTS, it evaluates HR, BP, RR, and mental status, but also includes oxygen saturation and temperature. NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock.

Despite their differences, all three systems—ATLS, RETTS, and NEWS2—share key similarities, particularly in their use of vital sign markers such as heart rate, blood pressure, respiratory rate, and mental status. This overlap suggests a potential framework for integrating these systems to better classify trauma patients presenting in any form of shock.

Infoga OFI

This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.

Methods
=======

*Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*

Study design
We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.

Setting
The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2012 and 2023. The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023. Karolinska University Hospital in Solna acts as a level one trauma center according to critera set by American College of Surgeons(), and treats all major trauma in the greater metropolitan area of Stockholm.

Participatns
Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters. Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences. The identified OFIs are then categorized into broader areas. The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields. The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.

To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine. We included all patients in ... between ... and ... We excluded patients younger than 18 and/or were dead on arrival.

Variable and measurements
To classify the patients "heart rate", "blood pressure", "respiratory rate" and "mental status" will be used as independent variables for classification of shock regardless of their cause.  Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications. The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock. The subgroups of OFI will be explored and reported in a list, and therefore non-binary.

Project Update and Timeline
===========================

*This is part B) in the project plan [sv: projektredogörelse]. *

*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*

![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)

Doing right now:
- Introduction
- Methods

To do:
- Coding for regression
- Analysis
- Results
- Discussion
- Conclusion and Abstract

Back-up Plan
============

*This is part C) in the project plan [sv: projektredogörelse]*

References
==========

<!-- Do not edit by hand, references will be inserted and formatted automatically once you knit this document -->

### 2024-09-11 manuscript.Rmd
lagt till ofi i pr. typ färdig med pr.

--- 2024-09-11 manuscript.Rmd
@@
     Förklara vad OFI är för något
-      To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
-OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the structure and clinical process of trauma care(Hussein).)
+      To be able to assess and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
+OFI is multidiciplinary morbidity and mortality reviews, where opportunity for improvements are identified in the structure and clinical process of trauma care[@albaaj_patient_2023].
+)
 
@@
 If we look into all three classifications we will be able to find similarities in that all use HF, BP, RR and mental status.
-Therefore we decided to use these markers as the independent variables for classification of shock regardless of their nature.
 (sista två meningarna borde tillhöra metod...)
@@
 
-The outcome is OFI recorded in the trauma care quality database and are binary (as in yes exist or no does not exist) when used to calculate the association with degree of shock. 
+To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
+Heart rate, bloodpressure, respiratory rate and mental status.
+Based on these four measurements, the patients will be calssified into four groups rouphly based on the same value as in the ATLS trauma shock classifications.
+The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
 The subgroups of OFI will be explored and reported in a list, and therefore non-binary.
-
-To classify the patients the following variables will be used: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output.
 

### 2024-09-11 project-plan.Rmd
lagt till ofi i pr. typ färdig med pr.

@@
 Trauma is a significant global public health concern, responsible for causing death and disability in over four million people each year.
-Among those affected, younger populations are disproportionately impacted, making trauma management a critical issue in healthcare systems worldwide[@noauthor_injuries_nodate].
-The initial management of trauma patients in the hospital setting is extremely time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010].
-Trauma patients presenting in shock have a notably higher mortality rate compared to those who are not in shock, underscoring the need for improved care strategies for these high-risk patients[@vang_shock_2022].
+Among those affected, younger populations are more impacted, making trauma management a worldwide health care concern[@noauthor_injuries_nodate].
+The initial management of trauma patients in the hospital setting is time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010].
+Trauma patients presenting with shock have a notably higher mortality compared to those who are not in shock, underscoring the need for improved care strategies for these patients[@vang_shock_2022].
 Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023].
-There are four primary causes of shock: hypovolemic, obstructive, cardiogenic, and distributive.
-These etiologies can occur in isolation or in combination.
+There are four primary causes of shock: hypovolemic, obstructive, cardiogenic and distributive.
+These etiologies can occur alone or in combination.
 In the context of trauma, hypovolemic shock, usually caused by massive hemorrhage, is the most common type of shock[@teixeira_preventable_2007].
@@
 To classify shock in trauma 
-The Advanced Trauma Life Support (ATLS) system provides a widely used classification for hypovolemic shock based on blood loss, heart rate, blood pressure, respiratory rate, mental status, and urine output.
-ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe, and helps guide resuscitative efforts.
-However, this classification is specific to hypovolemic shock and does not account for the other types of shock.
+The Advanced Trauma Life Support (ATLS) classification for hypovolemic shock based on the parameters of "blood loss", "heart rate", "blood pressure", "respiratory rate", "mental status" and "urine output".
+ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe.
+However, this classification is specific to hypovolemic shock and does not account for the other types of shock like obstructive.
 
-To address the broader spectrum of shock types encountered in trauma, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are commonly used in emergency care settings.
-While these systems are not shock-specific, they offer useful frameworks for assessing patients in various shock states by prioritizing vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR), and mental status.
+To address other types of diseases, such as other types of trauma shock, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are used in emergency care.
+While these systems are not shock-specific, they are useful for assessing patients in various shock states by using vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR) and mental status.
 
-RETTS is designed for rapid triage and treatment prioritization in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency.
-It incorporates physiological parameters similar to those used in ATLS, making it applicable to patients in various forms of shock, including hypovolemic, cardiogenic, and distributive shock.
+RETTS is designed for triage in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency.
+It incorporates parameters similar to those used in ATLS, but also include airway, saturation and temperature.
+This makes it applicable to patients in various forms of shock, including hypovolemic, cardiogenic and distributive shock(källa?).
 
-NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of hospitalized patients.
+NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of patients.
 Like ATLS and RETTS, it evaluates HR, BP, RR, and mental status, but also includes oxygen saturation and temperature.
-NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock.
+NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock(källa?).
 
-Despite their differences, all three systems—ATLS, RETTS, and NEWS2—share key similarities, particularly in their use of vital sign markers such as heart rate, blood pressure, respiratory rate, and mental status.
-This overlap suggests a potential framework for integrating these systems to better classify trauma patients presenting in any form of shock.
+Despite their differences, all three systems ATLS, RETTS and NEWS2 share similarities in their use of vital sign markers such as heart rate, blood pressure, respiratory rate and mental status.
+This overlap suggests using these four parameters as a base for general classification for all types of shock is justifiable(är det verkligen så, får man anta det?).
 
-Infoga OFI
+To improve trauma care quality, healthcare systems identify Opportunities for Improvement (OFI) through structured reviews of patient cases.
+Multidisciplinary morbidity and mortality reviews play a central role in this process, where clinicians from different specialities together analyse trauma cases to evaluate both the structure and clinical processes involved in trauma care following the Donabedian’s quality of care framework[@albaaj_patient_2023].
+For shock patients, timely interventions such as fluid resuscitation/blood transfusion, intervention, OR and radiology are critical, and delays or missteps can result in adverse outcomes (källa).
+By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity(källa).
 

### 2024-09-14 manuscript.Rmd
Comment on manuscript

@@
 Trauma is a major global public health concern in form of death and disability.
-It affects over four milion peole each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010].
+It affects over four milion peole each year, among those younger populations are the most affected <!-- You can be explicit and say that trauma causes over four million deaths each year.
+Trauma affects hundreds of millions -->[@noauthor_injuries_nodate].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010].
 This combined with the fact that trauma patients in shock have a higher mortality rate, than those not[@vang_shock_2022], makes it important to study and improve the care for trauma patients in shock.
+<!-- You can write that shock is a major cause of mortality in trauma patients.
+-->
 
@@
 The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
+<!-- Move the information about the types of shock here.
+-->
 
@@
 However this only suits hypovolemic shocks and not the other types.
-There are two other classifications, namely RETTS and NEWS2.
+There are two other classifications, namely RETTS and NEWS2<!-- You can discuss these in your KI paper but you don't need it here in your manuscript-->.
 These are used in emergency triaging and can be used on other types of shocks, however they are not shock specific classifications.
@@
 (sista två meningarna borde tillhöra metod...)
+
+<!-- Say something about opportunities for improvement here, and why it's important to study these among patients in shock -->
 
@@
 We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
+We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
 
@@
 
-The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2012 and 2023.
-The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023.
-Karolinska University Hospital in Solna acts as a level one trauma center according to critera set by American College of Surgeons(), and treats all major trauma in the greater metropolitan area of Stockholm.
+The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
+The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review.
+The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.
 
@@
 We included all patients in ...
-between ...
-and ...
-We excluded patients younger than 18 and/or were dead on arrival.
+between 2014 and 2023  We excluded patients younger than 15 and/or were dead on arrival.
 
@@
 
+<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
+
 To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
 Heart rate, bloodpressure, respiratory rate and mental status.
-Based on these four measurements, the patients will be calssified into four groups rouphly based on the same value as in the ATLS trauma shock classifications.
-The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
+Based on these four measurements, the patients will be classified into four groups rouphly based on the same value as in the ATLS trauma shock classifications.
+<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?--> The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
 The subgroups of OFI will be explored and reported in a list, and therefore non-binary.
@@
 
-Size of the data.
+Size of the data.<!-- Mention that you will include all available data-->
 
@@
 -->
+
+<!-- Start with stating what software you'll use. Then say something about how you will describe the data. Then say something about how you will analyse the data. -->
 

### 2024-09-14 manuscript.Rmd
nått ord här och där, inga större skillnad

--- 2024-09-11 manuscript.Rmd
@@
 
+Infoga OFI del och koppla samman
+
 This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
@@
 We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
+We then linked the two databases to first assess what types of OFI occurred in patients arriving in shock and then use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
 

### 2024-09-16 manuscript.Rmd
Fortsättning på metod, väntar på svar hur man använder ATLS kliniskt.

--- 2024-09-14 manuscript.Rmd
@@
 To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
-Heart rate, bloodpressure, respiratory rate and mental status.
-Based on these four measurements, the patients will be classified into four groups rouphly based on the same value as in the ATLS trauma shock classifications.
-<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?--> The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
+Heart rate(HR), blood pressure(BP), respiratory rate(RR) and mental status.
+Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.
+
+<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->
+The first group will be patients presenting with a HR below 100 beats per minute(BPM), respiratory rate between 14 to 20.
+The original ATLS classification only states normal BP, but no number.
+In our case, we will be using the systolic over 90mmhg as the lower limit up to 65 years old, and 110mmhg as the lower limit for patients older (geriatric patients).
+The...
+
+The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
 The subgroups of OFI will be explored and reported in a list, and therefore non-binary.
@@
 
-Size of the data.<!-- Mention that you will include all available data-->
+All available data in the trauma care quality database? will be included, regarding the studied variables. <!-- Mention that you will include all available data-->
 
@@
 <!-- Start with stating what software you'll use. Then say something about how you will describe the data. Then say something about how you will analyse the data. -->
+
+I will be using R to...
 

### 2024-09-16 project-plan.Rmd
rättade till referens platsen

@@
 The initial management of trauma patients in the hospital setting is time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010].
-Trauma patients presenting with shock have a notably higher mortality compared to those who are not in shock, underscoring the need for improved care strategies for these patients[@vang_shock_2022].
+Trauma patients presenting with shock have a notably higher mortality compared to those who are not in shock[@vang_shock_2022], underscoring the need for improved care strategies for these patients.
 Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023].

### 2024-09-16 manuscript.Rmd
Uppdaterad intro, med tillagd OFI del från projektplan samt även ändrat strukturen från innan och ändrat enligt kommentarerna

@@
 
-Trauma is a major global public health concern in form of death and disability.
-It affects over four milion peole each year, among those younger populations are the most affected <!-- You can be explicit and say that trauma causes over four million deaths each year.
-Trauma affects hundreds of millions -->[@noauthor_injuries_nodate].
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010].
-This combined with the fact that trauma patients in shock have a higher mortality rate, than those not[@vang_shock_2022], makes it important to study and improve the care for trauma patients in shock.
-<!-- You can write that shock is a major cause of mortality in trauma patients.
--->
-
-Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
+Trauma is a major global public health concern.
+It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
+Shock is a major cause of mortality in trauma patients.
+The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
 There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive.
@@
 The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
-<!-- Move the information about the types of shock here.
--->
 
-Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used.
-The classification are divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
-However this only suits hypovolemic shocks and not the other types.
-There are two other classifications, namely RETTS and NEWS2<!-- You can discuss these in your KI paper but you don't need it here in your manuscript-->.
-These are used in emergency triaging and can be used on other types of shocks, however they are not shock specific classifications.
-If we look into all three classifications we will be able to find similarities in that all use HF, BP, RR and mental status.
-(sista två meningarna borde tillhöra metod...)
+Clinically to be able to differentiate the degree of shock in context of trauma, the ATLS classification for major haemorrhage can be used.
+The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
+However, this only suits hypovolemic shocks and not the other types.
+(Källa) There is other classification for triaging or tracking the clinical deterioration of patients, which can be used for other causes of shocks.
+The common denominator for these classifications is they are all based on vital parameters.
 
-<!-- Say something about opportunities for improvement here, and why it's important to study these among patients in shock -->
-
-Infoga OFI del och koppla samman
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010].
+To improve trauma care quality, healthcare systems identify Opportunities for Improvement (OFI) through structured reviews of patient cases.
+Multidisciplinary morbidity and mortality reviews play a central role in this process, where clinicians from different specialities together analyse trauma cases to evaluate both the structure and clinical processes involved in trauma care following the Donabedian’s quality of care framework[@albaaj_patient_2023].
+For shock patients, timely interventions such as fluid resuscitation/blood transfusion, intervention, OR and radiology are critical, and delays or missteps can result in adverse outcomes (källa).
+By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity(källa).
 

### 2024-09-16 manuscript.Rmd
bättre ord
tog bort en ställe som krävde källa, för att det inte krävs

@@
 These causes can occur alone or in combination.
-The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
+The most usual one during trauma are hypovolemic in form of a major haemorrhage [@teixeira_preventable_2007].
 
@@
 For shock patients, timely interventions such as fluid resuscitation/blood transfusion, intervention, OR and radiology are critical, and delays or missteps can result in adverse outcomes (källa).
-By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity(källa).
+By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity.
 

### 2024-09-16 manuscript.Rmd
la till tanke idé för ofi och chock koppling. nu äntligen fattar jag typ vad som ska vara med i intro

@@
 Trauma is a major global public health concern.
-It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
-Shock is a major cause of mortality in trauma patients.
+It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected.
+Not only does it do bodily harm, but also places a burden on national economies, in form of billions of us dollars for health care, lost productivity and law enforcement[@noauthor_injuries_nodate].
+
+Shock is a major cause of mortality in trauma patients[@siegel_effect_1995].
 The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
@@
 The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010].
+
+This is especially worth noting when patients in shock also have a higher mortality rate than those not[@vang_shock_2022].
+
+Koppla ihop det mpå nått sätt med att, trauma i sig är tid sensitiv och kaotisk, kombinerat med att chock patienter har högre mortalitet (kanske källa som även säger chock pat är mer kaotisk och tid sensitiv), parallellt har pat med medel+hög iss hög OFI, dessa medel och hög har oftast chock(källa), därför kombinerat kaos natur + kaos från chock + medel skadade oftast med chock har hög ofi -> bör undersöka pat med olika nivåer av chock har ofi.
+
 To improve trauma care quality, healthcare systems identify Opportunities for Improvement (OFI) through structured reviews of patient cases.

### 2024-09-17 manuscript.Rmd
la till 2 källor

@@
 
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008].
 

### 2024-09-20 manuscript.Rmd
Version 2.2 för manus och 2.0 för projekt redogörelse.

--- 2024-09-16 manuscript.Rmd
@@
 
-1  Varför trauma är viktigt
-    Mortalitet för unga
-    Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
-      Kanske att chock är stor del av mortalitet.
-
-2  Definiton av chock
-    olika typer av chock: hypovolem, obstruktiv, neurogen, kardiogen och distributiv.
-    trauma oftas de 3 första??
-
-3  Klassifikationer av chock
-    ATLS klassifikation för blödningschock.
-      Hur ska jag ta med:
-        Obstruktiv som tamponad?
-        Neurogen som trauma mot halsrygg?
-      RETTS och NEWS istället?? de används vid triage men är inte chock specifik.
-    Kanske borde använda ATLS 4 kategorier + obstruktiv + neurogen vid sidan??
-
-  (4  Kliniskt för att förbättra omhändertagande används OFI
-    Förklara vad OFI är för något
-      To be able to assess and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI).
-OFI is multidiciplinary morbidity and mortality reviews, where opportunity for improvements are identified in the structure and clinical process of trauma care[@albaaj_patient_2023].
-)
-
-5  Frågeställning:
-    a, Vilka typer av OFI finns hos chockpatienter?
-    b, Koppling mellan nivå av chock och sannolikhet till OFI.
-    Vilken av de två är huvudfrågeställningen?
-    PICO: Population: Chock pat, bland 8000 (2014-2023)
-          Intervention: OFI efter vård?
-          Comparison: mellan chocknivåer?
-          Outcome: Sannolikhet till OFI?
-      Det blir ju inte direkt någon ja/nej svar på frågeställningen...funkar det?
-
 Trauma is a major global public health concern.
 It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
-Shock is a major cause of mortality in trauma patients.
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010] [@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury[@berry_shock_2015].
+It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not[@vang_shock_2022].
+
 The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
@@
 These causes can occur alone or in combination.
-The most usual one during trauma are hypovolemic in form of a major haemorrhage [@teixeira_preventable_2007].
+The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
 
-Clinically to be able to differentiate the degree of shock in context of trauma, the ATLS classification for major haemorrhage can be used.
+To differentiate the severity of trauma shock, we can divide shock into different degrees.
+One of such classification is the ATLS classification for major haemorrhage.
 The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
-However, this only suits hypovolemic shocks and not the other types.
-(Källa) There is other classification for triaging or tracking the clinical deterioration of patients, which can be used for other causes of shocks.
-The common denominator for these classifications is they are all based on vital parameters.
+However, this only meant to be used for hypovolemic shocks and not the other types.
+There is other classification for predicting massive transfusion (TASH, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
+The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
+ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
 
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010].
-To improve trauma care quality, healthcare systems identify Opportunities for Improvement (OFI) through structured reviews of patient cases.
-Multidisciplinary morbidity and mortality reviews play a central role in this process, where clinicians from different specialities together analyse trauma cases to evaluate both the structure and clinical processes involved in trauma care following the Donabedian’s quality of care framework[@albaaj_patient_2023].
-For shock patients, timely interventions such as fluid resuscitation/blood transfusion, intervention, OR and radiology are critical, and delays or missteps can result in adverse outcomes (källa).
-By systematically addressing OFIs, healthcare providers can improve the effectiveness of trauma care, directly impacting mortality and morbidity.
+One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI).
+This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals.
+There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [källa 8].
+In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
+It is based on the three factors of structure, process and outcome.
+By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?].
+This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].
 

### 2024-09-20 project-plan.Rmd
Version 2.2 för manus och 2.0 för projekt redogörelse.

--- 2024-09-11 project-plan.Rmd
@@
   rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
+```{=html}
 <!-- The project plan should be 3-4 pages long and written in
@@
 use this document as a template for your own project plan. -->
-
+```
+```{=html}
 <!-- This is an R Markdown document. Markdown is a simple formatting
@@
 details on using R Markdown see http://rmarkdown.rstudio.com -->
-
+```
+```{=html}
 <!-- This is a comment and will not be present in the compiled
@@
 the console and you should only see headings -->
+```
+# Introduction
 
-Introduction
-============
-
-*This is part A) in the project plan [sv: projektredogörelse]. *
+*This is part A) in the project plan [sv: projektredogörelse].*
 
@@
 what is known about the problem, what is not known, what specific
-knowledge gap the study is supposed to fill and why this is
-important. It should end be stating the aim of the study. The
-background should be 4-5 paragraphs long and each paragraph should be
-between 3-5 sentences. Strive to make the paragraphs about the same
-length.*
+knowledge gap the study is supposed to fill and why this is important.
+It should end be stating the aim of the study. The background should be
+4-5 paragraphs long and each paragraph should be between 3-5 sentences.
+Strive to make the paragraphs about the same length.*
 
@@
 "[@exampleKey9999]". One can convert DOIs' to bibtex entries using
-https://www.doi2bib.org/. If you use Zotero you can simply export 
-entries to bibtex `Ctrl-Shift-C`, however you need to first setup 
-Quick Copy to export to bibtex (https://www.zotero.org/support/creating_bibliographies#quick_copy)*
+<https://www.doi2bib.org/>. If you use Zotero you can simply export
+entries to bibtex `Ctrl-Shift-C`, however you need to first setup Quick
+Copy to export to bibtex
+(<https://www.zotero.org/support/creating_bibliographies#quick_copy>)*
 
-Trauma is a significant global public health concern, responsible for causing death and disability in over four million people each year.
-Among those affected, younger populations are disproportionately impacted, making trauma management a critical issue in healthcare systems worldwide[@noauthor_injuries_nodate].
-The initial management of trauma patients in the hospital setting is extremely time-sensitive and prone to errors, which can adversely affect patient outcomes[@mackersie_pitfalls_2010].
-Trauma patients presenting in shock have a notably higher mortality rate compared to those who are not in shock, underscoring the need for improved care strategies for these high-risk patients[@vang_shock_2022].
-Shock is generally defined as circulatory failure resulting in insufficient perfusion and oxygenation of tissues and organs[@leech_shock_2023].
-There are four primary causes of shock: hypovolemic, obstructive, cardiogenic, and distributive.
-These etiologies can occur in isolation or in combination.
-In the context of trauma, hypovolemic shock, usually caused by massive hemorrhage, is the most common type of shock[@teixeira_preventable_2007].
+Trauma is a major global public health concern.
+It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010] [@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury[@berry_shock_2015].
+It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not[@vang_shock_2022].
 
-To classify shock in trauma 
-The Advanced Trauma Life Support (ATLS) system provides a widely used classification for hypovolemic shock based on blood loss, heart rate, blood pressure, respiratory rate, mental status, and urine output.
-ATLS divides hypovolemic shock into four distinct classes, ranging from mild to severe, and helps guide resuscitative efforts.
-However, this classification is specific to hypovolemic shock and does not account for the other types of shock.
+The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
+There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive.
+These causes can occur alone or in combination.
+The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
 
-To address the broader spectrum of shock types encountered in trauma, triage systems like Rapid Emergency Triage and Treatment System (RETTS) and National Early Warning Score 2 (NEWS2) are commonly used in emergency care settings.
-While these systems are not shock-specific, they offer useful frameworks for assessing patients in various shock states by prioritizing vital signs such as heart rate (HR), blood pressure (BP), respiratory rate (RR), and mental status.
+To differentiate the severity of trauma shock, we can divide shock into different degrees.
+One of such classification is the ATLS classification for major haemorrhage.
+The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
+However, this only meant to be used for hypovolemic shocks and not the other types.
+There is other classification for predicting massive transfusion (TASH, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
+The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
+ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
 
-RETTS is designed for rapid triage and treatment prioritization in emergency departments and uses color-coded categories (red, orange, yellow, green, and blue) to rank patients by urgency.
-It incorporates physiological parameters similar to those used in ATLS, making it applicable to patients in various forms of shock, including hypovolemic, cardiogenic, and distributive shock.
-
-NEWS2, on the other hand, is a scoring system developed to track the clinical deterioration of hospitalized patients.
-Like ATLS and RETTS, it evaluates HR, BP, RR, and mental status, but also includes oxygen saturation and temperature.
-NEWS2 is especially valuable for monitoring patients over time and identifying those at risk of deteriorating, regardless of the underlying cause of shock.
-
-Despite their differences, all three systems—ATLS, RETTS, and NEWS2—share key similarities, particularly in their use of vital sign markers such as heart rate, blood pressure, respiratory rate, and mental status.
-This overlap suggests a potential framework for integrating these systems to better classify trauma patients presenting in any form of shock.
-
-Infoga OFI
+One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI).
+This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals.
+There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [källa 8].
+In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
+It is based on the three factors of structure, process and outcome.
+By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?].
+This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].
 
@@
 
-Methods
-=======
+# Methods
 
 *Refer to the appropriate reporting guideline for details. If you are
-developing, updating or validating a clinical prediction model then
-use
-[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
-you are conducting an observational study, for example a cohort or
+developing, updating or validating a clinical prediction model then use
+[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/).
+If you are conducting an observational study, for example a cohort or
 case control study in which you assess associations between some
@@
 
-Study design
-We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use adjusted logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
+Study design We conducted a registry based retrospective cohort study,
+using data from the trauma registry and trauma care quality database at
+the Karolinska University Hospital in Solna. We then linked the two
+databases to assess what types of OFI occurred in patients arriving in
+shock and use adjusted logistic regression to assess the association
+between the degree of shock and odds of opportunities for improvement.
 
-Setting
-The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2012 and 2023.
-The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review between 2014 and 2023.
-Karolinska University Hospital in Solna acts as a level one trauma center according to critera set by American College of Surgeons(), and treats all major trauma in the greater metropolitan area of Stockholm.
+Setting The trauma registry includes about 14000 patients treated at
+Karolinska University Hospital in Solna between 2012 and 2023. The
+trauma care quality database is a subset of the trauma registry and
+includes about 8000 patients selected for review between 2014 and 2023.
+Karolinska University Hospital in Solna acts as a level one trauma
+center according to critera set by American College of Surgeons(), and
+treats all major trauma in the greater metropolitan area of Stockholm.
 
-Participatns
-Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
-Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
-The identified OFIs are then categorized into broader areas.
-The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields.
-The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
+Participatns Every trauma patient at Karolinska University Hospital is
+included in a morbidity and mortality review process, which involves
+both individual case evaluations by specialized nurses and audit
+filters. Patients identified with a high potential for OFIs are
+discussed at multidisciplinary conferences. The identified OFIs are then
+categorized into broader areas. The multidisciplinary conferences are
+held every six to eight weeks, during which an average of ten patient
+cases are reviewed by experienced specialists from all trauma-related
+fields. The presence or absence of OFIs is determined by consensus among
+all participants and is documented in the trauma care quality database.
 
-To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
-We included all patients in ...
-between ...
-and ...
-We excluded patients younger than 18 and/or were dead on arrival.
+To qualify into the trauma registry, requires admission by trauma team
+activation or any admission with an Injury Severity Score (ISS) of more
+than nine. We included all patients in ... between ... and ... We
+excluded patients younger than 18 and/or were dead on arrival.
 
-Variable and measurements
-To classify the patients "heart rate", "blood pressure", "respiratory rate" and "mental status" will be used as independent variables for classification of shock regardless of their cause.
-Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.
-The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
-The subgroups of OFI will be explored and reported in a list, and therefore non-binary.
+Variable and measurements To classify the patients "heart rate", "blood
+pressure", "respiratory rate" and "mental status" will be used as
+independent variables for classification of shock regardless of their
+cause. Based on these four measurements, the patients will be classified
+into four groups roughly based on the same value as in the ATLS trauma
+shock classifications. The outcome will be OFI recorded in the trauma
+care quality database and are binary when used to calculate the
+association with degree of shock. The subgroups of OFI will be explored
+and reported in a list, and therefore non-binary.
 
-Project Update and Timeline
-===========================
+# Project Update and Timeline
 
-*This is part B) in the project plan [sv: projektredogörelse]. *
+*This is part B) in the project plan [sv: projektredogörelse].*
 
-*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*
+*Save the image generated from the Gantt chard spreadsheet into the
+assets folder, rename it to project-gantt.png*
 
-![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)
+![**The Gantt chart will be displayed here when you place it in the
+assets folder under "project-gantt.png"**](assets/project-gantt.png)
 
-Doing right now:
-- Introduction
-- Methods
+Doing right now: - Introduction - Methods
 
-To do:
-- Coding for regression
-- Analysis
-- Results
-- Discussion
-- Conclusion and Abstract
+To do: - Coding for regression - Analysis - Results - Discussion -
+Conclusion and Abstract
 
-Back-up Plan
-============
+# Back-up Plan
 
@@
 
-References
-==========
+# References
 

### 2024-09-20 manuscript.Rmd
glömde lägga till en källa

@@
 This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals.
-There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [källa 8].
+There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009].
 In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].

### 2024-09-22 manuscript.Rmd
la till ett par källor

@@
 However, this only meant to be used for hypovolemic shocks and not the other types.
-There is other classification for predicting massive transfusion (TASH, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
+There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
 The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
@@
 To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
-We included all patients in ...
-between 2014 and 2023  We excluded patients younger than 15 and/or were dead on arrival.
+We included all patients in "SWETRAU" and "Problem" between 2014 and 2023  We excluded patients younger than 15 and/or were dead on arrival.
 
@@
 <!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
+The outcome was defined as presence of OFI, determined by the multidisciplinary M&M conference.
+An OFI can be expressed as various types of care errors, which include but not limited to communication, lack of resources, treatment delays, level of care, lack of competence and so on.
+In this study, two outcomes will be used, one is all of the types of OFI described, and the other is binary where OFI exists or does not exist.
 
@@
 <!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->
+Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
+
+Version 1,
+Due to the newer versions of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017].
+
+Version 2,
+Due to the newer versions of ATLS classifications lack actual numbers, we will be approximating the values to numbers. For HF... For BP
+
 The first group will be patients presenting with a HR below 100 beats per minute(BPM), respiratory rate between 14 to 20.
@@
 The...
+
+Hur gör vi med de som hamnar mellan klasserna?
 
@@
 
-All available data in the trauma care quality database? will be included, regarding the studied variables. <!-- Mention that you will include all available data-->
+<!-- Mention that you will include all available data-->
+All available data in the trauma care quality database? will be included, regarding the studied variables.
 

### 2024-09-22 manuscript.Rmd
2 version för determine vilken class

@@
 To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
-Heart rate(HR), blood pressure(BP), respiratory rate(RR) and mental status.
+Heart rate(HR), systolic blood pressure(SBP), respiratory rate(RR) and mental status.
 Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.
@@
 Version 1,
-Due to the newer versions of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017].
+Due to the tenth edition of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017]. (INSERT PICTURE)
 
 Version 2,
-Due to the newer versions of ATLS classifications lack actual numbers, we will be approximating the values to numbers. For HF... For BP
-
-The first group will be patients presenting with a HR below 100 beats per minute(BPM), respiratory rate between 14 to 20.
-The original ATLS classification only states normal BP, but no number.
-In our case, we will be using the systolic over 90mmhg as the lower limit up to 65 years old, and 110mmhg as the lower limit for patients older (geriatric patients).
-The...
+Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
+For HR and RR we will be using the numbers from the ninth edition of the classification.
+The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg.
+This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV.
+These two studies, defined hypotension as 110mmhg respective 90mmhg dependent on age, therefore we choose class III as 110-90mmgh and class IV below 90mmhg.
+For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS.
+Therefore we decided to use GCS 15 for class I + II and GCS <15 for class III and IV.
 

### 2024-09-22 project-plan.Rmd
la till kod för knita till word

@@
 author: "Your name here"
+output: 
+  word_document: default
 bibliography: bibliography.bib

### 2024-09-22 manuscript.Rmd
test knit och idé om BE

@@
 
+```{=html}
 <!-- This is an R Markdown document. Markdown is a simple formatting
@@
 the console and you should only see headings -->
-
+```
+```{=html}
 <!-- Structure and most comments are from the Strengthening the
@@
 Epidemiology (STROBE):explanation and elaboration.  -->
-
+```
+```{=html}
 <!-- Indicate the study’s design with a commonly used term in the
@@
 study" -->
+```
+# Abstract
 
-Abstract
-========
-
+```{=html}
 <!-- Provide in the abstract an informative and balanced summary of
 what was done and what was found. Not more than 300 words. -->
+```
+## Background
 
-Background
-----------
+## Methods
 
-Methods
--------
+## Results
 
-Results
--------
+## Conclusion
 
-Conclusion
-----------
+# Introduction
 
-Introduction
-============
-
+```{=html}
 <!-- Explain the scientific background and rationale for the
@@
 6-8 lines long. Support every statement with a citation -->
-
-You can cite document here like this [@exampleKey9999]. Open the file
-bibliography.bib to learn more.
+```
+You can cite document here like this [@exampleKey9999]. Open the file bibliography.bib to learn more.
 
@@
 This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals.
-There are different processes to finding OFIs, but one of the primary ways is structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009].
+There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009].
 In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
@@
 
-Methods 
-=======
+# Methods
 
-Study design
-------------
+## Study design
 
@@
 
-Setting
--------
+## Setting
 
+```{=html}
 <!-- Describe the setting, locations, and relevant dates, including
 periods of recruitment, exposure, follow-up, and data collection -->
-
+```
 The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
@@
 
-Participants
-------------
+## Participants
 
+```{=html}
 <!-- Cohort study: Give the eligibility criteria, and the sources and
@@
 exposed and unexposed -->
-
+```
+```{=html}
 <!-- Case-control study: Give the eligibility criteria, and the
@@
 of controls per case -->
-
+```
+```{=html}
 <!-- Cross-sectional study: Give the eligibility criteria, and the
 sources and methods of selection of participants -->
-
+```
 Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
@@
 
-Variables and data sources/measurements
----------------------------------------
+## Variables and data sources/measurements
 
+```{=html}
 <!-- Clearly define all outcomes, exposures, predictors, potential
@@
 of assessment methods if there is more than one group -->
+```
+<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
 
-<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
 The outcome was defined as presence of OFI, determined by the multidisciplinary M&M conference.
@@
 <!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->
+
 Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
@@
 This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV.
-These two studies, defined hypotension as 110mmhg respective 90mmhg dependent on age, therefore we choose class III as 110-90mmgh and class IV below 90mmhg.
+These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90mmgh and class IV below 90mmhg.
 For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS.
@@
 
-Hur gör vi med de som hamnar mellan klasserna?
+ - Hur gör vi med de som hamnar mellan klasserna?
+To classify, all of the parameters in one class needs to be fulfilled.
+In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
+At same time ATLS manual states one of the common pitfalls is "Diagnosis of shock can be missed when only a single parameter is used." therefore to improve sensitivity we decided to use the worst parameter for classification.
 
-The outcome will be OFI recorded in the trauma care quality database and are binary when used to calculate the association with degree of shock.
-The subgroups of OFI will be explored and reported in a list, and therefore non-binary.
+To prevent classifying non-shock traumatic brain injury as class III and worse shock based on GCS, we decided to only include those who also present a circulatory parameter in accordance with class III or IV.
+ - eller ta bort GCS helt och hållet? För i enlighet med ovan kommer GCS alltid att prioriteras bort och då finns ingen anlending att ha med det...
 
-Bias 
-----
+Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
+We will be using the classification of tenth edition of ATLS, but only the base deficit parameter as it is most accurate [@dunham_comparison_2017].
+
+## Bias
 
@@
 
-Study size 
-----------
+## Study size
 
@@
 <!-- Mention that you will include all available data-->
-All available data in the trauma care quality database? will be included, regarding the studied variables.
 
-Quantitative variables
-----------------------
+All available data in "SWETRAU" and "Problem" database will be included.
 
+## Quantitative variables
+
+```{=html}
 <!-- Explain how quantitative variables were handled in the
@@
 -->
+```
+## Statistical methods
 
-Statistical methods
--------------------
-
+```{=html}
 <!--
@@
 -->
-
+```
 <!-- Start with stating what software you'll use. Then say something about how you will describe the data. Then say something about how you will analyse the data. -->
@@
 
-Results
-=======
+# Results
 
+```{=html}
 <!--
@@
 -->
-
+```
 You can include code in this document like this:
@@
 
-Discussion
-==========
+# Discussion
 
+```{=html}
 <!--
@@
 -->
-
-Conclusion
-==========
+```
+# Conclusion
 

### 2024-09-22 manuscript.Rmd
skrivit version 3

@@
 Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
-We will be using the classification of tenth edition of ATLS, but only the base deficit parameter as it is most accurate [@dunham_comparison_2017].
+We will be only using the base deficit parameter acccording to the tenth edition of the ATLS classification for hemorrhage.
+There are two reasons.
+The first one is that it is superior to vital parameter[@dunham_comparison_2017].
+The second one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
+The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).
 

### 2024-09-22 manuscript.Rmd
bara statistical analysis kvar.

@@
 The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
-The trauma care quality database is a subset of the trauma registry and includes about 8000 patients selected for review.
+The "problem" database is a subset of the trauma registry and includes about ?
+patients selected for review.
 The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.
@@
 ```
+To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
+
 Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
@@
 
-To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
 We included all patients in "SWETRAU" and "Problem" between 2014 and 2023 We excluded patients younger than 15 and/or were dead on arrival.
@@
 
-I will be using R to...
+The statistical analysis will be performed using R, a language and environment for statistical computing.
 

### 2024-09-23 manuscript.Rmd
la till en till anledning

@@
 Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
-We will be only using the base deficit parameter acccording to the tenth edition of the ATLS classification for hemorrhage.
-There are two reasons.
-The first one is that it is superior to vital parameter[@dunham_comparison_2017].
-The second one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
+We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage.
+There are three reasons.
+The first one is ATLS tenth edition lack numerical values for vital parameters.
+The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
+The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
 The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).

### 2024-09-23 manuscript.Rmd
la till statistisk analys delen. Följde kommentaren, men inte intruktionen ovan pga förstår inte begreppen.

@@
 
-The outcome was defined as presence of OFI, determined by the multidisciplinary M&M conference.
-An OFI can be expressed as various types of care errors, which include but not limited to communication, lack of resources, treatment delays, level of care, lack of competence and so on.
-In this study, two outcomes will be used, one is all of the types of OFI described, and the other is binary where OFI exists or does not exist.
+The outcome was defined as presence of atleast one OFI, determined by the multidisciplinary M&M conference in the "problem" registry.
+An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, "preventable death" (den här var väl inte ens med i "problem", fast fanns med i studien som Martin delade på discord) and other errors.
+
+In this study, the outcomes will be studied in two ways, one is the amount of occurrence for various types of OFI in shock patients, and the other is binary where odds of OFI for every shock class i calculated out of presence of OFI or not.
+(kanske inte ens en del av metod)
 
@@
 
-The statistical analysis will be performed using R, a language and environment for statistical computing.
-
-(a) Logistic regression to determine the association between the binary results of OFI existence and teh four degrees of shock.
-
-(b) ?
-
-(c) patient case removed?
-
-(d) patient case removed?
-
-(e) ?
+The statistical analysis will be performed using R, a programing language and environment for statistical computing.
+We will present the type of OFI as percentage distributions and then visualize it in a pie chart.
+Bivariable logistic regression will be used to determine the association between the OFI and degree/class of shock according to ATLS.
+It will be presented as odds ratios (OR) between presence of OFI, and shock classes.
+The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
+All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the “problem” database to ensure objectivity.
 

### 2024-09-23 project-plan.Rmd
uppdaterad metod enligt manuscript

@@
 
-*Refer to the appropriate reporting guideline for details. If you are
-developing, updating or validating a clinical prediction model then use
-[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/).
 If you are conducting an observational study, for example a cohort or
@@
 
-Study design We conducted a registry based retrospective cohort study,
-using data from the trauma registry and trauma care quality database at
-the Karolinska University Hospital in Solna. We then linked the two
-databases to assess what types of OFI occurred in patients arriving in
-shock and use adjusted logistic regression to assess the association
-between the degree of shock and odds of opportunities for improvement.
+## Study design
 
-Setting The trauma registry includes about 14000 patients treated at
-Karolinska University Hospital in Solna between 2012 and 2023. The
-trauma care quality database is a subset of the trauma registry and
-includes about 8000 patients selected for review between 2014 and 2023.
-Karolinska University Hospital in Solna acts as a level one trauma
-center according to critera set by American College of Surgeons(), and
-treats all major trauma in the greater metropolitan area of Stockholm.
+<!-- Present key elements of study design -->
 
-Participatns Every trauma patient at Karolinska University Hospital is
-included in a morbidity and mortality review process, which involves
-both individual case evaluations by specialized nurses and audit
-filters. Patients identified with a high potential for OFIs are
-discussed at multidisciplinary conferences. The identified OFIs are then
-categorized into broader areas. The multidisciplinary conferences are
-held every six to eight weeks, during which an average of ten patient
-cases are reviewed by experienced specialists from all trauma-related
-fields. The presence or absence of OFIs is determined by consensus among
-all participants and is documented in the trauma care quality database.
+We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
 
-To qualify into the trauma registry, requires admission by trauma team
-activation or any admission with an Injury Severity Score (ISS) of more
-than nine. We included all patients in ... between ... and ... We
-excluded patients younger than 18 and/or were dead on arrival.
+## Setting
 
-Variable and measurements To classify the patients "heart rate", "blood
-pressure", "respiratory rate" and "mental status" will be used as
-independent variables for classification of shock regardless of their
-cause. Based on these four measurements, the patients will be classified
-into four groups roughly based on the same value as in the ATLS trauma
-shock classifications. The outcome will be OFI recorded in the trauma
-care quality database and are binary when used to calculate the
-association with degree of shock. The subgroups of OFI will be explored
-and reported in a list, and therefore non-binary.
+The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
+The "problem" database is a subset of the trauma registry and includes about ?
+patients selected for review.
+The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.
+
+## Participants
+
+To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
+
+Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
+Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
+The identified OFIs are then categorized into broader areas.
+The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields.
+The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
+
+We included all patients in "SWETRAU" and "Problem" between 2014 and 2023 We excluded patients younger than 15 and/or were dead on arrival.
+
+## Variables and data sources/measurements
+
+<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
+
+The outcome was defined as presence of atleast one OFI, determined by the multidisciplinary M&M conference in the "problem" registry.
+An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, "preventable death" (den här var väl inte ens med i "problem", fast fanns med i studien som Martin delade på discord) and other errors.
+
+In this study, the outcomes will be studied in two ways, one is the amount of occurrence for various types of OFI in shock patients, and the other is binary where odds of OFI for every shock class i calculated out of presence of OFI or not.
+(kanske inte ens en del av metod)
+
+To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
+Heart rate(HR), systolic blood pressure(SBP), respiratory rate(RR) and mental status.
+Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.
+
+<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->
+
+Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
+
+Version 1, 
+Due to the tenth edition of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017]. (INSERT PICTURE)
+
+Version 2, 
+Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
+For HR and RR we will be using the numbers from the ninth edition of the classification.
+The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg.
+This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV.
+These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90mmgh and class IV below 90mmhg.
+For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS.
+Therefore we decided to use GCS 15 for class I + II and GCS <15 for class III and IV.
+
+ - Hur gör vi med de som hamnar mellan klasserna?
+To classify, all of the parameters in one class needs to be fulfilled.
+In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
+At same time ATLS manual states one of the common pitfalls is "Diagnosis of shock can be missed when only a single parameter is used." therefore to improve sensitivity we decided to use the worst parameter for classification.
+
+To prevent classifying non-shock traumatic brain injury as class III and worse shock based on GCS, we decided to only include those who also present a circulatory parameter in accordance with class III or IV.
+ - eller ta bort GCS helt och hållet? För i enlighet med ovan kommer GCS alltid att prioriteras bort och då finns ingen anlending att ha med det...
+
+Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
+We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage.
+There are three reasons.
+The first one is ATLS tenth edition lack numerical values for vital parameters.
+The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
+The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
+The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).
+
+## Bias
+
+## Study size
+
+All available data in "SWETRAU" and "Problem" database will be included.
+
+## Quantitative variables
+
+## Statistical methods
+
+The statistical analysis will be performed using R, a programing language and environment for statistical computing.
+We will present the type of OFI as percentage distributions and then visualize it in a pie chart.
+Bivariable logistic regression will be used to determine the association between the OFI and degree/class of shock according to ATLS.
+It will be presented as odds ratios (OR) between presence of OFI, and shock classes.
+The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
+All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the “problem” database to ensure objectivity.
 

### 2024-09-26 Sparade snuttar.R
testat börjat med att dela shock klasserna

@@ -1,0 +1,10 @@
+function(x){
+  x <- as.character(x)
+  x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)
+  x <- as.numeric(x)
+  return(x)
+}
+
+BE <- convert_number(data$swetrau_scrambled$ed_be_art)
+
+BEclass1 <- swetrau_scrambled[BE<0 & BE>-2, ]

### 2024-09-27 Sparade snuttar.R
anndra klasser

@@ -8,3 +8,7 @@ function(x){
 BE <- convert_number(data$swetrau_scrambled$ed_be_art)
 
 BEclass1 <- swetrau_scrambled[BE<0 & BE>-2, ]
+BEclass2 <- swetrau_scrambled[BE<-2 & BE>-6, ]
+BEclass3 <- swetrau_scrambled[BE<-6 & BE>-10, ]
+BEclass3 <- swetrau_scrambled[BE<-10, ]
+

### 2024-09-27 Sparade snuttar.R
fel variabel namn

@@ -10,5 +10,5 @@ BE <- convert_number(data$swetrau_scrambled$ed_be_art)
 BEclass1 <- swetrau_scrambled[BE<0 & BE>-2, ]
 BEclass2 <- swetrau_scrambled[BE<-2 & BE>-6, ]
 BEclass3 <- swetrau_scrambled[BE<-6 & BE>-10, ]
-BEclass3 <- swetrau_scrambled[BE<-10, ]
+BEclass4 <- swetrau_scrambled[BE<-10, ]
 

### 2024-09-29 project-plan.Rmd
knit
ändrade rätt i snuttarna så att det är faktiska negativa tal
la till nya intro

@@
 
+This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
+
+## Epidemiology
+Trauma is a major global public health concern.
+It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected (2).
+Globally low and middle income countries are disproportionately affected, but high income countries like Sweden also face considerable challenges(källa).
+In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old.
+(Swetrau ÅR, 2021).
+This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs.
+
+## Trauma
+Trauma are generally divided into two groups, blunt and penetrating injury.
+The most common type in Sweden is blunt injury, constituting around 90% of all trauma.
+The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma(Swetrau ÅR, 2021).
+
+Blunt and penetrating trauma have their own distinct clinical implications.
+Blunt trauma often results from impacts of different energy factor, and depending on the energy can result in anything from abbrations to fractures and internal organ damage.
+The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), at 75% of all blunt trauma(advances of abdominal trauma).
+The most common cause is traffic related such as, rapid deceleration leading to driver crashing into car furniture and causing contusions or rupture of internal organs.
+
+Penetrating trauma on the other hand are often caused by gunshots or stab wounds(Emergency care and transportation of the sick and injured).
+The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and heart tamponade.
+
+A consequence of both types of trauma is possibility of significant bleeding, which in turn can result in hemorrhagic shock.
+
+## Shock
+The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ(9).
+Some of the common symptoms are tachycardia, hypotenstion and altered mental status.
+Shock are usually classified into four different types based on physiological mechanism.
+The types are, hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics(källa).
+These causes can occur alone or in combination, as there can be multiple disruptions on the circulatory system.
+Common examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - pulmonary embolism, cardiogenic - STEMI  and, distributive - sepsis(källa).
+
+The most common type of shock during trauma is the hypovolemic type due to a major hemorrhage(4).
+Hemorrhage leads to low blood volume, which beside the common central symptoms noted before, also results in peripheral vasoconstriction, which corresponds with cold, wet and pale extremities(källa).
+
+## Shock classifications
+To differentiate the severity of trauma shock, we can divide it into different degrees.
+One of such classification is the ATLS classification for major haemorrhage.
+ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
+However, this only meant to be used for hypovolemic shocks and not the other types.
+There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
+The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
+ATLS in this regard, even though having its own shortcomings(10), still is one of the most used classifications for this topic(Källa).
+
+## Trauma quality improvement and oppertunities for improvement
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone (3) [(4)](5), and shock is the most common potentially preventable death within the first 24 hour of injury(6), as the midian time to death is 2 hours(Tisherman, 2015).
+To address these challenges and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
+QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality(WHO).
+By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally (Juillard + Hashmi).
+
+One way to improve trauma care quality is through identifying Opportunities for Improvement (OFI).
+This method has an advantage compared to traditional mortality reviews in that it also includes non-fatal outcomes, which gives a broader perspective and a chance to reduce chronic morbidity.
+There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality (M&M) reviews of patient cases (11).
+In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used(12).
+It is based on the three factors of structure, process and outcome.
+By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?].
+This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].
+
+## Aims
 This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
@@
 
-To do: - Coding for regression - Analysis - Results - Discussion -
+To do: - Coding to sort data - Coding for regression - Analysis - Results - Discussion -
 Conclusion and Abstract

### 2024-09-29 Sparade snuttar.R

@@ -8,7 +8,11 @@ function(x){
 BE <- convert_number(data$swetrau_scrambled$ed_be_art)
 
 BEclass1 <- swetrau_scrambled[BE<0 & BE>-2, ]
-BEclass2 <- swetrau_scrambled[BE<-2 & BE>-6, ]
-BEclass3 <- swetrau_scrambled[BE<-6 & BE>-10, ]
-BEclass4 <- swetrau_scrambled[BE<-10, ]
+BEclass2 <- swetrau_scrambled[BE<(-2) & BE>(-6), ]
+BEclass3 <- swetrau_scrambled[BE<(-6) & BE>-10, ]
+BEclass4 <- swetrau_scrambled[BE<(-10), ]
 
+BEclass1num <- convert_number(BEclass1$ed_be_art)
+BEclass2num <- convert_number(BEclass2$ed_be_art)
+BEclass3num <- convert_number(BEclass3$ed_be_art)
+BEclass4num <- convert_number(BEclass4$ed_be_art)

### 2024-09-29 project-plan.Rmd
la till metod

@@
 ## Epidemiology
+
 Trauma is a major global public health concern.
@@
 ## Trauma
+
 Trauma are generally divided into two groups, blunt and penetrating injury.
@@
 ## Shock
+
 The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ(9).
@@
 ## Shock classifications
+
 To differentiate the severity of trauma shock, we can divide it into different degrees.
@@
 ## Trauma quality improvement and oppertunities for improvement
+
 The initial management of trauma patients once they reach hospital is very time sensitive and error prone (3) [(4)](5), and shock is the most common potentially preventable death within the first 24 hour of injury(6), as the midian time to death is 2 hours(Tisherman, 2015).
@@
 
-If you are conducting an observational study, for example a cohort or
-case control study in which you assess associations between some
-exposure and an outcome then use
-[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).*
-
 ## Study design
 
-<!-- Present key elements of study design -->
-
 We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
 
@@
 
-The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
-The "problem" database is a subset of the trauma registry and includes about ?
-patients selected for review.
+The trauma registry includes patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
+The trauma care quality database is a subset of the trauma registry and includes patients selected for review.
 The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.
@@
 
-To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
+We included all patients in trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.
 
-Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
+The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or a already admitted patient with an Injury Severity Score (ISS) greater than nine.
+Every trauma patient (registered in the trauma registry) at Karolinska University Hospital is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
 Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
@@
 
-We included all patients in "SWETRAU" and "Problem" between 2014 and 2023 We excluded patients younger than 15 and/or were dead on arrival.
-
 ## Variables and data sources/measurements
 
-<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
-
-The outcome was defined as presence of atleast one OFI, determined by the multidisciplinary M&M conference in the "problem" registry.
-An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, "preventable death" (den här var väl inte ens med i "problem", fast fanns med i studien som Martin delade på discord) and other errors.
-
-In this study, the outcomes will be studied in two ways, one is the amount of occurrence for various types of OFI in shock patients, and the other is binary where odds of OFI for every shock class i calculated out of presence of OFI or not.
-(kanske inte ens en del av metod)
+The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the “problem” registry.
+An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
 
@@
 
-<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->
-
-Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
-
-Version 1, 
-Due to the tenth edition of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017]. (INSERT PICTURE)
-
-Version 2, 
+Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[(13)](14)(15).
 Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
@@
 The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg.
-This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV.
-These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90mmgh and class IV below 90mmhg.
-For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS.
-Therefore we decided to use GCS 15 for class I + II and GCS <15 for class III and IV.
+This selection is based on tenth edition, where SBP is defined as normal in class I to II and normal/below in class III and clearly below normal in class IV.
+These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[(17)](18), therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
+We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).
+Urine output and pulse pressure were also neglected, due to lack of data.
 
- - Hur gör vi med de som hamnar mellan klasserna?
 To classify, all of the parameters in one class needs to be fulfilled.
-In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
-At same time ATLS manual states one of the common pitfalls is "Diagnosis of shock can be missed when only a single parameter is used." therefore to improve sensitivity we decided to use the worst parameter for classification.
+In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al(16).
+At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification.
 
-To prevent classifying non-shock traumatic brain injury as class III and worse shock based on GCS, we decided to only include those who also present a circulatory parameter in accordance with class III or IV.
- - eller ta bort GCS helt och hållet? För i enlighet med ovan kommer GCS alltid att prioriteras bort och då finns ingen anlending att ha med det...
-
-Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
 We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage.
@@
 The first one is ATLS tenth edition lack numerical values for vital parameters.
-The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
+The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality(16).
 The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
@@
 
+The trauma care quality database is build up in a way where every error can be classified into a certain type of OFI, however for paitent that died there is only the category “preventable death”.
+Therefore, survivorship bias may occur as we dont have the detail which type of OFI actually occured.
+
 ## Study size
 
-All available data in "SWETRAU" and "Problem" database will be included.
+All available data in trauma registry and trauma care quality database will be included.
 
@@
 
-The statistical analysis will be performed using R, a programing language and environment for statistical computing.
+The statistical analysis will be performed using R, a programming language and environment for statistical computing.
 We will present the type of OFI as percentage distributions and then visualize it in a pie chart.
-Bivariable logistic regression will be used to determine the association between the OFI and degree/class of shock according to ATLS.
+Unadjusted and adjusted logistic regression will be used to determine the association between the OFI, degree/class of shock according to ATLS and other patient factors such as age, sex, preinjury ASA and ISS.
 It will be presented as odds ratios (OR) between presence of OFI, and shock classes.
 The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
-All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the “problem” database to ensure objectivity.
+All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.
+
+# Ethics
+
+The handling of registry data and the extraction of information from medical records pose a potential risk to patient privacy.
+To mitigate this, patients were anonymized after data collection, with personal identity numbers replaced by unique study identifiers.
+
+From a self-determination standpoint, the Swedish trauma registry does not require patient consent for data registration.
+However, it is mandatory for the caregiver to inform patients of their inclusion in the registry.
+Patients have the right to view, correct, opt out of, and request the deletion of their data, without any impact on the care they receive.
+
+In this study, data will be presented at the group level, ensuring that no individual patient details can be identified.
+Ethical approvals has been obtained for this study, with the record numbers: 2021-02541, 2021-03531 and 2023-02975-02.
+
+The benefits of this study, in relation to its risks, are substantial due to its potential to improve care for future critically ill patients.
+This research is a step toward reducing mortality and morbidity in a significant portion of the younger and economically active population worldwide.
 
@@
 
-*This is part C) in the project plan [sv: projektredogörelse]*
+After discussion with my supervisor, it seems like the part which have the highest possibility of taking too much time is coding, and therefore he and other cosupervisor will assist with coding if needed.
 

### 2024-09-29 project-plan.Rmd
la till källor

@@
 Trauma is a major global public health concern.
-It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected (2).
+It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
 Globally low and middle income countries are disproportionately affected, but high income countries like Sweden also face considerable challenges(källa).
@@
 Blunt trauma often results from impacts of different energy factor, and depending on the energy can result in anything from abbrations to fractures and internal organ damage.
-The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), at 75% of all blunt trauma(advances of abdominal trauma).
+The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), at 75% of all blunt trauma[@isenhour_advances_2007].
 The most common cause is traffic related such as, rapid deceleration leading to driver crashing into car furniture and causing contusions or rupture of internal organs.
 
-Penetrating trauma on the other hand are often caused by gunshots or stab wounds(Emergency care and transportation of the sick and injured).
+Penetrating trauma on the other hand are often caused by gunshots or stab wounds[@internet_archive_emergency_2002].
 The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and heart tamponade.
@@
 
-The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ(9).
+The definition for shock is circulatory failure leading to insufficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
 Some of the common symptoms are tachycardia, hypotenstion and altered mental status.
@@
 
-The most common type of shock during trauma is the hypovolemic type due to a major hemorrhage(4).
+The most common type of shock during trauma is the hypovolemic type due to a major hemorrhage[@teixeira_preventable_2007].
 Hemorrhage leads to low blood volume, which beside the common central symptoms noted before, also results in peripheral vasoconstriction, which corresponds with cold, wet and pale extremities(källa).
@@
 
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone (3) [(4)](5), and shock is the most common potentially preventable death within the first 24 hour of injury(6), as the midian time to death is 2 hours(Tisherman, 2015).
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24 hour of injury[@berry_shock_2015], as the midian time to death is 2 hours[@tisherman_detailed_2015].
 To address these challenges and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
-QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality(WHO).
-By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally (Juillard + Hashmi).
+QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality[@world_health_organization_guidelines_2009].
+By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally[@juillard_establishing_2009][@hashmi_hospital-based_2013].
 
@@
 This method has an advantage compared to traditional mortality reviews in that it also includes non-fatal outcomes, which gives a broader perspective and a chance to reduce chronic morbidity.
-There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality (M&M) reviews of patient cases (11).
-In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used(12).
+There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality (M&M) reviews of patient cases[@world_health_organization_guidelines_2009].
+In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
 It is based on the three factors of structure, process and outcome.
@@
 
-Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[(13)](14)(15).
+Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
 Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
@@
 This selection is based on tenth edition, where SBP is defined as normal in class I to II and normal/below in class III and clearly below normal in class IV.
-These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[(17)](18), therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
+These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
 We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).
@@
 To classify, all of the parameters in one class needs to be fulfilled.
-In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al(16).
-At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification.
+In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
+At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification[ATLS].
 
@@
 The first one is ATLS tenth edition lack numerical values for vital parameters.
-The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality(16).
+The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
 The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.

### 2024-09-29 project-plan.Rmd
missade en källa och tog bart mallen

@@
 # Introduction
-
-*This is part A) in the project plan [sv: projektredogörelse].*
-
-*The background/introduction should summarise the scope of the problem,
-what is known about the problem, what is not known, what specific
-knowledge gap the study is supposed to fill and why this is important.
-It should end be stating the aim of the study. The background should be
-4-5 paragraphs long and each paragraph should be between 3-5 sentences.
-Strive to make the paragraphs about the same length.*
-
-*bibliography.bib includes an example reference. Add additional
-references in that file as bibtex entries and then cite in text using:
-"[@exampleKey9999]". One can convert DOIs' to bibtex entries using
-<https://www.doi2bib.org/>. If you use Zotero you can simply export
-entries to bibtex `Ctrl-Shift-C`, however you need to first setup Quick
-Copy to export to bibtex
-(<https://www.zotero.org/support/creating_bibliographies#quick_copy>)*
-
-Trauma is a major global public health concern.
-It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010] [@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury[@berry_shock_2015].
-It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not[@vang_shock_2022].
-
-The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
-There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive.
-These causes can occur alone or in combination.
-The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
-
-To differentiate the severity of trauma shock, we can divide shock into different degrees.
-One of such classification is the ATLS classification for major haemorrhage.
-The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
-However, this only meant to be used for hypovolemic shocks and not the other types.
-There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
-The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
-ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
-
-One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI).
-This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals.
-There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009].
-In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
-It is based on the three factors of structure, process and outcome.
-By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?].
-This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].
-
-This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
 
@@
 The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
-ATLS in this regard, even though having its own shortcomings(10), still is one of the most used classifications for this topic(Källa).
+ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
 

### 2024-09-29 project-plan.Rmd
fixat de flesta källor, två kvar och kanske lägga några extra om jag hinner i trauma delen

@@
 Trauma is a major global public health concern.
-It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
-Globally low and middle income countries are disproportionately affected, but high income countries like Sweden also face considerable challenges(källa).
-In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old.
-(Swetrau ÅR, 2021).
-This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs.
+It causes over four million death and affects hundreds of millions each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
+Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall number but the younger demographic are also affected.
+In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old[@noauthor_arsrapport_2022].
+This not only leads to premature loss of life but also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs[@noauthor_injuries_nodate].
 
@@
 The most common type in Sweden is blunt injury, constituting around 90% of all trauma.
-The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma(Swetrau ÅR, 2021).
+The most common causes for blunt trauma are traffic related and falls, which account for roughly ½ and ⅓ of all blunt trauma[@noauthor_arsrapport_2022].
 
 Blunt and penetrating trauma have their own distinct clinical implications.
-Blunt trauma often results from impacts of different energy factor, and depending on the energy can result in anything from abbrations to fractures and internal organ damage.
-The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), at 75% of all blunt trauma[@isenhour_advances_2007].
-The most common cause is traffic related such as, rapid deceleration leading to driver crashing into car furniture and causing contusions or rupture of internal organs.
+Blunt trauma often results from impacts of different energy factors, and depending on the energy can result in anything from abbrations to fractures and internal organ damage.
+The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), which is often caused by traffic accidents.
+When counting motor vehicle collision and pedestrian versus auto accidents together, they account for 75% of all blunt trauma[@isenhour_advances_2007].
 
 Penetrating trauma on the other hand are often caused by gunshots or stab wounds[@internet_archive_emergency_2002].
-The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and heart tamponade.
+The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and cardiac tamponade.
 
@@
 Shock are usually classified into four different types based on physiological mechanism.
-The types are, hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics(källa).
+The types are: hypovolemic, obstructive, cardiogenic and distributive, and they all have their own special characteristics[@standl_nomenclature_2018].
 These causes can occur alone or in combination, as there can be multiple disruptions on the circulatory system.
-Common examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - pulmonary embolism, cardiogenic - STEMI  and, distributive - sepsis(källa).
+Examples of diagnoses for each type are, hypovolemic - bleeding, obstructive - tension pneumothorax, cardiogenic - myocardial pump failure and distributive - sepsis[@standl_nomenclature_2018].
 
 The most common type of shock during trauma is the hypovolemic type due to a major hemorrhage[@teixeira_preventable_2007].
-Hemorrhage leads to low blood volume, which beside the common central symptoms noted before, also results in peripheral vasoconstriction, which corresponds with cold, wet and pale extremities(källa).
+Hemorrhage leads to low blood volume, which beside the common central symptoms noted before, also results in peripheral vasoconstriction, which corresponds with cold, wet and pale extremities[@raja_initial_2024].
 
@@
 One of such classification is the ATLS classification for major haemorrhage.
-ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
+ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output[@noauthor_atls_2018].
 However, this only meant to be used for hypovolemic shocks and not the other types.
@@
 The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
-ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
+ATLS in this regard, even though having its own shortcomings, still is one of the common classifications and part of the student course manual of the internationally accepted ATLS trauma management system[@guly_vital_2011][@noauthor_atls_2018].
 
@@
 The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24 hour of injury[@berry_shock_2015], as the midian time to death is 2 hours[@tisherman_detailed_2015].
-To address these challenges and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
+To address these and other challenges within trauma care and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
 QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality[@world_health_organization_guidelines_2009].
@@
 In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
-At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification[ATLS].
+At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification[@noauthor_atls_2018].
 

### 2024-09-30 project-plan.Rmd
fixat alla källor

@@
 There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality (M&M) reviews of patient cases[@world_health_organization_guidelines_2009].
-In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
+In order for the review content to be as comprehensive as possible and taking in a system oriented approach, the Donabedian quality of care framework can be used[@donabedian_quality_1988][@world_health_organization_guidelines_2009].
 It is based on the three factors of structure, process and outcome.
-By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?].
-This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].
+By following this framework, healthcare providers will be able to systematically find OFIs as part of QI.
+This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[@juillard_establishing_2009][@hashmi_hospital-based_2013].
 

### 2024-09-30 project-plan.Rmd
nya penetrating trauma del

@@
 
-Penetrating trauma on the other hand are often caused by gunshots or stab wounds[@internet_archive_emergency_2002].
-The most common location is thorax and injuries there can lead to damage of the lungs and heart, which in turn interfere with breathing and circulation, such as pneumo/hemothorax and cardiac tamponade.
+Penetrating trauma, often caused by gunshots or stab wounds, commonly affects the small intestine, colon, liver, and intra-abdominal blood vessels.
+These injuries are life-threatening due to significant bleeding from large vessels and the liver’s extensive blood supply.
+There is also a risk of other complications, such as pancreatic damage leading to autodigestion, and intestinal perforation resulting in contamination from fecal matter.
+Consequently, penetrating trauma frequently leads to hypovolemic shock and peritonitis[@lotfollahzadeh_penetrating_2023].
 
@@
 The statistical analysis will be performed using R, a programming language and environment for statistical computing.
-We will present the type of OFI as percentage distributions and then visualize it in a pie chart.
+We will present the types of OFI as percentage distributions and then visualize it in a pie chart.
 Unadjusted and adjusted logistic regression will be used to determine the association between the OFI, degree/class of shock according to ATLS and other patient factors such as age, sex, preinjury ASA and ISS.

### 2024-10-03 Sparade snuttar.R
la till i main

@@ -16,3 +16,9 @@ BEclass1num <- convert_number(BEclass1$ed_be_art)
 BEclass2num <- convert_number(BEclass2$ed_be_art)
 BEclass3num <- convert_number(BEclass3$ed_be_art)
 BEclass4num <- convert_number(BEclass4$ed_be_art)
+
+newNiss <- as.numeric(NISS)
+newISS <- as.numeric(ISS)
+newhost_care_level <- as.numeric(host_care_level)
+
+DataFrame1 <- data.frame(host_care_level, newNiss, newISS)

### 2024-10-03 main.R

@@ -24,3 +24,22 @@ noacsr::source_all_functions()
 data <- import_data(test = TRUE)
 
 ## Whatever you do next, maybe clean data?
+
+function(x){
+  x <- as.character(x)
+  x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)
+  x <- as.numeric(x)
+  return(x)
+}
+
+BE <- convert_number(data$swetrau_scrambled$ed_be_art)
+
+BEclass1 <- swetrau_scrambled[BE<0 & BE>-2, ]
+BEclass2 <- swetrau_scrambled[BE<(-2) & BE>(-6), ]
+BEclass3 <- swetrau_scrambled[BE<(-6) & BE>-10, ]
+BEclass4 <- swetrau_scrambled[BE<(-10), ]
+
+BEclass1num <- convert_number(BEclass1$ed_be_art)
+BEclass2num <- convert_number(BEclass2$ed_be_art)
+BEclass3num <- convert_number(BEclass3$ed_be_art)
+BEclass4num <- convert_number(BEclass4$ed_be_art)

### 2024-10-05 project-plan.Rmd
la till i main
glömde ändra i gammal stycke
tog bort en "problem" som jag glömde ta bort.

@@
 
+(Infoga bild på ATLS tenth edition)
+
 ## Trauma quality improvement and oppertunities for improvement
@@
 
-The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the “problem” registry.
+The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the trauma care quality registry.
 An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
@@
 To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
-Heart rate(HR), systolic blood pressure(SBP), respiratory rate(RR) and mental status.
+Heart rate(HR), systolic blood pressure(SBP) and respiratory rate(RR).
 Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.

### 2024-10-11 main.R
fixat BE classificering.

@@ -51,22 +51,52 @@ study.data <- merged.data |>
          res_survival,
          ofi)
 
-OfiBE <- merged.data |> 
-  select(ed_be_art,
-         ofi)
-
 # Exclude patients who were not reviewed for the presence of OFI
 study.sample <- study.data |>
   filter(!is.na(ofi))
 
+# Exclude patients who did not have BE data
 study.sampleBE <- study.sample |>
   filter(!is.na(ed_be_art))
 
-OfiBEO <- OfiBE |>
-  filter(!is.na(ofi))
+# The display sample
+display.sample <- study.sample |> 
+  select(pt_age_yrs, 
+         pt_Gender, 
+         pt_asa_preinjury,
+         ed_sbp_value,
+         ISS,
+         ofi)
 
-OfiBEOB <- OfiBEO |>
-  filter(!is.na(ed_be_art))
+# Adding shock classification
+
+convert_number <- function(x){
+  x <- as.character(x)
+  x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)
+  x <- as.numeric(x)
+  return(x)
+}
+
+BEnum <- convert_number(study.sample$ed_be_art)
+
+study.sample <- study.sample %>%
+  mutate(BE_class = case_when(
+    BEnum < (-10) ~ "Class 4",
+    BEnum >= (-10) & BEnum < (-6) ~ "Class 3",
+    BEnum >= (-6) & BEnum < (-2) ~ "Class 2",  
+    BEnum >= (-2) & BEnum < (0) ~ "Class 1",   
+    is.na(BEnum) ~ NA_character_,        
+    TRUE ~ "no shock"
+  ))
+
+# Remove unused variables. 
+study.sample <- study.sample |> 
+  select( -inj_mechanism, 
+         -ed_gcs_sum,
+         -ed_rr_value,
+         -ed_be_art,
+         -host_care_level,
+         -res_survival)
 
 # Label variables
 var_label(study.sample$pt_age_yrs) <- "Age in years"
@@ -78,7 +108,7 @@ sample.characteristics.table <- tbl_summary(study.sample,
 
 ## Whatever you do next, maybe clean data?
 
-function(x){
+convert_number <- function(x){
   x <- as.character(x)
   x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)
   x <- as.numeric(x)
[…]

### 2024-10-11 main.R
la till BE som numerisk

@@ -89,6 +89,10 @@ study.sample <- study.sample %>%
     TRUE ~ "no shock"
   ))
 
+# Re-add the BE column as numeric to `study.sample`
+study.sample <- study.sample %>%
+  mutate(ed_be_art_numeric = BEnum)
+
 # Remove unused variables. 
 study.sample <- study.sample |> 
   select( -inj_mechanism, 
@@ -101,6 +105,7 @@ study.sample <- study.sample |>
 # Label variables
 var_label(study.sample$pt_age_yrs) <- "Age in years"
 var_label(study.sample$ofi) <- "Opportunities for improvement"
+var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
 
 # Create a table of sample characteristics
 sample.characteristics.table <- tbl_summary(study.sample,

### 2024-10-11 main.R
la till sbp och visa tabell i koden

@@ -68,8 +68,7 @@ display.sample <- study.sample |>
          ISS,
          ofi)
 
-# Adding shock classification
-
+# Converting ed_be_art to numeric
 convert_number <- function(x){
   x <- as.character(x)
   x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)
@@ -79,6 +78,7 @@ convert_number <- function(x){
 
 BEnum <- convert_number(study.sample$ed_be_art)
 
+# BE shock classification
 study.sample <- study.sample %>%
   mutate(BE_class = case_when(
     BEnum < (-10) ~ "Class 4",
@@ -93,6 +93,15 @@ study.sample <- study.sample %>%
 study.sample <- study.sample %>%
   mutate(ed_be_art_numeric = BEnum)
 
+# SBP shock classifiaction
+study.sample <- study.sample %>%
+  mutate(SBP_class = case_when(
+    ed_sbp_value < (90) ~ "Class 2",
+    ed_sbp_value >= (90) & ed_sbp_value < (110) ~ "Class 1",
+    is.na(ed_sbp_value) ~ NA_character_,        
+    TRUE ~ "no shock"
+  ))
+
 # Remove unused variables. 
 study.sample <- study.sample |> 
   select( -inj_mechanism, 
@@ -103,14 +112,17 @@ study.sample <- study.sample |>
          -res_survival)
 
 # Label variables
-var_label(study.sample$pt_age_yrs) <- "Age in years"
-var_label(study.sample$ofi) <- "Opportunities for improvement"
+var_label(study.sample$pt_age_yrs) <- "Age (Years)"
+var_label(study.sample$ofi) <- "Opportunities for improvement (Y/N)"
 var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
+var_label(study.sample$SBP_class) <- "Shock class classfied according to SBP"
+var_label(study.sample$BE_class) <- "Shock class classified according to BE"
 
 # Create a table of sample characteristics
 sample.characteristics.table <- tbl_summary(study.sample,
                                             by = ofi)
 
+sample.characteristics.table
 ## Whatever you do next, maybe clean data?
 
 convert_number <- function(x){

### 2024-10-11 main.R
la till rubriker

@@ -117,6 +117,9 @@ var_label(study.sample$ofi) <- "Opportunities for improvement (Y/N)"
 var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
 var_label(study.sample$SBP_class) <- "Shock class classfied according to SBP"
 var_label(study.sample$BE_class) <- "Shock class classified according to BE"
+var_label(study.sample$pt_Gender) <- "Gender (M/F)"
+var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
+var_label(study.sample$ISS) <- "Injury Severity Score"
 
 # Create a table of sample characteristics
 sample.characteristics.table <- tbl_summary(study.sample,

### 2024-10-11 skräpsnuttar.R
la till fil där jag kan lägga in skräp som jag kanske senare behöver.

@@ -1,0 +1,131 @@
+convert_number <- function(x){
+  x <- as.character(x)
+  x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)
+  x <- as.numeric(x)
+  return(x)
+}
+
+BE <- convert_number(study.sampleBE$ed_be_art)
+
+BEclass1 <- study.sampleBE[BE<0 & BE>=(-2), ]
+BEclass2 <- study.sampleBE[BE<(-2) & BE>=(-6), ]
+BEclass3 <- study.sampleBE[BE<(-6) & BE>=-10, ]
+BEclass4 <- study.sampleBE[BE<(-10), ]
+
+BEclass1num <- convert_number(BEclass1$ed_be_art)
+BEclass2num <- convert_number(BEclass2$ed_be_art)
+BEclass3num <- convert_number(BEclass3$ed_be_art)
+BEclass4num <- convert_number(BEclass4$ed_be_art)
+
+antalBE1 <- sum(length(BEclass1num))
+antalBE2 <- sum(length(BEclass2num))
+antalBE3 <- sum(length(BEclass3num))
+antalBE4 <- sum(length(BEclass4num))
+
+
+ofi10 <- ifelse(study.sample$ofi == "Yes", 1, 0)
+
+
+BEc1Y <- filter(BEclass1, ofi == "Yes")
+BEc1N <- filter(BEclass1, ofi == "No")
+BEc2Y <- filter(BEclass2, ofi == "Yes")
+BEc2N <- filter(BEclass2, ofi == "No")
+BEc3Y <- filter(BEclass3, ofi == "Yes")
+BEc3N <- filter(BEclass3, ofi == "No")
+BEc4Y <- filter(BEclass4, ofi == "Yes")
+BEc4N <- filter(BEclass4, ofi == "No")
+
+tabell <- as.table(rbind(c(sum(length(convert_number(BEc1Y$ed_be_art))), sum(length(convert_number(BEc2Y$ed_be_art))), sum(length(convert_number(BEc3Y$ed_be_art))), sum(length(convert_number(BEc4Y$ed_be_art)))), 
+                         c(sum(length(convert_number(BEc1N$ed_be_art))), sum(length(convert_number(BEc2N$ed_be_art))), sum(length(convert_number(BEc3N$ed_be_art))), sum(length(convert_number(BEc4N$ed_be_art))))))
+dimnames(tabell) <- list(OFI = c("Yes", "No"),
[…]

### 2024-10-11 main.R
la till kön som kategoriska istället för numerisk

@@ -55,19 +55,6 @@ study.data <- merged.data |>
 study.sample <- study.data |>
   filter(!is.na(ofi))
 
-# Exclude patients who did not have BE data
-study.sampleBE <- study.sample |>
-  filter(!is.na(ed_be_art))
-
-# The display sample
-display.sample <- study.sample |> 
-  select(pt_age_yrs, 
-         pt_Gender, 
-         pt_asa_preinjury,
-         ed_sbp_value,
-         ISS,
-         ofi)
-
 # Converting ed_be_art to numeric
 convert_number <- function(x){
   x <- as.character(x)
@@ -102,9 +89,17 @@ study.sample <- study.sample %>%
     TRUE ~ "no shock"
   ))
 
+# Convert gender from numeric to categorical
+study.sample <- study.sample %>%
+  mutate(pt_Gender = case_when(
+    pt_Gender == 1 ~ "Male",
+    pt_Gender == 2 ~ "Female",
+    TRUE ~ NA_character_
+  ))
+
 # Remove unused variables. 
 study.sample <- study.sample |> 
-  select( -inj_mechanism, 
+  select(-inj_mechanism, 
          -ed_gcs_sum,
          -ed_rr_value,
          -ed_be_art,
@@ -117,9 +112,9 @@ var_label(study.sample$ofi) <- "Opportunities for improvement (Y/N)"
 var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
 var_label(study.sample$SBP_class) <- "Shock class classfied according to SBP"
 var_label(study.sample$BE_class) <- "Shock class classified according to BE"
-var_label(study.sample$pt_Gender) <- "Gender (M/F)"
 var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
 var_label(study.sample$ISS) <- "Injury Severity Score"
+var_label(display.sample$pt_Gender) <- "Gender"
 
 # Create a table of sample characteristics
 sample.characteristics.table <- tbl_summary(study.sample,
@@ -128,137 +123,7 @@ sample.characteristics.table <- tbl_summary(study.sample,
 sample.characteristics.table
[…]

### 2024-10-11 skräpsnuttar.R

@@ -1,3 +1,21 @@
+# Exclude patients who did not have BE data
+study.sampleBE <- study.sample |>
+  filter(!is.na(ed_be_art))
+
+# The display sample
+display.sample <- study.sample |> 
+  select(pt_age_yrs, 
+         pt_Gender, 
+         pt_asa_preinjury,
+         ed_sbp_value,
+         ISS,
+         ofi)
+
+
+
+
+
+
 convert_number <- function(x){
   x <- as.character(x)
   x <- gsub(pattern = ",", replacement = ".",x = x, fixed = TRUE)

### 2024-10-11 main.R
fixat asa klassificering

@@ -97,6 +97,13 @@ study.sample <- study.sample %>%
     TRUE ~ NA_character_
   ))
 
+# Converting 999 to unknown for pt_asa_preinjury
+study.sample <- study.sample %>%
+  mutate(pt_asa_preinjury = case_when(
+    pt_asa_preinjury == 999 ~ "Unknown",
+    TRUE ~ as.character(pt_asa_preinjury)
+  ))
+
 # Remove unused variables. 
 study.sample <- study.sample |> 
   select(-inj_mechanism, 
@@ -114,7 +121,8 @@ var_label(study.sample$SBP_class) <- "Shock class classfied according to SBP"
 var_label(study.sample$BE_class) <- "Shock class classified according to BE"
 var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
 var_label(study.sample$ISS) <- "Injury Severity Score"
-var_label(display.sample$pt_Gender) <- "Gender"
+var_label(study.sample$pt_Gender) <- "Gender"
+var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure"
 
 # Create a table of sample characteristics
 sample.characteristics.table <- tbl_summary(study.sample,

### 2024-10-11 main.R
La till R

@@ -128,7 +128,12 @@ var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure"
 sample.characteristics.table <- tbl_summary(study.sample,
                                             by = ofi)
 
+# Display table
 sample.characteristics.table
+
+# Testing P
+add_p(sample.characteristics.table)
+
 ## Whatever you do next, maybe clean data?
 
 
@@ -136,3 +141,4 @@ sample.characteristics.table
 #recodar variabler - tidy verse
 #conituerlig av BE i tabellen o.s.v 
 #add p - gt summary 
+

### 2024-10-11 main.R
tog bort tomma linjer

@@ -141,4 +141,3 @@ add_p(sample.characteristics.table)
 #recodar variabler - tidy verse
 #conituerlig av BE i tabellen o.s.v 
 #add p - gt summary 
-

### 2024-10-13 project-plan.Rmd
la till stycke om ofi och trauma/medical gap

@@
 
-Trauma is a major global public health concern.
-It causes over four million death and affects hundreds of millions each year, among those younger populations are the most affected[@noauthor_injuries_nodate].
-Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall number but the younger demographic are also affected.
+Trauma is a major global public health concern. It causes over four million deaths and affects hundreds of millions each year
+#otydligt
+, among those younger populations are the most affected[@noauthor_injuries_nodate].
+Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall number but the younger and economically active demographic are also affected.
 In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old[@noauthor_arsrapport_2022].
@@
 The most anatomically common type of blunt trauma is blunt abdominal trauma (BAT), which is often caused by traffic accidents.
-When counting motor vehicle collision and pedestrian versus auto accidents together, they account for 75% of all blunt trauma[@isenhour_advances_2007].
+When counting motor vehicle collision and pedestrian versus auto accidents together, they account for 75% of all blunt trauma[@isenhour_advances_2007]
+#sweden or worldwide?
+.
 
@@
 
-A consequence of both types of trauma is possibility of significant bleeding, which in turn can result in hemorrhagic shock.
+A consequence of both types of trauma is the possibility of significant bleeding, which in turn can result in hemorrhagic shock.
 
@@
 One of such classification is the ATLS classification for major haemorrhage.
-ATLS stratifies patients into four classes, based on the patients: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output[@noauthor_atls_2018].
-However, this only meant to be used for hypovolemic shocks and not the other types.
-There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
+ATLS stratifies patients into four classes, based on the following patient factors: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output[@noauthor_atls_2018].
+However, this is only meant to be used for hypovolemic shocks and not the other types.
+There are other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2)
+#referens?
+.
 The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
@@
 
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24 hour of injury[@berry_shock_2015], as the midian time to death is 2 hours[@tisherman_detailed_2015].
-To address these and other challenges within trauma care and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008].
+To address this and other challenges within trauma care and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
 QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality[@world_health_organization_guidelines_2009].
@@
 This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[@juillard_establishing_2009][@hashmi_hospital-based_2013].
+
+## Shock and OFI
+Shock is the leading cause of preventable death within the first 24 hours of injury[@berry_shock_2015], with a median time to death from hemorrhagic shock of just 2 hours[@tisherman_detailed_2015].
+Despite this, the relationship between shock, its severity, and Opportunities for Improvement (OFI) remains largely unexamined.
+Most studies on OFI focus on trauma as a whole and do not explore shock in detail, for example, one study [@christie_how_2023] noted that 80% of patients had bleeding or hypotension but did not investigate further.
+However, studying shock more closely is crucial, as it is common, highly preventable and one of the most deadly.
+This knowledge would help clinicians better identify and avoid preventable mistakes in shock patients who need urgent care.
 
@@
 This selection is based on tenth edition, where SBP is defined as normal in class I to II and normal/below in class III and clearly below normal in class IV.
-These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
+These two studies by Eastridge et al and Oyetunji et al, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
 We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).

### 2024-10-13 thesis.Rmd
la till titel o namn

--- 2024-10-13 project-plan.Rmd
@@
 ---
-title: "Manuscript title here"
+title: "Opportunities for improvement in the care of adult trauma patients arriving in shock"
 subtitle: "Add a Subtitle if Needed"
-author: "Your name here"
+author: "Carl Skog"
 output: 
@@
 
-Trauma is a major global public health concern. It causes over four million deaths and affects hundreds of millions each year
-#otydligt
-, among those younger populations are the most affected[@noauthor_injuries_nodate].
+Trauma is a major global public health concern.
+It causes over four million deaths, among those younger populations are the most affected[@noauthor_injuries_nodate].
 Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall number but the younger and economically active demographic are also affected.

### 2024-10-15 main.R
gissade kön korrekt från början och ändrade nu till na

@@ -100,7 +100,7 @@ study.sample <- study.sample %>%
 # Converting 999 to unknown for pt_asa_preinjury
 study.sample <- study.sample %>%
   mutate(pt_asa_preinjury = case_when(
-    pt_asa_preinjury == 999 ~ "Unknown",
+    pt_asa_preinjury == 999 ~ NA_character_,
     TRUE ~ as.character(pt_asa_preinjury)
   ))
 

### 2024-10-15 main.R
ändrade stavning o la till enhet

@@ -117,12 +117,12 @@ study.sample <- study.sample |>
 var_label(study.sample$pt_age_yrs) <- "Age (Years)"
 var_label(study.sample$ofi) <- "Opportunities for improvement (Y/N)"
 var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
-var_label(study.sample$SBP_class) <- "Shock class classfied according to SBP"
+var_label(study.sample$SBP_class) <- "Shock class classified according to SBP"
 var_label(study.sample$BE_class) <- "Shock class classified according to BE"
 var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
 var_label(study.sample$ISS) <- "Injury Severity Score"
-var_label(study.sample$pt_Gender) <- "Gender"
-var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure"
+var_label(study.sample$pt_Gender) <- "Gender (M/F)"
+var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure (mmhg)"
 
 # Create a table of sample characteristics
 sample.characteristics.table <- tbl_summary(study.sample,

### 2024-10-15 manuscript.Rmd
la till kod för att kunna presentera tabell i dokument

@@
 output:
+  html_document: default
   word_document: default
-  html_document: default
 bibliography: bibliography.bib
@@
 
-```{r main, echo=FALSE}
+```{r main, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
 source("main.R") ## This "imports" the main script file of your project and run any code in it

### 2024-10-15 main.R

@@ -134,10 +134,5 @@ sample.characteristics.table
 # Testing P
 add_p(sample.characteristics.table)
 
-## Whatever you do next, maybe clean data?
-
-
-
-#recodar variabler - tidy verse
-#conituerlig av BE i tabellen o.s.v 
-#add p - gt summary 
+# Print
+print(sample.characteristics.table)

### 2024-10-15 thesis.Rmd
la till resultat delen mall från manuscrpt

@@
 
-# Project Update and Timeline
+# Results
 
-*This is part B) in the project plan [sv: projektredogörelse].*
+```{=html}
+<!--
 
-*Save the image generated from the Gantt chard spreadsheet into the
-assets folder, rename it to project-gantt.png*
+Participants
+------------
 
-![**The Gantt chart will be displayed here when you place it in the
-assets folder under "project-gantt.png"**](assets/project-gantt.png)
+(a) Report numbers of individuals at each stage of study—eg numbers
+potentially eligible, examined for eligibility, confirmed eligible,
+included in the study, completing follow-up, and analysed
 
-Doing right now: - Introduction - Methods
+(b) Give reasons for non-participation at each stage
 
-To do: - Coding to sort data - Coding for regression - Analysis - Results - Discussion -
-Conclusion and Abstract
+(c) Consider use of a flow diagram
 
-# Back-up Plan
+Descriptive data
+----------------
 
-After discussion with my supervisor, it seems like the part which have the highest possibility of taking too much time is coding, and therefore he and other cosupervisor will assist with coding if needed.
+(a) Give characteristics of study participants (eg demographic,
+clinical, social) and information on exposures and potential
+confounders
+
+(b) Indicate number of participants with missing data for each
+variable of interest
+
+(c) Cohort study — Summarise follow-up time (eg, average and total
+amount)
+
+Outcome data
+------------
+
+Cohort study — Report numbers of outcome events or summary measures
+over time
+
+Case-control study — Report numbers in each exposure category, or
+summary measures of exposure
+
+Cross-sectional study — Report numbers of outcome events or summary
+measures
+
+Main results
+------------
+
+(a) Give unadjusted estimates and, if applicable, confounder-adjusted
+estimates and their precision (eg, 95% confidence interval). Make
+clear which confounders were adjusted for and why they were included
+
+(b) Report category boundaries when continuous variables were
+categorized
+
+(c) If relevant, consider translating estimates of relative risk into
+absolute risk for a meaningful time period
+
+Other analyses 
+--------------
+
+Report other analyses done—eg analyses of subgroups and interactions,
+and sensitivity analyses 
+-->
+```
+You can include code in this document like this:
+
+```{r main, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
+source("main.R") ## This "imports" the main script file of your project and run any code in it
+```
+
+You can also embed plots:
+
+```{r plot, echo=FALSE}
+plot(pressure)
+```
+
+You can also mix text and code, so called inline code, like this: `r 2+5`.
+
+# Discussion
+
+```{=html}
+<!--
+
+Key results
+-----------
+Summarise key results with reference to study objectives
+
+Limitations
+-----------
+
+Discuss limitations of the study, taking into account sources of
+potential bias or imprecision.  Discuss both direction and magnitude
+of any potential bias
+
+Interpretation
+--------------
+
+Give a cautious overall interpretation of results considering
+objectives, limitations, multiplicity of analyses, results from
+similar studies, and other relevant evidence
+
+Generalisability
+----------------
+
+Discuss the generalisability (external validity) of the study results
+
+-->
+```
+# Conclusion
+
+<!-- Give a short conclusion, corresponding to the aim -->
 

### 2024-10-24 main.R
tog bort lite onödiga grejer

@@ -45,15 +45,10 @@ merged.data$ofi <- create_ofi(merged.data)
 study.data <- merged.data |> 
   select(pt_age_yrs, 
          pt_Gender, 
-         inj_mechanism, 
          pt_asa_preinjury, 
-         ed_gcs_sum,
          ed_sbp_value,
-         ed_rr_value,
          ed_be_art,
          ISS,
-         host_care_level,
-         res_survival,
          ofi,
          ed_sbp_rtscat,
          ed_inr,
@@ -163,19 +158,20 @@ study.sample <- study.sample %>%
 true_noNA <- na.omit(study.sample)
 
 # Binary logistic regression model
-log_reg <- glm(ofinum ~ pt_age_yrs + pt_Gender + ISS + ed_inr_numeric + pt_asa_preinjury + V3SBP_class + BE_class, 
-               family = binomial, 
-               data = true_noNA)
-
-log_reg <- glm(ofinum ~ ed_sbp_value, 
+log_reg <- glm(ofinum ~ pt_age_yrs + pt_Gender + ISS + ed_inr_numeric + pt_asa_preinjury + ed_sbp_value + BE_class, 
                family = binomial, 
                data = study.sample)
+
+#log_reg <- glm(ofinum ~ ed_sbp_value, 
+               #family = binomial, 
+               #data = study.sample)
 
 
 # Summary of the model
 summary(log_reg)
 
-plot(log_reg)
+# Generate predicted probabilities
+study.sample$predicted_prob <- predict(log_reg, type = "response")
 
 
 # Box tidwell test for linearity (add 1e-10? to avoid log(0))
@@ -191,13 +187,10 @@ summary(box_tidwell_sbp)
 
 # Remove unused variables. 
 study.sample <- study.sample |> 
-  select(-inj_mechanism, 
-         -ed_gcs_sum,
-         -ed_rr_value,
-         -ed_be_art,
-         -host_care_level,
-         -res_survival,
-         -ed_inr
+  select(-ed_be_art,
+         -ed_inr,
+         -log_ed_sbp_value,
+         -ofinum
          )
 
 # Label variables

### 2024-10-24 main.R
Tog bort allt onödigt skräp och la till det som kan användas till skräp snuttar
La till tabell även för reg

omvandlade resutlatet från log till odds

@@ -27,6 +27,7 @@ library(labelled)
 library(gtsummary)
 library(ggplot2)
 library(nnet)
+library(broom.helpers)
 
 ## Import data
 data <- import_data(test = TRUE)
@@ -88,15 +89,6 @@ study.sample <- study.sample %>%
 study.sample <- study.sample %>%
   mutate(ed_be_art_numeric = BEnum)
 
-# SBP shock classifiaction
-study.sample <- study.sample %>%
-  mutate(SBP_class = case_when(
-    ed_sbp_value < (90) ~ "Class 2",
-    ed_sbp_value >= (90) & ed_sbp_value < (110) ~ "Class 1",
-    is.na(ed_sbp_value) ~ NA_character_,        
-    TRUE ~ "no shock"
-  ))
-
 # Convert gender from numeric to categorical
 study.sample <- study.sample %>%
   mutate(pt_Gender = case_when(
@@ -112,25 +104,6 @@ study.sample <- study.sample %>%
     TRUE ~ as.character(pt_asa_preinjury)
   ))
 
-# Converting 999 to unknown for ed_sbp_rtscat
-study.sample <- study.sample %>%
-  mutate(ed_sbp_rtscat = case_when(
-    ed_sbp_rtscat == 999 ~ NA_character_,
-    TRUE ~ as.character(ed_sbp_rtscat)
-  ))
-
-# Reclassify those without ed_sbp_rtscat class
-study.sample <- study.sample %>%
-  mutate(ed_sbp_rtscat = case_when(
-    ed_sbp_rtscat != "999" & !is.na(ed_sbp_rtscat) ~ ed_sbp_rtscat,  
-    ed_sbp_value >89 ~ "4",
-    ed_sbp_value <= 89 & ed_sbp_value >= 76 ~ "3",                                   
-    ed_sbp_value <= 75 & ed_sbp_value >= 50 ~ "2",
-    ed_sbp_value <= 49 & ed_sbp_value >= 1 ~ "1",
-    ed_sbp_value >= 110 ~ "no shock",                               
-    TRUE ~ NA_character_                                              
-  ))
-
 # Converting INR to numeric
 ed_inr_numeric <- convert_number(study.sample$ed_inr)
 
@@ -138,58 +111,28 @@ ed_inr_numeric <- convert_number(study.sample$ed_inr)
 study.sample <- study.sample %>%
   mutate(ed_inr_numeric = ed_inr_numeric)
 
[…]

### 2024-10-25 main.R
ändrade lite ordningen

@@ -51,7 +51,6 @@ study.data <- merged.data |>
          ed_be_art,
          ISS,
          ofi,
-         ed_sbp_rtscat,
          ed_inr,
          ofi.categories.broad,
          ofi.categories.detailed
@@ -74,16 +73,6 @@ study.sample$ofinum <- ifelse(study.sample$ofi == "Yes", 1, 0)
 
 # Converting ed_be_art to numeric
 BEnum <- convert_number(study.sample$ed_be_art)
-
-# BE shock classification
-study.sample <- study.sample %>%
-  mutate(BE_class = case_when(
-    BEnum < (-10) ~ "Class 4 (severe)",
-    BEnum >= (-10) & BEnum < (-6) ~ "Class 3 (moderate)",
-    BEnum >= (-6) & BEnum < (-2) ~ "Class 2 (mild)",  
-    BEnum >= (-2) ~ "Class 1 (no shock)",   
-    is.na(BEnum) ~ NA_character_,
-  ))
 
 # Re-add the BE column as numeric to `study.sample`
 study.sample <- study.sample %>%
@@ -111,18 +100,28 @@ ed_inr_numeric <- convert_number(study.sample$ed_inr)
 study.sample <- study.sample %>%
   mutate(ed_inr_numeric = ed_inr_numeric)
 
+# BE shock classification
+study.sample <- study.sample %>%
+  mutate(BE_class = case_when(
+    BEnum < (-10) ~ "Class 4 (severe)",
+    BEnum >= (-10) & BEnum < (-6) ~ "Class 3 (moderate)",
+    BEnum >= (-6) & BEnum < (-2) ~ "Class 2 (mild)",  
+    BEnum >= (-2) ~ "Class 1 (no shock)",   
+    is.na(BEnum) ~ NA_character_,
+  ))
+
 # V4 SBP class
 study.sample <- study.sample %>%
   mutate(V4SBP_class = case_when(
-    ed_sbp_value < (90) ~ "Class 4 - severe",
-    ed_sbp_value >= 90 & ed_sbp_value < (100) ~"Class 3 - moderate",
-    ed_sbp_value >= 100 & ed_sbp_value < (110) ~ "Class 2 - mild",
+    ed_sbp_value < (90) ~ "Class 4 (severe)",
+    ed_sbp_value >= 90 & ed_sbp_value < (100) ~"Class 3 (moderate)",
[…]

### 2024-10-25 main.R
la till frågor till nästa gång

@@ -178,4 +178,5 @@ print(log_reg_sample.characteristics.tabel)
 
 
 # hantera missingdata - Hur hantera de 33%? kör listwise deletion och sedan diskutera det i diskussionen
-# tolka värdena och läsa på
+# lägga till noNA kanske för summary tabellen?
+# lägga till cirkel diagram för ofi broad och shock class?

### 2024-10-26 thesis.Rmd
Ändrade delen om shock classifikation

@@
 
-To differentiate the severity of trauma shock, we can divide it into different degrees.
-One of such classification is the ATLS classification for major haemorrhage.
-ATLS stratifies patients into four classes, based on the following patient factors: estimated blood loss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output[@noauthor_atls_2018].
-However, this is only meant to be used for hypovolemic shocks and not the other types.
-There are other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2)
-#referens?
-.
-The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
-ATLS in this regard, even though having its own shortcomings, still is one of the common classifications and part of the student course manual of the internationally accepted ATLS trauma management system[@guly_vital_2011][@noauthor_atls_2018].
+To assess the severity of trauma shock, patients can be categorized by stratifying symptoms and physiological parameters into degrees.
+Although no single international standard exists for such classifications, there is the Advanced Trauma Life Support (ATLS) system for major hemorrhage, which is part of the ATLS trauma management course curriculum[@noauthor_atls_2018].
+This system stratifies patients into four shock classes based on estimated blood loss, heart rate (HR), blood pressure (BP), pulse pressure, respiratory rate (RR), mental status, and urine output[@noauthor_atls_2018].
+However, studies have indicated that the ATLS classification has limited prognostic accuracy[@guly_vital_2011] and is often impractical for clinical use[@mutschler_critical_2013], which may explain the low uptake in practice[@mutschler_is_2015].
+
+Alternative scoring systems have been developed to address specific clinical needs, such as the Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion[@el-menyar_review_2019], and the National Early Warning Score 2 (NEWS2) for monitoring general patient deterioration [källa?].
+While none of these systems is explicitly designed to classify trauma shock across all shock types, each can, to some extent, aid in categorizing shock severity based on core physiological parameters.
+
+Ultimately, the shared foundation of these classifications lies in their reliance on vital signs and additional clinical indicators, underscoring the central role of physiological metrics in assessing shock severity across different settings and applications.
 
@@
 
-Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
+Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
 Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.

### 2024-10-30 main.R
Edit and comment on code

@@ -61,6 +61,8 @@ study.data <- merged.data |>
 study.sample <- study.data |>
   filter(!is.na(ofi))
 
+# Add code to remove patients younger than 15 and dead on arrival
+
 # Function for converting to numeric
 convert_number <- function(x) {
   x <- as.character(x)
@@ -122,6 +124,7 @@ study.sample <- study.sample %>%
   ))
 
 # Binary logistic regression model
+# Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
 log_reg <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + BE_class + V4SBP_class,
   family = binomial,
   data = study.sample
@@ -155,10 +158,11 @@ var_label(study.sample$V4SBP_class) <- "Shock classification - SBP"
 sample.characteristics.table <- tbl_summary(study.sample,
   by = ofi
 ) |>
+  add_overall() |>
   add_p()
 
 # Create a table of regression of sample
-log_reg_sample.characteristics.tabel <- tbl_regression(log_reg,
+log_reg_sample.characteristics.table <- tbl_regression(log_reg,
   exponentiate = TRUE,
   label = list(
     pt_age_yrs ~ "Age (Years)",
@@ -173,14 +177,16 @@ log_reg_sample.characteristics.tabel <- tbl_regression(log_reg,
 
 # Display tables
 sample.characteristics.table
-log_reg_sample.characteristics.tabel
+log_reg_sample.characteristics.table
 
 # Print
 print(sample.characteristics.table)
-print(log_reg_sample.characteristics.tabel)
+print(log_reg_sample.characteristics.table)
 
-
-
+# Create objects for descriptive data
+ofi <- paste0(sum(study.sample$ofi == "Yes"), " (", round(sum(study.sample$ofi == "Yes") / nrow(study.sample) * 100, 1), "%)")
+age <- inline_text(sample.characteristics.table, variable = pt_age_yrs, column = stat_0)
+gender <- inline_text(sample.characteristics.table, variable = pt_Gender, column = stat_0, level = "Male")
 
 # hantera missingdata - Hur hantera de 33%? kör listwise deletion och sedan diskutera det i diskussionen
 # lägga till noNA kanske för summary tabellen?

### 2024-10-30 manuscript.Rmd

@@
 ```
-You can cite document here like this [@exampleKey9999]. Open the file bibliography.bib to learn more.
-
 Trauma is a major global public health concern.
@@
 
+```{r main, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
+source("main.R") ## This "imports" the main script file of your project and run any code in it
+```
+
 ```{=html}
@@
 
+There were a total of `r nrow(merged.data)` patients in the trauma registry.
+Out of these, `r nrow(study.sample)` patients had been reviewed for the presence of opportunities for improvement.
+After exluding patients younger than 15 and/or were dead on arrival, `r nrow(study.sample)` patients remained.
+A total of X patients were finally excluded due to missing data, leaving us with Y patients for the final analysis.
+
 Descriptive data
@@
 amount)
+
+Sample characteristics are presented in Table \@ref(tab:sample-characteristics).
+A total of `r ofi` patients had at least one opportunity for improvement.
+The median age was `r age`.
+Most patients were male `r gender`.
+
+```{r sample-characteristics, echo=FALSE}
+sample.characteristics.table
+```
 
@@
 You can include code in this document like this:
-
-```{r main, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
-source("main.R") ## This "imports" the main script file of your project and run any code in it
-```
 

### 2024-10-30 main.R
la till både ta bort fel ålder och död la till ny/separata log reg för BE och SBP Behöll ta bort senare istället för i början, pga behöver värdena i början för kalkylering, men använder senare antigen den numeriska eller klassade varianten istället och därför tar bort de senare, då jag har behov av original datan i början Behöll BE num, då SBP var kvar. Kanske tanken är att jag ska ta bort båda?

@@ -54,14 +54,16 @@ study.data <- merged.data |>
     ofi,
     ed_inr,
     ofi.categories.broad,
-    ofi.categories.detailed
+    ofi.categories.detailed,
+    Deceased,
   )
 
 # Exclude patients who were not reviewed for the presence of OFI
 study.sample <- study.data |>
-  filter(!is.na(ofi))
-
-# Add code to remove patients younger than 15 and dead on arrival
+  filter(!is.na(ofi),
+         Deceased == "False",
+         pt_age_yrs >= 15
+         )
 
 # Function for converting to numeric
 convert_number <- function(x) {
@@ -77,9 +79,9 @@ study.sample$ofinum <- ifelse(study.sample$ofi == "Yes", 1, 0)
 # Converting ed_be_art to numeric
 BEnum <- convert_number(study.sample$ed_be_art)
 
-# Re-add the BE column as numeric to `study.sample`
-# study.sample <- study.sample %>%
-#  mutate(ed_be_art_numeric = BEnum)
+#Re-add the BE column as numeric to `study.sample`
+study.sample <- study.sample %>%
+  mutate(ed_be_art_numeric = BEnum)
 
 # Convert gender from numeric to categorical
 study.sample <- study.sample %>%
@@ -123,15 +125,18 @@ study.sample <- study.sample %>%
     TRUE ~ "Class 1 (no shock)"
   ))
 
-# Binary logistic regression model
+# Binary logistic regression model - BE
 # Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
-log_reg <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + BE_class + V4SBP_class,
+log_regBE <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + BE_class,
   family = binomial,
   data = study.sample
 )
 
-# Summary of the model
-summary(log_reg)
+# Binary logistic regression model - SBP
+log_regSBP <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + V4SBP_class,
+               family = binomial,
+               data = study.sample
+)
 
 # Remove unused variables.
[…]

### 2024-10-30 thesis.Rmd
Ändrade två grammatik delar

@@
 It causes over four million deaths, among those younger populations are the most affected[@noauthor_injuries_nodate].
-Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall number but the younger and economically active demographic are also affected.
+Globally low and middle income countries are disproportionately affected, while high income countries like Sweden have lower overall numbers but the younger and economically active demographic are also affected.
 In Sweden between 2018 and 2021, the most common age demographic for trauma is the working aged population, between 18 and 64 years old[@noauthor_arsrapport_2022].
@@
 
-Trauma are generally divided into two groups, blunt and penetrating injury.
+Trauma is generally divided into two groups, blunt and penetrating injury.
 The most common type in Sweden is blunt injury, constituting around 90% of all trauma.

### 2024-10-30 main.R
tog bort gamla frågor

@@ -207,7 +207,3 @@ print(log_regSBP_sample.characteristics.table)
 ofi <- paste0(sum(study.sample$ofi == "Yes"), " (", round(sum(study.sample$ofi == "Yes") / nrow(study.sample) * 100, 1), "%)")
 age <- inline_text(sample.characteristics.table, variable = pt_age_yrs, column = stat_0)
 gender <- inline_text(sample.characteristics.table, variable = pt_Gender, column = stat_0, level = "Male")
-
-# hantera missingdata - Hur hantera de 33%? kör listwise deletion och sedan diskutera det i diskussionen
-# lägga till noNA kanske för summary tabellen?
-# lägga till cirkel diagram för ofi broad och shock class?

### 2024-11-01 manuscript.Rmd
La till ATLS tabell

@@
 ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
+
+| **Parameter**                     | **Class I** | **Class II** (Mild) | **Class III** (Moderate) | **Class IV** (Severe)        |
+|-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
+| Approximate blood loss            | <15%        | 15–30%              | 31–40%                    | >40%                          |
+| Heart rate                        | ↔           | ↔/↑                 | ↑                         | ↑/↑↑                          |
+| Blood pressure                    | ↔           | ↔                   | ↔/↓                       | ↓                             |
+| Pulse pressure                    | ↔           | ↓                   | ↓                         | ↓                             |
+| Respiratory rate                  | ↔           | ↔                   | ↔/↑                       | ↑                             |
+| Urine output                      | ↔           | ↔                   | ↓                         | ↓↓                            |
+| Glasgow Coma Scale score          | ↔           | ↔                   | ↓                         | ↓                             |
+| Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L     | –6 to –10 mEq/L          | –10 mEq/L or less             |
+| Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
 

### 2024-11-01 main.R

@@ -167,7 +167,7 @@ sample.characteristics.table <- tbl_summary(study.sample,
   add_overall() |>
   add_p()
 
-# Create a table of regression of sample
+# Create a table of regression of sample - BE
 log_regBE_sample.characteristics.table <- tbl_regression(log_regBE,
   exponentiate = TRUE,
   label = list(
@@ -180,7 +180,7 @@ log_regBE_sample.characteristics.table <- tbl_regression(log_regBE,
   )
 )
 
-# Create a table of regression of sample
+# Create a table of regression of sample - SBP
 log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
   exponentiate = TRUE,
   label = list(

### 2024-11-01 thesis.Rmd
ändrade till html

@@
 output:
+  html_document: default
   word_document: default

### 2024-11-01 thesis.Rmd
tabell för ATLS och min approximation

@@
 
+**ATLS tenth edition**
+
+| **Parameter**                     | **Class I** | **Class II** (Mild) | **Class III** (Moderate) | **Class IV** (Severe)        |
+|-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
+| Approximate blood loss            | <15%        | 15–30%              | 31–40%                    | >40%                          |
+| Heart rate                        | ↔           | ↔/↑                 | ↑                         | ↑/↑↑                          |
+| Blood pressure                    | ↔           | ↔                   | ↔/↓                       | ↓                             |
+| Pulse pressure                    | ↔           | ↓                   | ↓                         | ↓                             |
+| Respiratory rate                  | ↔           | ↔                   | ↔/↑                       | ↑                             |
+| Urine output                      | ↔           | ↔                   | ↓                         | ↓↓                            |
+| Glasgow Coma Scale score          | ↔           | ↔                   | ↓                         | ↓                             |
+| Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L     | –6 to –10 mEq/L          | –10 mEq/L or less             |
+| Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
+
 Alternative scoring systems have been developed to address specific clinical needs, such as the Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion[@el-menyar_review_2019], and the National Early Warning Score 2 (NEWS2) for monitoring general patient deterioration [källa?].
@@
 Ultimately, the shared foundation of these classifications lies in their reliance on vital signs and additional clinical indicators, underscoring the central role of physiological metrics in assessing shock severity across different settings and applications.
-
-(Infoga bild på ATLS tenth edition)
 
@@
 ## Quantitative variables
+
+**ATLS tenth edition - Systolic blood pressure and Base excess (BD)**
+
+| **Parameter**                     | **Class I** | **Class II** (Mild)  | **Class III** (Moderate)  | **Class IV** (Severe)         |
+|-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
+| Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L     | –6 to –10 mEq/L           | –10 mEq/L or less             |
+| Blood pressure                    | ↔           | ↔                    | ↔/↓                       | ↓                             |
+| **Parameter - numerical approximated SBP**                                                                                         |
+| Blood pressure                    | >110 mmhg   | 109-100 mmhg         | 99-90 mmhg                | <90  mmhg                     |
 

### 2024-11-01 thesis.Rmd
ändrade ltie i format

@@
 
-| **Parameter**                     | **Class I** | **Class II** (Mild)  | **Class III** (Moderate)  | **Class IV** (Severe)         |
+| **Parameter - original ATLS 10th edition**| **Class I** | **Class II** (Mild)  | **Class III** (Moderate)  | **Class IV** (Severe) |
 |-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
 | Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L     | –6 to –10 mEq/L           | –10 mEq/L or less             |
-| Blood pressure                    | ↔           | ↔                    | ↔/↓                       | ↓                             |
+| Systolic Blood pressure           | ↔           | ↔                    | ↔/↓                       | ↓                             |
 | **Parameter - numerical approximated SBP**                                                                                         |
-| Blood pressure                    | >110 mmhg   | 109-100 mmhg         | 99-90 mmhg                | <90  mmhg                     |
+| Systolic Blood pressure           | >110 mmhg   | 109-100 mmhg         | 99-90 mmhg                | <90  mmhg                     |
 

### 2024-11-02 main.R
lagt till plot får inte till en förklarande rutan utanför diagram området

@@ -108,21 +108,21 @@ study.sample <- study.sample %>%
 # BE shock classification
 study.sample <- study.sample %>%
   mutate(BE_class = case_when(
-    BEnum < (-10) ~ "Class 4 (severe)",
-    BEnum >= (-10) & BEnum < (-6) ~ "Class 3 (moderate)",
-    BEnum >= (-6) & BEnum < (-2) ~ "Class 2 (mild)",
-    BEnum >= (-2) ~ "Class 1 (no shock)",
+    BEnum < (-10) ~ "Class 4",
+    BEnum >= (-10) & BEnum < (-6) ~ "Class 3",
+    BEnum >= (-6) & BEnum < (-2) ~ "Class 2",
+    BEnum >= (-2) ~ "Class 1",
     is.na(BEnum) ~ NA_character_,
   ))
 
 # V4 SBP class
 study.sample <- study.sample %>%
   mutate(V4SBP_class = case_when(
-    ed_sbp_value < (90) ~ "Class 4 (severe)",
-    ed_sbp_value >= 90 & ed_sbp_value < (100) ~ "Class 3 (moderate)",
-    ed_sbp_value >= 100 & ed_sbp_value < (110) ~ "Class 2 (mild)",
+    ed_sbp_value < (90) ~ "Class 4",
+    ed_sbp_value >= 90 & ed_sbp_value < (100) ~ "Class 3",
+    ed_sbp_value >= 100 & ed_sbp_value < (110) ~ "Class 2",
     is.na(ed_sbp_value) ~ NA_character_,
-    TRUE ~ "Class 1 (no shock)"
+    TRUE ~ "Class 1"
   ))
 
 # Binary logistic regression model - BE
@@ -192,6 +192,22 @@ log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
     V4SBP_class ~ "Shock classification - SBP"
   )
 )
+
+# Create a contingency table of counts
+ofi_table <- table(study.sample$BE_class, study.sample$ofi.categories.broad)
+
+# Convert the table into proportions by row (BE_class)
+ofi_prop <- prop.table(ofi_table, margin = 1)
+
+# Plotting
+barplot(
+  t(ofi_prop),  
+  beside = TRUE,  
+  col = rainbow(ncol(ofi_prop)),  
+  main = "Distribution of OFI Categories by BE Shock Class",
+  xlab = "Shock class - BE",
+  ylab = "Proportion (%)",
+)  
 
 # Display tables
 sample.characteristics.table

### 2024-11-03 main.R
ggplot2 version av diagrammet

@@ -193,13 +193,11 @@ log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
   )
 )
 
-# Create a contingency table of counts
+# plot
 ofi_table <- table(study.sample$BE_class, study.sample$ofi.categories.broad)
 
-# Convert the table into proportions by row (BE_class)
 ofi_prop <- prop.table(ofi_table, margin = 1)
 
-# Plotting
 barplot(
   t(ofi_prop),  
   beside = TRUE,  
@@ -208,6 +206,30 @@ barplot(
   xlab = "Shock class - BE",
   ylab = "Proportion (%)",
 )  
+
+# ggplot2
+ofi_counts <- study.sample %>%
+  group_by(BE_class, ofi.categories.broad) %>%
+  summarize(count = n()) %>%
+  ungroup() %>%
+  group_by(BE_class) %>%
+  mutate(percent = count / sum(count) * 100)
+
+p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.broad)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(
+    title = "Distribution of OFI Broad Categories by BE Shock Class",
+    x = "BE Shock Class",
+    y = "Percentage of OFI Broad Categories"
+  ) +
+  theme_minimal() +
+  theme(
+    legend.title = element_text(size = 10),
+    legend.position = "bottom"
+  ) +
+  scale_fill_brewer(palette = "Set1", name = "OFI Categories")
+
+print(p)
 
 # Display tables
 sample.characteristics.table

### 2024-11-03 thesis.Rmd
lagt till plot får inte till en förklarande rutan utanför diagram området
fattade och fixade och slog ihop ställen med flera källor

@@
 To assess the severity of trauma shock, patients can be categorized by stratifying symptoms and physiological parameters into degrees.
-Although no single international standard exists for such classifications, there is the Advanced Trauma Life Support (ATLS) system for major hemorrhage, which is part of the ATLS trauma management course curriculum[@noauthor_atls_2018].
+Although no single international standard exists for such classifications, there is Advanced Trauma Life Support (ATLS) system for major hemorrhage, which is part of the  internationally recognised ATLS trauma management course curriculum[@noauthor_atls_2018].
 This system stratifies patients into four shock classes based on estimated blood loss, heart rate (HR), blood pressure (BP), pulse pressure, respiratory rate (RR), mental status, and urine output[@noauthor_atls_2018].
@@
 
-**ATLS tenth edition**
+**ATLS 10th edition**
 
@@
 
-Ultimately, the shared foundation of these classifications lies in their reliance on vital signs and additional clinical indicators, underscoring the central role of physiological metrics in assessing shock severity across different settings and applications.
+Ultimately, the shared foundation of these classifications lies in their reliance on vital signs and additional clinical indicators, underscoring the importance of physiological metrics in assessing shock severity across different clinical settings.
 
@@
 
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010][@teixeira_preventable_2007][@ivatury_patient_2008].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010;@teixeira_preventable_2007;@ivatury_patient_2008].
 To address this and other challenges within trauma care and reduce preventable errors, different types of Quality Improvement (QI) initiatives were initiated.
 QI focuses on systematically evaluating healthcare processes and patient outcomes, aiming to reduce morbidity and mortality[@world_health_organization_guidelines_2009].
-By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally[@juillard_establishing_2009][@hashmi_hospital-based_2013].
+By implementing formal quality measures and conducting collaborative case reviews, QI programs have strengthened trauma systems and improved patient outcomes in various settings globally[@juillard_establishing_2009;@hashmi_hospital-based_2013].
 
@@
 There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality (M&M) reviews of patient cases[@world_health_organization_guidelines_2009].
-In order for the review content to be as comprehensive as possible and taking in a system oriented approach, the Donabedian quality of care framework can be used[@donabedian_quality_1988][@world_health_organization_guidelines_2009].
+In order for the review content to be as comprehensive as possible and taking in a system oriented approach, the Donabedian quality of care framework can be used[@donabedian_quality_1988;@world_health_organization_guidelines_2009].
 It is based on the three factors of structure, process and outcome.
 By following this framework, healthcare providers will be able to systematically find OFIs as part of QI.
-This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[@juillard_establishing_2009][@hashmi_hospital-based_2013].
+This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[@juillard_establishing_2009;@hashmi_hospital-based_2013].
 
@@
 
-Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
+Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006;@ogura_modified_2016;@hanlin_prehospital_2024].
 Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
@@
 This selection is based on tenth edition, where SBP is defined as normal in class I to II and normal/below in class III and clearly below normal in class IV.
-These two studies by Eastridge et al and Oyetunji et al, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
+These two studies by Eastridge et al and Oyetunji et al, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007;@oyetunji_redefining_2011], therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
 We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).

### 2024-11-04 main.R
tog bort plot

@@ -193,20 +193,6 @@ log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
   )
 )
 
-# plot
-ofi_table <- table(study.sample$BE_class, study.sample$ofi.categories.broad)
-
-ofi_prop <- prop.table(ofi_table, margin = 1)
-
-barplot(
-  t(ofi_prop),  
-  beside = TRUE,  
-  col = rainbow(ncol(ofi_prop)),  
-  main = "Distribution of OFI Categories by BE Shock Class",
-  xlab = "Shock class - BE",
-  ylab = "Proportion (%)",
-)  
-
 # ggplot2
 ofi_counts <- study.sample %>%
   group_by(BE_class, ofi.categories.broad) %>%

### 2024-11-04 main.R
la till för SBP

@@ -217,6 +217,30 @@ p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.bro
 
 print(p)
 
+# ggplot2 SBP
+ofi_counts <- study.sample %>%
+  group_by(V4SBP_class, ofi.categories.broad) %>%
+  summarize(count = n()) %>%
+  ungroup() %>%
+  group_by(V4SBP_class) %>%
+  mutate(percent = count / sum(count) * 100)
+
+pSBP <- ggplot(ofi_counts, aes(x = V4SBP_class, y = percent, fill = ofi.categories.broad)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  labs(
+    title = "Distribution of OFI Broad Categories by SBP Shock Class",
+    x = "SBP Shock Class",
+    y = "Percentage of OFI Broad Categories"
+  ) +
+  theme_minimal() +
+  theme(
+    legend.title = element_text(size = 10),
+    legend.position = "bottom"
+  ) +
+  scale_fill_brewer(palette = "Set2", name = "OFI Categories")
+
+print(pSBP)
+
 # Display tables
 sample.characteristics.table
 log_regBE_sample.characteristics.table

### 2024-11-04 manuscript.Rmd
La till kod i main som räknar det log reg använder och sedan la till det i theisis

@@
 # Results
-
 ```{r main, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

### 2024-11-04 thesis.Rmd
La till kod i main som räknar det log reg använder och sedan la till det i theisis

@@
 
+There were a total of `r merged` patients in the trauma registry.
+After excluding the patients younger than 15 and/or were dead on arrival, there were `r excluded` patients left.
+Out of those, `r sample` patients had been reviewed for the presence of OFI.
+A total of `r sample - log_reg_countBE` patients were excluded for the BE classification group due to missing data, resulting in `r log_reg_countBE` patients for the final analysis.
+For the SBP group, missing data accounted for exclusion of `r sample - log_reg_countSBP` patients, leaving a sum of `r log_reg_countSBP` patients for the final analysis.
+
 You can also embed plots:

### 2024-11-04 main.R

@@ -58,6 +58,12 @@ study.data <- merged.data |>
     Deceased,
   )
 
+# Those excluded
+excluded <- study.data |>
+  filter(Deceased == "False",
+         pt_age_yrs >= 15
+  )
+
 # Exclude patients who were not reviewed for the presence of OFI
 study.sample <- study.data |>
   filter(!is.na(ofi),
@@ -251,7 +257,35 @@ print(sample.characteristics.table)
 print(log_regBE_sample.characteristics.table)
 print(log_regSBP_sample.characteristics.table)
 
+# Counting log reg BE
+log_reg_dataBE <- study.sample %>% 
+  filter(!is.na(pt_age_yrs) & 
+           !is.na(pt_Gender) & 
+           !is.na(ISS) & 
+           !is.na(ed_inr_numeric) & 
+           !is.na(pt_asa_preinjury) & 
+           !is.na(BE_class))
+
+# Total rows used in the logistic regression
+log_reg_countBE <- nrow(log_reg_dataBE)
+
+# Counting log reg SBP
+log_reg_dataSBP <- study.sample %>% 
+  filter(!is.na(pt_age_yrs) & 
+           !is.na(pt_Gender) & 
+           !is.na(ISS) & 
+           !is.na(ed_inr_numeric) & 
+           !is.na(pt_asa_preinjury) & 
+           !is.na(V4SBP_class))
+
+# Total rows used in the logistic regression
+log_reg_countSBP <- nrow(log_reg_dataSBP)
+
 # Create objects for descriptive data
 ofi <- paste0(sum(study.sample$ofi == "Yes"), " (", round(sum(study.sample$ofi == "Yes") / nrow(study.sample) * 100, 1), "%)")
 age <- inline_text(sample.characteristics.table, variable = pt_age_yrs, column = stat_0)
-gender <- inline_text(sample.characteristics.table, variable = pt_Gender, column = stat_0, level = "Male")
+male <- inline_text(sample.characteristics.table, variable = pt_Gender, column = stat_0, level = "Male")
+merged <- nrow(merged.data)
+sample <- nrow(study.sample)
+excluded <- nrow(excluded)
+

### 2024-11-05 main.R
unadjusted log reg

@@ -144,6 +144,18 @@ log_regSBP <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_nu
                data = study.sample
 )
 
+# Binary logistic regression model unadjusted - BE
+log_regBEun <- glm(ofinum ~ BE_class,
+                 family = binomial,
+                 data = study.sample
+)
+
+# Binary logistic regression model unadjusted - SBP
+log_regSBPun <- glm(ofinum ~ V4SBP_class,
+                  family = binomial,
+                  data = study.sample
+)
+
 # Remove unused variables.
 # I suggest removing them from the lines 40-52 where the study data is created instead
 study.sample <- study.sample |>
@@ -199,6 +211,22 @@ log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
   )
 )
 
+# Create a table of regression of sample unadjusted - BE
+log_regBE_sample.characteristics.table_unadjusted <- tbl_regression(log_regBEun,
+                                                         exponentiate = TRUE,
+                                                         label = list(
+                                                           BE_class ~ "Shock classification - BE"
+                                                         )
+)
+
+# Create a table of regression of sample unadjusted - SBP
+log_regSBP_sample.characteristics.table_unadjusted <- tbl_regression(log_regSBPun,
+                                                          exponentiate = TRUE,
+                                                          label = list(
+                                                            V4SBP_class ~ "Shock classification - SBP"
+                                                          )
+)
+
 # ggplot2
 ofi_counts <- study.sample %>%
   group_by(BE_class, ofi.categories.broad) %>%
@@ -251,6 +279,8 @@ print(pSBP)
 sample.characteristics.table
 log_regBE_sample.characteristics.table
 log_regSBP_sample.characteristics.table
+log_regSBP_sample.characteristics.table_unadjusted
+log_regSBP_sample.characteristics.table_unadjusted
 
 # Print
 print(sample.characteristics.table)

### 2024-11-05 main.R
ändrade plats till innan radering av variabler för räkning av reg antal

@@ -156,6 +156,30 @@ log_regSBPun <- glm(ofinum ~ V4SBP_class,
                   data = study.sample
 )
 
+# Counting log reg BE
+log_reg_dataBE <- study.sample %>% 
+  filter(!is.na(pt_age_yrs) & 
+           !is.na(pt_Gender) & 
+           !is.na(ISS) & 
+           !is.na(ed_inr_numeric) & 
+           !is.na(pt_asa_preinjury) & 
+           !is.na(BE_class))
+
+# Total rows used in the logistic regression
+log_reg_countBE <- nrow(log_reg_dataBE)
+
+# Counting log reg SBP
+log_reg_dataSBP <- study.sample %>% 
+  filter(!is.na(pt_age_yrs) & 
+           !is.na(pt_Gender) & 
+           !is.na(ISS) & 
+           !is.na(ed_inr_numeric) & 
+           !is.na(pt_asa_preinjury) & 
+           !is.na(V4SBP_class))
+
+# Total rows used in the logistic regression
+log_reg_countSBP <- nrow(log_reg_dataSBP)
+
 # Remove unused variables.
 # I suggest removing them from the lines 40-52 where the study data is created instead
 study.sample <- study.sample |>
@@ -275,41 +299,12 @@ pSBP <- ggplot(ofi_counts, aes(x = V4SBP_class, y = percent, fill = ofi.categori
 
 print(pSBP)
 
-# Display tables
-sample.characteristics.table
-log_regBE_sample.characteristics.table
-log_regSBP_sample.characteristics.table
-log_regSBP_sample.characteristics.table_unadjusted
-log_regSBP_sample.characteristics.table_unadjusted
-
 # Print
 print(sample.characteristics.table)
 print(log_regBE_sample.characteristics.table)
 print(log_regSBP_sample.characteristics.table)
-
-# Counting log reg BE
-log_reg_dataBE <- study.sample %>% 
-  filter(!is.na(pt_age_yrs) & 
-           !is.na(pt_Gender) & 
-           !is.na(ISS) & 
-           !is.na(ed_inr_numeric) & 
-           !is.na(pt_asa_preinjury) & 
-           !is.na(BE_class))
-
-# Total rows used in the logistic regression
-log_reg_countBE <- nrow(log_reg_dataBE)
-
-# Counting log reg SBP
-log_reg_dataSBP <- study.sample %>% 
-  filter(!is.na(pt_age_yrs) & 
[…]

### 2024-11-06 thesis.Rmd
la till, men fråga om dubbel study sample

@@
 
+## Participants
 There were a total of `r merged` patients in the trauma registry.
@@
 For the SBP group, missing data accounted for exclusion of `r sample - log_reg_countSBP` patients, leaving a sum of `r log_reg_countSBP` patients for the final analysis.
+
+## Descriptive data
+The study demographic median age is ... and the most common gender is male, `r male`.
+
+The variable with most missing data is ... which lack the data for ... patients.
+
+### BE classification
+For the outcome OFI, out of the `r sample` patients reviewed, `r ofi` patients had a positive OFI. Clinically, the
+
+### SBP classification
+
+Descriptive data
+----------------
+
+(a) Give characteristics of study participants (eg demographic,
+clinical, social) and information on exposures and potential
+confounders
+
+(b) Indicate number of participants with missing data for each
+variable of interest
+
+(c) Cohort study — Summarise follow-up time (eg, average and total
+amount)
+
+Outcome data
+------------
+
+Cohort study — Report numbers of outcome events or summary measures
+over time
+
+Case-control study — Report numbers in each exposure category, or
+summary measures of exposure
+
+Cross-sectional study — Report numbers of outcome events or summary
+measures
+
+Main results
+------------
+
+(a) Give unadjusted estimates and, if applicable, confounder-adjusted
+estimates and their precision (eg, 95% confidence interval). Make
+clear which confounders were adjusted for and why they were included
+
+(b) Report category boundaries when continuous variables were
+categorized
+
+(c) If relevant, consider translating estimates of relative risk into
+absolute risk for a meaningful time period
+
+Other analyses 
+--------------
+
+Report other analyses done—eg analyses of subgroups and interactions,
+and sensitivity analyses 
+-->
+
+Table 3 present the sample characteristics.
+
+Table 4 present
 

### 2024-11-06 main.R

@@ -156,6 +156,27 @@ log_regSBPun <- glm(ofinum ~ V4SBP_class,
                   data = study.sample
 )
 
+# Remove unused variables.
+# I suggest removing them from the lines 40-52 where the study data is created instead
+study.sample <- study.sample |>
+  select(
+    -ed_be_art,
+    -ed_inr,
+    -ofinum,
+    -Deceased
+  )
+# Label variables
+var_label(study.sample$pt_age_yrs) <- "Age (Years)"
+var_label(study.sample$ofi) <- "Opportunities for improvement (Y/N)"
+var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
+var_label(study.sample$BE_class) <- "Shock classification - BE"
+var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
+var_label(study.sample$ISS) <- "Injury Severity Score"
+var_label(study.sample$pt_Gender) <- "Gender (M/F)"
+var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure (mmhg)"
+var_label(study.sample$ed_inr_numeric) <- "INR"
+var_label(study.sample$V4SBP_class) <- "Shock classification - SBP"
+
 # Counting log reg BE
 log_reg_dataBE <- study.sample %>% 
   filter(!is.na(pt_age_yrs) & 
@@ -180,31 +201,24 @@ log_reg_dataSBP <- study.sample %>%
 # Total rows used in the logistic regression
 log_reg_countSBP <- nrow(log_reg_dataSBP)
 
-# Remove unused variables.
-# I suggest removing them from the lines 40-52 where the study data is created instead
-study.sample <- study.sample |>
-  select(
-    -ed_be_art,
-    -ed_inr,
-    -ofinum,
-    -Deceased
-  )
-
-# Label variables
-var_label(study.sample$pt_age_yrs) <- "Age (Years)"
-var_label(study.sample$ofi) <- "Opportunities for improvement (Y/N)"
-var_label(study.sample$ed_be_art_numeric) <- "Base Excess (BE)"
-var_label(study.sample$BE_class) <- "Shock classification - BE"
-var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
-var_label(study.sample$ISS) <- "Injury Severity Score"
-var_label(study.sample$pt_Gender) <- "Gender (M/F)"
[…]

### 2024-11-06 main.R
la till detailed variant av ofi sub

@@ -176,6 +176,8 @@ var_label(study.sample$pt_Gender) <- "Gender (M/F)"
 var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure (mmhg)"
 var_label(study.sample$ed_inr_numeric) <- "INR"
 var_label(study.sample$V4SBP_class) <- "Shock classification - SBP"
+var_label(study.sample$ofi.categories.broad) <- "OFI categories broad"
+var_label(study.sample$ofi.categories.detailed) <- "OFI categories detailed"
 
 # Counting log reg BE
 log_reg_dataBE <- study.sample %>% 
@@ -267,13 +269,13 @@ log_regSBP_sample.characteristics.table_unadjusted <- tbl_regression(log_regSBPu
 
 # ggplot2
 ofi_counts <- study.sample %>%
-  group_by(BE_class, ofi.categories.broad) %>%
+  group_by(BE_class, ofi.categories.detailed) %>%
   summarize(count = n()) %>%
   ungroup() %>%
   group_by(BE_class) %>%
   mutate(percent = count / sum(count) * 100)
 
-p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.broad)) +
+p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.detailed)) +
   geom_bar(stat = "identity", position = "dodge") +
   labs(
     title = "Distribution of OFI Broad Categories by BE Shock Class",
@@ -285,7 +287,7 @@ p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.bro
     legend.title = element_text(size = 10),
     legend.position = "bottom"
   ) +
-  scale_fill_brewer(palette = "Set1", name = "OFI Categories")
+  scale_fill_brewer(palette = "Set3", name = "OFI Categories")
 
 print(p)
 
@@ -309,7 +311,7 @@ pSBP <- ggplot(ofi_counts, aes(x = V4SBP_class, y = percent, fill = ofi.categori
     legend.title = element_text(size = 10),
     legend.position = "bottom"
   ) +
-  scale_fill_brewer(palette = "Set2", name = "OFI Categories")
+  scale_fill_brewer(palette = "Set1", name = "OFI Categories")
 
 print(pSBP)
 

### 2024-11-08 thesis.Rmd
la till detailed variant av ofi sub
flyttat över den nya ändrade metoden

@@
 Most studies on OFI focus on trauma as a whole and do not explore shock in detail, for example, one study [@christie_how_2023] noted that 80% of patients had bleeding or hypotension but did not investigate further.
-However, studying shock more closely is crucial, as it is common, highly preventable and one of the most deadly.
+However, studying shock more closely is crucial, as it is common, highly preventable, and one of the most deadly.
 This knowledge would help clinicians better identify and avoid preventable mistakes in shock patients who need urgent care.
@@
 
-The trauma registry includes patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
+The trauma registry includes patients treated at Karolinska University Hospital in Solna, which treats all major trauma in the greater metropolitan area of Stockholm.
+We included patients registered between 2014 and 2023.
 The trauma care quality database is a subset of the trauma registry and includes patients selected for review.
-The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.
 
@@
 
-We included all patients in trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.
-
-The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or a already admitted patient with an Injury Severity Score (ISS) greater than nine.
+The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or an already admitted patient with an Injury Severity Score (ISS) greater than nine.
 Every trauma patient (registered in the trauma registry) at Karolinska University Hospital is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
@@
 The identified OFIs are then categorized into broader areas.
-The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields.
 The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
+
+We included all patients in the trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.
 
@@
 The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the trauma care quality registry.
-An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
-
-To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
-Heart rate(HR), systolic blood pressure(SBP) and respiratory rate(RR).
-Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.
-
-Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006;@ogura_modified_2016;@hanlin_prehospital_2024].
-Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
-For HR and RR we will be using the numbers from the ninth edition of the classification.
-The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg.
-This selection is based on tenth edition, where SBP is defined as normal in class I to II and normal/below in class III and clearly below normal in class IV.
-These two studies by Eastridge et al and Oyetunji et al, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007;@oyetunji_redefining_2011], therefore we choose class III as 110-90 mmhg and class IV below 90mmhg.
-We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).
-Urine output and pulse pressure were also neglected, due to lack of data.
-
-To classify, all of the parameters in one class needs to be fulfilled.
-In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
-At same time ATLS manual states one of the common pitfalls is “Diagnosis of shock can be missed when only a single parameter is used.” therefore to improve sensitivity we decided to use the worst parameter for classification[@noauthor_atls_2018].
-
-We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage.
-There are three reasons.
-The first one is ATLS tenth edition lack numerical values for vital parameters.
-The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
-The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
-The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).
+An OFI can be various types of errors, which are categorized into: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
+The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure(SBP) and once with base excess (BE), regardless of their cause of shock.
+Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications.
+We included five other metrics to be adjusted for, due to potential for confounding.
+Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous as they were directly from the register.
 
 ## Bias
-
-The trauma care quality database is build up in a way where every error can be classified into a certain type of OFI, however for paitent that died there is only the category “preventable death”.
-Therefore, survivorship bias may occur as we dont have the detail which type of OFI actually occured.
 
@@
 ## Quantitative variables
+
+The BE parameter was divided into four classes with these cutoffs: Class I (above -2), Class II (-2 to -6), Class III (-6 to -10), and Class IV (below -10).
+Similarly, SBP was divided into four classes: Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and Class IV (below 90 mmHg).
+This classification follows the tenth edition ATLS hemorrhage classification, which does not specify exact values for any parameter other than BE.
+To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.
+and Oyetunji et al., who redefined hypotension as 110 mmHg respective 90 mmHg dependent on age[@eastridge_hypotension_2007; @oyetunji_redefining_2011].
+Therefore, we used 110 mmHg as the lower limit for (Class I: no shock) - (ATLS: normal SBP) and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
+This also aligns with the classification done by Mutschler et al.[@mutschler_critical_2013].
 
@@
 
+Please note that ATLS classification defines class I for the BE parameter as (0 to -2) however, we choose to follow its original source[@mutschler_renaissance_2013], which defines class I as base deficit (Inverted BE) ≤ 2, with no lower limit (i.e.
+an upper limit for BE).
+
 ## Statistical methods
@@
 The statistical analysis will be performed using R, a programming language and environment for statistical computing.
-We will present the types of OFI as percentage distributions and then visualize it in a pie chart.
-Unadjusted and adjusted logistic regression will be used to determine the association between the OFI, degree/class of shock according to ATLS and other patient factors such as age, sex, preinjury ASA and ISS.
-It will be presented as odds ratios (OR) between presence of OFI, and shock classes.
+We will present the types of OFI as percentage distributions and then visualize them in a bar chart.
+Unadjusted logistic regression will be used to determine the association between the OFI, and degree of shock defined separately by classifying BE and SBP roughly according to ATLS classification of hemorrhagic shock.
+Adjusted logistic regression will then incorporate other patient factors such as age, sex, preinjury ASA, INR, and ISS.
+It will be presented as odds ratios (OR) between the presence of OFI, and shock classes.
 The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
 All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.
+Missing data will be addressed by listwise deletion.
 
@@
 In this study, data will be presented at the group level, ensuring that no individual patient details can be identified.
-Ethical approvals has been obtained for this study, with the record numbers: 2021-02541, 2021-03531 and 2023-02975-02.
+Ethical approvals have been obtained for this study, with the record numbers: 2021-02541, 2021-03531, and 2023-02975-02.
 
@@
 
-Descriptive data
-----------------
-
-(a) Give characteristics of study participants (eg demographic,
-clinical, social) and information on exposures and potential
-confounders
-
-(b) Indicate number of participants with missing data for each
-variable of interest
-
-(c) Cohort study — Summarise follow-up time (eg, average and total
-amount)
-
-Outcome data
-------------
-
-Cohort study — Report numbers of outcome events or summary measures
-over time
-
-Case-control study — Report numbers in each exposure category, or
-summary measures of exposure
-
-Cross-sectional study — Report numbers of outcome events or summary
-measures
-
-Main results
-------------
-
-(a) Give unadjusted estimates and, if applicable, confounder-adjusted
-estimates and their precision (eg, 95% confidence interval). Make
-clear which confounders were adjusted for and why they were included
-
-(b) Report category boundaries when continuous variables were
-categorized
-
-(c) If relevant, consider translating estimates of relative risk into
-absolute risk for a meaningful time period
-
-Other analyses 
---------------
-
-Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses 
--->
-
 Table 3 present the sample characteristics.
@@
 ```
+Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
+Results from the BE approximation done by Mutschler et al [mutschler_renaissance_2013] which is the same as the tenth edition ATLS shock classification, shows that class I had a mean SBP of 132,6 mmhg and class IV have 94,8 mmhg at the emergency department.
+As a SBP of 132,6 is at the higher range of normotensive, we will not be able to use it as a upper limit, but the 94,8 mmhg is close to our approximation of 90mmhg for class IV.
+
+We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).
+Urine output and pulse pressure were also neglected, due to lack of data.
+
+There are three reasons.
+The first one is ATLS tenth edition lack numerical values for vital parameters.
+The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
+The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
+
+Hantera missing med listwise deletion.
+
 # Conclusion

### 2024-11-08 thesis.Rmd
verkar ej funka på det viset
lyckats sortera ut så att jag visar tabellerna där jag vill att de ska visas

@@
 
-# Ethics
+## Ethics
 
@@
 
+```{r sample-characteristics, echo=FALSE}
+sample.characteristics.table
+```
+
 Table 4 present
+
+```{r BE-subofi, echo=FALSE}
+print(p)
+```
+
+Table 5
+
+```{r SBP-subofi, echo=FALSE}
+print(pSBP)
+```
+
+Table 6 - Subofi detailed
+
+```{r SBP-subofi-detailed, echo=FALSE}
+print(pd)
+```
+
+```{r BE-subofi-detailed, echo=FALSE}
+print(pSBPd)
+```
+
+print(sample.characteristics.table)
+print(sample.characteristics.tableBE)
+print(sample.characteristics.tableSBP)
+print(log_regBE_sample.characteristics.table)
+print(log_regSBP_sample.characteristics.table)
+print(log_regBE_sample.characteristics.table_unadjusted)
+print(log_regSBP_sample.characteristics.table_unadjusted)
 

### 2024-11-08 main.R

@@ -269,13 +269,13 @@ log_regSBP_sample.characteristics.table_unadjusted <- tbl_regression(log_regSBPu
 
 # ggplot2
 ofi_counts <- study.sample %>%
-  group_by(BE_class, ofi.categories.detailed) %>%
+  group_by(BE_class, ofi.categories.broad) %>%
   summarize(count = n()) %>%
   ungroup() %>%
   group_by(BE_class) %>%
   mutate(percent = count / sum(count) * 100)
 
-p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.detailed)) +
+p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.broad)) +
   geom_bar(stat = "identity", position = "dodge") +
   labs(
     title = "Distribution of OFI Broad Categories by BE Shock Class",
@@ -287,9 +287,7 @@ p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.det
     legend.title = element_text(size = 10),
     legend.position = "bottom"
   ) +
-  scale_fill_brewer(palette = "Set3", name = "OFI Categories")
-
-print(p)
+  scale_fill_brewer(palette = "Set1", name = "OFI Categories")
 
 # ggplot2 SBP
 ofi_counts <- study.sample %>%
@@ -313,16 +311,50 @@ pSBP <- ggplot(ofi_counts, aes(x = V4SBP_class, y = percent, fill = ofi.categori
   ) +
   scale_fill_brewer(palette = "Set1", name = "OFI Categories")
 
-print(pSBP)
+#sub ofi detailed
+# ggplot2
+ofi_counts <- study.sample %>%
+  group_by(BE_class, ofi.categories.detailed) %>%
+  summarize(count = n()) %>%
+  ungroup() %>%
+  group_by(BE_class) %>%
+  mutate(percent = count / sum(count) * 100)
 
-# Print
-print(sample.characteristics.table)
-print(sample.characteristics.tableBE)
-print(sample.characteristics.tableSBP)
-print(log_regBE_sample.characteristics.table)
-print(log_regSBP_sample.characteristics.table)
-print(log_regBE_sample.characteristics.table_unadjusted)
-print(log_regSBP_sample.characteristics.table_unadjusted)
+pd <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.detailed)) +
[…]

### 2024-11-08 thesis.Rmd
la till alla tabeller o dånt

--- 2024-11-08 thesis.Rmd
@@
 
+```{r sample-characteristics, echo=FALSE}
+sample.characteristics.table
+sample.characteristics.tableBE
+sample.characteristics.tableSBP
+```
+
 Table 4 present
+
+```{r BE-subofi, echo=FALSE}
+print(p)
+```
+
+Table 5
+
+```{r SBP-subofi, echo=FALSE}
+print(pSBP)
+```
+
+Table 6 - Subofi detailed
+
+```{r SBP-subofi-detailed, echo=FALSE}
+print(pd)
+```
+
+```{r BE-subofi-detailed, echo=FALSE}
+print(pSBPd)
+```
+
+Table adjusted log
+```{r Adjusted_log, echo=FALSE}
+log_regBE_sample.characteristics.table
+log_regSBP_sample.characteristics.table
+```
+
+Table unadjusted log
+```{r Unadjusted_log, echo=FALSE}
+log_regBE_sample.characteristics.table_unadjusted
+log_regSBP_sample.characteristics.table_unadjusted
+```
 

### 2024-11-09 thesis.Rmd
ny struktur och lite mer text

--- 2024-11-08 thesis.Rmd
@@
 There were a total of `r merged` patients in the trauma registry.
-After excluding the patients younger than 15 and/or were dead on arrival, there were `r excluded` patients left.
+After excluding the patients younger than 15 and/or were dead on arrival, there were `r excludednum` patients left.
 Out of those, `r sample` patients had been reviewed for the presence of OFI.
@@
 
-## Descriptive data
-The study demographic median age is ... and the most common gender is male, `r male`.
-
 The variable with most missing data is ... which lack the data for ... patients.
 
+```{r sample-characteristics, echo=FALSE}
+sample.characteristics.table
+```
+
+## Descriptive data
+
 ### BE classification
-For the outcome OFI, out of the `r sample` patients reviewed, `r ofi` patients had a positive OFI. Clinically, the
+The study demographics median age is `r age` and the most common gender is male, `r male`.
+Clinically, the median ISS is (...) which is class (...), median INR is (...), and the most common Preinjury ASA class is (...), at (...)%.
 
+```{r sample-characteristics BE, echo=FALSE}
+sample.characteristics.tableBE
+```
 ### SBP classification
+The study demographics median age is `r age` and the most common gender is male, `r male`.
+Clinically, the median ISS is (...) which is class (...), median INR is (...), and the most common Preinjury ASA class is (...), at (...)%.
 
-Table 3 present the sample characteristics.
+Table 3, 4, 5 present the sample characteristics.
 
-Table 4 present
+```{r sample-characteristics SBP, echo=FALSE}
+sample.characteristics.tableSBP
+```
+
+## Outcome data
+Out of the `r sample` patients reviewed, `r ofi` patients had at least one positive OFI.
+
+The most common
+
+Table 6 and 7 present sub-ofi broad
+
+```{r BE-subofi, echo=FALSE}
+print(p)
+```
+
+```{r SBP-subofi, echo=FALSE}
+print(pSBP)
+```
+
+Table 8 and 9 present sub-ofi detailed
+
+```{r SBP-subofi-detailed, echo=FALSE}
+print(pd)
+```
+
+```{r BE-subofi-detailed, echo=FALSE}
+print(pSBPd)
+```
+
+Table 10 and 11 adjusted log
+```{r Adjusted_log, echo=FALSE}
+log_regBE_sample.characteristics.table
+log_regSBP_sample.characteristics.table
+```
+
+Table 12 and 13 unadjusted log
+```{r Unadjusted_log, echo=FALSE}
+log_regBE_sample.characteristics.table_unadjusted
+log_regSBP_sample.characteristics.table_unadjusted
+```
 

### 2024-11-09 main.R

@@ -362,5 +362,5 @@ age <- inline_text(sample.characteristics.table, variable = pt_age_yrs, column =
 male <- inline_text(sample.characteristics.table, variable = pt_Gender, column = stat_0, level = "Male")
 merged <- nrow(merged.data)
 sample <- nrow(study.sample)
-excluded <- nrow(excluded)
+excludednum <- nrow(excluded)
 

### 2024-11-10 manuscript.Rmd
försökt fixa, men fortfarande bug
förde över metod från theisis och tog bort ethics

@@
 
-<!-- Present key elements of study design -->
-
 We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
-We then linked the two databases to assess what types of OFI occurred in patients arriving in shock and use logistic regression to assess the association between the degree of shock and odds of opportunities for improvement.
 
@@
 
-```{=html}
-<!-- Describe the setting, locations, and relevant dates, including
-periods of recruitment, exposure, follow-up, and data collection -->
-```
-The trauma registry includes about 14000 patients treated at Karolinska University Hospital in Solna between 2014 and 2023.
-The "problem" database is a subset of the trauma registry and includes about ?
-patients selected for review.
-The Karolinska University Hospital in Solna treats all major trauma in the greater metropolitan area of Stockholm.
+The trauma registry includes patients treated at Karolinska University Hospital in Solna, which treats all major trauma in the greater metropolitan area of Stockholm.
+We included patients registered between 2014 and 2023.
+The trauma care quality database is a subset of the trauma registry and includes patients selected for review.
 
@@
 
-```{=html}
-<!-- Cohort study: Give the eligibility criteria, and the sources and
-methods of selection of participants. Describe methods of
-follow-up. For matched studies, give matching criteria and number of
-exposed and unexposed -->
-```
-```{=html}
-<!-- Case-control study: Give the eligibility criteria, and the
-sources and + methods of case ascertainment and control
-selection. Give the rationale for the choice of cases and
-controls. For matched studies, give matching criteria and the number
-of controls per case -->
-```
-```{=html}
-<!-- Cross-sectional study: Give the eligibility criteria, and the
-sources and methods of selection of participants -->
-```
-To qualify into the trauma registry, requires admission by trauma team activation or any admission with an Injury Severity Score (ISS) of more than nine.
-
-Every trauma patient at Karolinska University Hospital is included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
+The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or an already admitted patient with an Injury Severity Score (ISS) greater than nine.
+Every trauma patient (registered in the trauma registry) at Karolinska University Hospital is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
 Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
 The identified OFIs are then categorized into broader areas.
-The multidisciplinary conferences are held every six to eight weeks, during which an average of ten patient cases are reviewed by experienced specialists from all trauma-related fields.
 The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
 
-We included all patients in "SWETRAU" and "Problem" between 2014 and 2023 We excluded patients younger than 15 and/or were dead on arrival.
+We included all patients in the trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.
 
@@
 
-```{=html}
-<!-- Clearly define all outcomes, exposures, predictors, potential
-confounders, and effect modifiers. Give diagnostic criteria, if
-applicable. For each variable of interest, give sources of data and
-details of methods of assessment (measurement). Describe comparability
-of assessment methods if there is more than one group -->
-```
-<!-- Elaborate on the outcome OFI. How is it defined? What are the different types of OFI? -->
-
-The outcome was defined as presence of atleast one OFI, determined by the multidisciplinary M&M conference in the "problem" registry.
-An OFI can be various types of errors, which are categorized into: clinical judgement error, inadequate resources, delay in treatment, missed injury, inadequate protocols, "preventable death" (den här var väl inte ens med i "problem", fast fanns med i studien som Martin delade på discord) and other errors.
-
-In this study, the outcomes will be studied in two ways, one is the amount of occurrence for various types of OFI in shock patients, and the other is binary where odds of OFI for every shock class i calculated out of presence of OFI or not.
-(kanske inte ens en del av metod)
-
-To classify the patients the following measurement will be used as independent variables for classification of shock regardless of their cause.
-Heart rate(HR), systolic blood pressure(SBP), respiratory rate(RR) and mental status.
-Based on these four measurements, the patients will be classified into four groups roughly based on the same value as in the ATLS trauma shock classifications.
-
-<!-- Elaborate on how you will categorise the degree of shock, what are the different levels?-->
-
-Due to lack of data on FAST, type of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
-
-Version 1, 
-Due to the tenth edition of ATLS classifications lack actual numbers, we will be using simplified and numerical version approximated by Dunham et al[@dunham_comparison_2017]. (INSERT PICTURE)
-
-Version 2, 
-Due to the tenth edition of ATLS classifications lack actual numbers, we will be approximating numbers to the values.
-For HR and RR we will be using the numbers from the ninth edition of the classification.
-The SBP will be divded into class I and II where the SBP is >110, class III 110-90mmhg and class IV <90mmhg.
-This selection is based on tenth edition, where SBP is defined as normal i class I to II and normal/below in class III and clearly below normal i class IV.
-These two studies, redefined hypotension as 110mmhg respective 90mmhg dependent on age[@eastridge_hypotension_2007][@oyetunji_redefining_2011], therefore we choose class III as 110-90mmgh and class IV below 90mmhg.
-For mental status, according to the tenth edition ATLS, class I and II should have normal GCS, and class III and IV have lower GCS.
-Therefore we decided to use GCS 15 for class I + II and GCS <15 for class III and IV.
-
- - Hur gör vi med de som hamnar mellan klasserna?
-To classify, all of the parameters in one class needs to be fulfilled.
-In the cases where the parameters belong in different classes, we will be classifying according to the most severe parameter in the same way as Dunham et al[@dunham_comparison_2017].
-At same time ATLS manual states one of the common pitfalls is "Diagnosis of shock can be missed when only a single parameter is used." therefore to improve sensitivity we decided to use the worst parameter for classification.
-
-To prevent classifying non-shock traumatic brain injury as class III and worse shock based on GCS, we decided to only include those who also present a circulatory parameter in accordance with class III or IV.
- - eller ta bort GCS helt och hållet? För i enlighet med ovan kommer GCS alltid att prioriteras bort och då finns ingen anlending att ha med det...
-
-Version 3, simplaste och den jag gillar mest, men kan andra sjukdomar som gluykolen förgiftning sabba tanken?
-We will be only using the base deficit parameter according to the tenth edition of the ATLS classification for haemorrhage.
-There are three reasons.
-The first one is ATLS tenth edition lack numerical values for vital parameters.
-The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
-The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
-The classes will be divided into class I: 0 - (-2), class II: (-2)-(-6), class III: (-6)-(-10), and class IV: <(-10).
+The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the trauma care quality registry.
+An OFI can be various types of errors, which are categorized into: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
+The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure(SBP) and once with base excess (BE), regardless of their cause of shock.
+Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications.
+We included five other metrics to be adjusted for, due to potential for confounding.
+Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous as they were directly from the register.
 
@@
 
-<!-- Describe any efforts to address potential sources of bias -->
-
 ## Study size
 
-<!-- Explain how the study size was arrived at -->
-
-<!-- Mention that you will include all available data-->
-
-All available data in "SWETRAU" and "Problem" database will be included.
+All available data in trauma registry and trauma care quality database will be included.
 
@@
 
-```{=html}
-<!-- Explain how quantitative variables were handled in the
-analyses. If applicable, describe which groupings were chosen and why
--->
-```
+The BE parameter was divided into four classes with these cutoffs: Class I (above -2), Class II (-2 to -6), Class III (-6 to -10), and Class IV (below -10).
+Similarly, SBP was divided into four classes: Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and Class IV (below 90 mmHg).
+This classification follows the tenth edition ATLS hemorrhage classification, which does not specify exact values for any parameter other than BE.
+To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.
+and Oyetunji et al., who redefined hypotension as 110 mmHg respective 90 mmHg dependent on age[@eastridge_hypotension_2007; @oyetunji_redefining_2011].
+Therefore, we used 110 mmHg as the lower limit for (Class I: no shock) - (ATLS: normal SBP) and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
+This also aligns with the classification done by Mutschler et al.[@mutschler_critical_2013].
+
+**ATLS tenth edition - Systolic blood pressure and Base excess (BD)**
+
+| **Parameter - original ATLS 10th edition**| **Class I** | **Class II** (Mild)  | **Class III** (Moderate)  | **Class IV** (Severe) |
+|-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
+| Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L     | –6 to –10 mEq/L           | –10 mEq/L or less             |
+| Systolic Blood pressure           | ↔           | ↔                    | ↔/↓                       | ↓                             |
+| **Parameter - numerical approximated SBP**                                                                                         |
+| Systolic Blood pressure           | >110 mmhg   | 109-100 mmhg         | 99-90 mmhg                | <90  mmhg                     |
+
+Please note that ATLS classification defines class I for the BE parameter as (0 to -2) however, we choose to follow its original source[@mutschler_renaissance_2013], which defines class I as base deficit (Inverted BE) ≤ 2, with no lower limit (i.e.
+an upper limit for BE).
+
 ## Statistical methods
 
-```{=html}
-<!--
-
-(a) Describe all statistical methods, including those used to control
-for confounding
-
-(b) Describe any methods used to examine subgroups and interactions
-
-(c) Explain how missing data were addressed
-
-(d)
-
-Cohort study: If applicable, explain how loss to follow-up was addressed
-
-Case-control study:If applicable, explain how matching of cases and
-controls was addressed
-
-Cross-sectional study: If applicable, describe analytical methods
-taking account of sampling strategy
-
-(e) Describe any sensitivity analyses
--->
-```
-<!-- Start with stating what software you'll use. Then say something about how you will describe the data. Then say something about how you will analyse the data. -->
-
-The statistical analysis will be performed using R, a programing language and environment for statistical computing.
-We will present the type of OFI as percentage distributions and then visualize it in a pie chart.
-Bivariable logistic regression will be used to determine the association between the OFI and degree/class of shock according to ATLS.
-It will be presented as odds ratios (OR) between presence of OFI, and shock classes.
+The statistical analysis will be performed using R, a programming language and environment for statistical computing.
+We will present the types of OFI as percentage distributions and then visualize them in a bar chart.
+Unadjusted logistic regression will be used to determine the association between the OFI, and degree of shock defined separately by classifying BE and SBP roughly according to ATLS classification of hemorrhagic shock.
+Adjusted logistic regression will then incorporate other patient factors such as age, sex, preinjury ASA, INR, and ISS.
+It will be presented as odds ratios (OR) between the presence of OFI, and shock classes.
 The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
-All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the “problem” database to ensure objectivity.
+All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.
+Missing data will be addressed by listwise deletion.
 
@@
 
-```{r sample-characteristics, echo=FALSE}
-sample.characteristics.table
-```
-
 Outcome data
@@
 You can include code in this document like this:
+
+```{r sample-characteristics, echo=FALSE}
+sample.characteristics.table
+```
 

### 2024-11-10 manuscript.Rmd
ändrade till passiv grej för theisis

@@
 
-The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or an already admitted patient with an Injury Severity Score (ISS) greater than nine.
-Every trauma patient (registered in the trauma registry) at Karolinska University Hospital is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
+Inclusion in the trauma registry requires either admission through trauma team activation or presenting with an Injury Severity Score (ISS) greater than nine after admission to the Karolinska University Hopsital.
+Each trauma patient is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
 Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
-The identified OFIs are then categorized into broader areas.
+The identified OFIs are then categorized into broad and detailed categories.
 The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.

### 2024-11-10 thesis.Rmd
ändrade till passiv grej för theisis

@@
 
-We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+A registry-based retrospective cohort study was conducted using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
 
@@
 
-The trauma registry includes patients treated at Karolinska University Hospital in Solna, which treats all major trauma in the greater metropolitan area of Stockholm.
+The trauma registry is compromised of patients treated at Karolinska University Hospital in Solna, which treats all major trauma in the greater metropolitan area of Stockholm.
 We included patients registered between 2014 and 2023.
@@
 
-The inclusion criteria for the trauma registry is a patient either admitted through trauma team activation or an already admitted patient with an Injury Severity Score (ISS) greater than nine.
-Every trauma patient (registered in the trauma registry) at Karolinska University Hospital is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
+Inclusion in the trauma registry requires either admission through trauma team activation or presenting with an Injury Severity Score (ISS) greater than nine after admission to the Karolinska University Hopsital.
+Each trauma patient is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
 Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
-The identified OFIs are then categorized into broader areas.
+The identified OFIs are then categorized into broad and detailed categories.
 The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
 
-We included all patients in the trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.
+All patients in the trauma registry and Trauma care quality database were included. Patients younger than 15 and/or were dead on arrival were excluded.
 
@@
 An OFI can be various types of errors, which are categorized into: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
-The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure(SBP) and once with base excess (BE), regardless of their cause of shock.
+Patients were classified into different shock severity categories in parallel and separately, once with systolic blood pressure (SBP) and once with base excess (BE), irrespective of shock cause.
 Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications.
-We included five other metrics to be adjusted for, due to potential for confounding.
+Five other metrics were included to be adjusted for, due to potential for confounding.
 Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous as they were directly from the register.
@@
 
-All available data in trauma registry and trauma care quality database will be included.
+All available data in trauma registry and trauma care quality database were included.
 
@@
 This classification follows the tenth edition ATLS hemorrhage classification, which does not specify exact values for any parameter other than BE.
-To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.
+To solve this, numerical SBP values were assigned to each class based on findings from Eastridge et al.
 and Oyetunji et al., who redefined hypotension as 110 mmHg respective 90 mmHg dependent on age[@eastridge_hypotension_2007; @oyetunji_redefining_2011].
-Therefore, we used 110 mmHg as the lower limit for (Class I: no shock) - (ATLS: normal SBP) and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
+Therefore, 110 mmHg was set as the lower limit for (Class I: no shock) - (ATLS: normal SBP), and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
 This also aligns with the classification done by Mutschler et al.[@mutschler_critical_2013].
@@
 
-Please note that ATLS classification defines class I for the BE parameter as (0 to -2) however, we choose to follow its original source[@mutschler_renaissance_2013], which defines class I as base deficit (Inverted BE) ≤ 2, with no lower limit (i.e.
+Please note that ATLS classification defines class I for the BE parameter as (0 to -2) however, the original source[@mutschler_renaissance_2013] defines class I as base deficit (Inverted BE) ≤ 2, with no lower limit (i.e.
 an upper limit for BE).
+The original source classification was chosen for this study.
 
@@
 
-The statistical analysis will be performed using R, a programming language and environment for statistical computing.
-We will present the types of OFI as percentage distributions and then visualize them in a bar chart.
-Unadjusted logistic regression will be used to determine the association between the OFI, and degree of shock defined separately by classifying BE and SBP roughly according to ATLS classification of hemorrhagic shock.
-Adjusted logistic regression will then incorporate other patient factors such as age, sex, preinjury ASA, INR, and ISS.
-It will be presented as odds ratios (OR) between the presence of OFI, and shock classes.
-The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
-All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.
-Missing data will be addressed by listwise deletion.
+The statistical analysis was performed using R, a programming language and environment for statistical computing.
+OFI subcategories was presented as percentage distributions and visualized in a bar chart.
+Unadjusted logistic regression was used to determine the association between the OFI, and degree of shock defined by BE and SBP classifications.
+Adjusted logistic regression incorporated other patient factors such as age, sex, preinjury ASA, INR, and ISS.
+It was presented as odds ratios (OR) between the presence of OFI, and shock classes.
+The OR was determined with 95% confidence intervals, and a significance level of 5% was used.
+All statistical analysis was first done on synthetic data and then later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.
+Missing data was addressed by listwise deletion.
 

### 2024-11-15 manuscript.Rmd
fixat errorn som inte kan knitta fixat ny kod i main så att jag ger ut en noNA dataframe istället för två varierande för BE och SBP. ändrat i theisis texten kort för att prova att den nya noNA fungerar.

@@
 
-```{=html}
-<!--
-
 Participants
@@
 
-There were a total of `r nrow(merged.data)` patients in the trauma registry.
-Out of these, `r nrow(study.sample)` patients had been reviewed for the presence of opportunities for improvement.
-After exluding patients younger than 15 and/or were dead on arrival, `r nrow(study.sample)` patients remained.
-A total of X patients were finally excluded due to missing data, leaving us with Y patients for the final analysis.
+There were a total of `r merged` patients in the trauma registry.
+After excluding the patients younger than 15 and/or were dead on arrival, there were `r excludednum` patients left.
+Out of those, `r sample` patients had been reviewed for the presence of OFI.
+A total of `r sample - log_reg_countBE` patients were excluded for the BE classification group due to missing data, resulting in `r log_reg_countBE` patients for the final analysis.
+For the SBP group, missing data accounted for exclusion of `r sample - log_reg_countSBP` patients, leaving a sum of `r log_reg_countSBP` patients for the final analysis.
 
@@
 The median age was `r age`.
-Most patients were male `r gender`.
+Most patients were male `r male`.
 
@@
 and sensitivity analyses
--->
-```
+
 You can include code in this document like this:

### 2024-11-15 thesis.Rmd
fixat errorn som inte kan knitta fixat ny kod i main så att jag ger ut en noNA dataframe istället för två varierande för BE och SBP. ändrat i theisis texten kort för att prova att den nya noNA fungerar.

@@
 Out of those, `r sample` patients had been reviewed for the presence of OFI.
-A total of `r sample - log_reg_countBE` patients were excluded for the BE classification group due to missing data, resulting in `r log_reg_countBE` patients for the final analysis.
-For the SBP group, missing data accounted for exclusion of `r sample - log_reg_countSBP` patients, leaving a sum of `r log_reg_countSBP` patients for the final analysis.
+A total of `r sample - log_reg_count` patients were excluded due to missing data, resulting in `r log_reg_count` patients for the final analysis.
 
@@
 Table 3, 4, 5 present the sample characteristics.
-
-```{r sample-characteristics SBP, echo=FALSE}
-sample.characteristics.tableSBP
-```
 

### 2024-11-15 main.R

@@ -131,32 +131,46 @@ study.sample <- study.sample %>%
     TRUE ~ "Class 1"
   ))
 
+# no missing data data frame for regression models
+reg.sample <- study.sample %>% 
+  filter(!is.na(pt_age_yrs) & 
+           !is.na(pt_Gender) & 
+           !is.na(ISS) & 
+           !is.na(ed_inr_numeric) & 
+           !is.na(pt_asa_preinjury) & 
+           !is.na(BE_class) &
+           !is.na(V4SBP_class)
+         )
+
+# Total rows used in the logistic regression
+log_reg_count <- nrow(reg.sample)
+
 # Binary logistic regression model - BE
 # Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
 log_regBE <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + BE_class,
   family = binomial,
-  data = study.sample
+  data = reg.sample
 )
 
 # Binary logistic regression model - SBP
 log_regSBP <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + V4SBP_class,
                family = binomial,
-               data = study.sample
+               data = reg.sample
 )
 
 # Binary logistic regression model unadjusted - BE
 log_regBEun <- glm(ofinum ~ BE_class,
                  family = binomial,
-                 data = study.sample
+                 data = reg.sample
 )
 
 # Binary logistic regression model unadjusted - SBP
 log_regSBPun <- glm(ofinum ~ V4SBP_class,
                   family = binomial,
-                  data = study.sample
+                  data = reg.sample
 )
 
-# Remove unused variables.
+# Remove no longer used variables - study sample
 # I suggest removing them from the lines 40-52 where the study data is created instead
 study.sample <- study.sample |>
   select(
@@ -165,6 +179,17 @@ study.sample <- study.sample |>
     -ofinum,
     -Deceased
   )
+
+# Remove no longer used variables - regression sample 
+# I suggest removing them from the lines 40-52 where the study data is created instead
[…]

### 2024-11-18 thesis.Rmd
test av olika tabell format

@@
 
-The variable with most missing data is ... which lack the data for ... patients.
+The variable with most missing data is BE which lack the data for 1721 patients.
 
@@
 ## Descriptive data
-
-### BE classification
 The study demographics median age is `r age` and the most common gender is male, `r male`.
@@
 ```
-### SBP classification
-The study demographics median age is `r age` and the most common gender is male, `r male`.
-Clinically, the median ISS is (...) which is class (...), median INR is (...), and the most common Preinjury ASA class is (...), at (...)%.
-
-Table 3, 4, 5 present the sample characteristics.
 
@@
 Table 6 and 7 present sub-ofi broad
+
+```{r ofitable, echo=FALSE, results='asis'}
+library(kableExtra)
+
+ofi_table %>%
+  kbl(
+    col.names = c("BE Shock Class", "OFI Category", "Count", "Percentage (%)"),
+    caption = "Counts and Percentages of OFI Broad Categories by BE Shock Class"
+  ) %>%
+  kable_classic(full_width = FALSE) %>%
+  row_spec(0, bold = TRUE) %>% # Optional: Makes header bold
+  group_rows(index = table(ofi_table$BE_class)) # Automatically merges BE_class rows
+
+```
+
+```{r combined_ofitable, echo=FALSE, results='asis'}
+library(kableExtra)
+
+combined_table %>%
+  kbl(
+    col.names = c(
+      "BE Shock Class",
+      "OFI Category",
+      "BE Count",
+      "BE Percentage (%)",
+      "SBP Shock Class",
+      "SBP Count",
+      "SBP Percentage (%)"
+    ),
+    caption = "Counts and Percentages of OFI Broad Categories by BE and SBP Shock Classes"
+  ) %>%
+  kable_classic(full_width = FALSE) %>%
+  row_spec(0, bold = TRUE) # Optional: Makes header bold
+
+```
 

### 2024-11-18 main.R

@@ -261,26 +261,30 @@ log_regSBP_sample.characteristics.table_unadjusted <- tbl_regression(log_regSBPu
 )
 
 # ggplot2
-ofi_counts <- reg.sample %>%
+ofi_counts_filtered <- reg.sample %>%
+  filter(ofi.categories.broad != "no") %>% 
   group_by(BE_class, ofi.categories.broad) %>%
-  summarize(count = n()) %>%
-  ungroup() %>%
-  group_by(BE_class) %>%
-  mutate(percent = count / sum(count) * 100)
+  summarize(count = n()) %>% 
+  ungroup()
 
-p <- ggplot(ofi_counts, aes(x = BE_class, y = percent, fill = ofi.categories.broad)) +
+Yaxis <- max(ofi_counts_filtered$count) + 10
+
+p <- ggplot(ofi_counts_filtered, aes(x = BE_class, y = count, fill = ofi.categories.broad)) +
   geom_bar(stat = "identity", position = "dodge") +
   labs(
     title = "Distribution of OFI Broad Categories by BE Shock Class",
     x = "BE Shock Class",
-    y = "Percentage of OFI Broad Categories"
+    y = "Count of OFI Broad Categories"
   ) +
   theme_minimal() +
   theme(
     legend.title = element_text(size = 10),
     legend.position = "bottom"
   ) +
-  scale_fill_brewer(palette = "Set1", name = "OFI Categories")
+  scale_fill_brewer(palette = "Set1", name = "OFI Categories") +
+  scale_y_continuous(limits = c(0, Yaxis)) 
+
+print(p)
 
 # ggplot2 SBP
 ofi_counts <- reg.sample %>%
@@ -357,3 +361,87 @@ merged <- nrow(merged.data)
 sample <- nrow(study.sample)
 excludednum <- nrow(excluded)
 
+
+#table
+library(dplyr)
+
+# Prepare the data
+ofi_table <- reg.sample %>%
+  filter(ofi.categories.broad != "no") %>% 
+  group_by(BE_class, ofi.categories.broad) %>%
+  summarize(count = n(), .groups = "drop") %>% 
+  group_by(BE_class) %>%
+  mutate(percentage = round((count / sum(count)) * 100, 1)) %>%
+  ungroup()
+
+
+#table 2.0
+library(dplyr)
+library(tidyr)
+
+# Define all possible OFI categories
+all_ofi_categories <- unique(reg.sample$ofi.categories.broad)
+
+# Define all BE and SBP classes
[…]

### 2024-11-20 manuscript.Rmd
Review manuscript and code

@@
 It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010] [@teixeira_preventable_2007][@ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury[@berry_shock_2015].
-It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not[@vang_shock_2022].
+The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010; @teixeira_preventable_2007; @ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury [@berry_shock_2015].
+It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not [@vang_shock_2022].
 
-The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023].
+The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ [@leech_shock_2023].
 There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive.
 These causes can occur alone or in combination.
-The most usual one during trauma are hypovolemic in form of a large bleeding[@teixeira_preventable_2007].
+The most usual one during trauma are hypovolemic in form of a large bleeding [@teixeira_preventable_2007].
 
 To differentiate the severity of trauma shock, we can divide shock into different degrees.
-One of such classification is the ATLS classification for major haemorrhage.
+One of such classification is the ATLS classification for major haemorrhage (See Table \@ref(tab:alts) (ADD CITATION).
 The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
 However, this only meant to be used for hypovolemic shocks and not the other types.
-There is other classifications for predicting massive transfusion (TASH,TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
+There is other classifications for predicting massive transfusion (TASH, TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
 The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
-ATLS in this regard, even though having its own shortcomings[@guly_vital_2011], still is one of the most used classifications for this topic(Källa).
+ATLS in this regard, even though having its own shortcomings [@guly_vital_2011], still is one of the most used classifications for this topic (Källa).
 
 | **Parameter**                     | **Class I** | **Class II** (Mild) | **Class III** (Moderate)  | **Class IV** (Severe)         |
-|-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
+|-----------------------------------|-------------|---------------------|---------------------------|-------------------------------|
 | Approximate blood loss            | <15%        | 15–30%              | 31–40%                    | >40%                          |
@@
 | Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
+
+Table: ATLS classification for haemorrhagic shock, adapted from (ADD CITATION) {#tbl:atls}.
 
@@
 The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the trauma care quality registry.
-An OFI can be various types of errors, which are categorized into: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
-The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure(SBP) and once with base excess (BE), regardless of their cause of shock.
+An OFI can be various types of preventable events, and can be categorized as: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
+
+The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure (SBP) and once with base excess (BE), regardless of their cause of shock.
 Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications.
-We included five other metrics to be adjusted for, due to potential for confounding.
-Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous as they were directly from the register.
+
+We included five other metrics to be adjusted for, due to potential for confounding. Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous.
 

### 2024-11-20 main.R

@@ -30,10 +30,10 @@ library(nnet)
 library(broom.helpers)
 
 ## Import data
-data <- import_data(test = TRUE)
+data <- import_data()
 
 # Merge data
-merged.data <- merge_data(data, test = TRUE)
+merged.data <- merge_data(data)
 
 # New add ofi categories
 merged.data <- add_ofi_categories(merged.data)
@@ -60,16 +60,18 @@ study.data <- merged.data |>
 
 # Those excluded
 excluded <- study.data |>
-  filter(Deceased == "False",
-         pt_age_yrs >= 15
+  filter(
+    Deceased == "False",
+    pt_age_yrs >= 15
   )
 
 # Exclude patients who were not reviewed for the presence of OFI
 study.sample <- study.data |>
-  filter(!is.na(ofi),
-         Deceased == "False",
-         pt_age_yrs >= 15
-         )
+  filter(
+    !is.na(ofi),
+    Deceased == "False",
+    pt_age_yrs >= 15
+  )
 
 # Function for converting to numeric
 convert_number <- function(x) {
@@ -85,7 +87,7 @@ study.sample$ofinum <- ifelse(study.sample$ofi == "Yes", 1, 0)
 # Converting ed_be_art to numeric
 BEnum <- convert_number(study.sample$ed_be_art)
 
-#Re-add the BE column as numeric to `study.sample`
+# Re-add the BE column as numeric to `study.sample`
 study.sample <- study.sample %>%
   mutate(ed_be_art_numeric = BEnum)
 
@@ -132,15 +134,14 @@ study.sample <- study.sample %>%
   ))
 
 # no missing data data frame for regression models
-reg.sample <- study.sample %>% 
-  filter(!is.na(pt_age_yrs) & 
-           !is.na(pt_Gender) & 
-           !is.na(ISS) & 
-           !is.na(ed_inr_numeric) & 
-           !is.na(pt_asa_preinjury) & 
-           !is.na(BE_class) &
-           !is.na(V4SBP_class)
-         )
+reg.sample <- study.sample %>%
+  filter(!is.na(pt_age_yrs) &
+    !is.na(pt_Gender) &
+    !is.na(ISS) &
+    !is.na(ed_inr_numeric) &
+    !is.na(pt_asa_preinjury) &
+    !is.na(BE_class) &
+    !is.na(V4SBP_class))
 
 # Total rows used in the logistic regression
[…]

### 2024-11-20 manuscript.Rmd
Further edits to manuscript

@@
 Similarly, SBP was divided into four classes: Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and Class IV (below 90 mmHg).
-This classification follows the tenth edition ATLS hemorrhage classification, which does not specify exact values for any parameter other than BE.
+This classification follows the tenth edition ATLS hemorrhage classification (ADD CITATION), which does not specify exact values for any parameter other than BE.
 To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.

### 2024-11-20 main.R
la till test för subofi

@@ -53,8 +53,8 @@ study.data <- merged.data |>
     ISS,
     ofi,
     ed_inr,
-    ofi.categories.broad,
     ofi.categories.detailed,
+    ofi.categories.broad,
     Deceased,
   )
 
@@ -218,8 +218,11 @@ sample.characteristics.tableBE <- tbl_summary(reg.sample,
   add_overall() |>
   add_p()
 
+print(sample.characteristics.tableBE)
+
 # Create a table of regression of sample - BE
-log_regBE_sample.characteristics.table <- tbl_regression(log_regBE,
+log_regBE_sample.characteristics.table <- tbl_regression(
+  log_regBE,
   exponentiate = TRUE,
   label = list(
     pt_age_yrs ~ "Age (Years)",
@@ -229,7 +232,11 @@ log_regBE_sample.characteristics.table <- tbl_regression(log_regBE,
     ISS ~ "Injury Severity Score",
     BE_class ~ "Shock classification - BE"
   )
-)
+) |>
+  add_nevent() |>
+  bold_p()
+
+print(log_regBE_sample.characteristics.table)
 
 # Create a table of regression of sample - SBP
 log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
@@ -374,6 +381,16 @@ ofi_table <- reg.sample %>%
   mutate(percentage = round((count / sum(count)) * 100, 1)) %>%
   ungroup()
 
+#table summery version
+BEsubofi <- reg.sample |>
+  select(
+    BE_class,
+    ofi.categories.broad,
+  )
+
+BEsubofi_tbl <- tbl_summary(BEsubofi, by = BE_class)
+
+print(BEsubofi_tbl)
 
 #table 2.0
 library(dplyr)

### 2024-11-21 main.R
la till combined

@@ -462,3 +462,11 @@ combined_table <- be_data_complete %>%
     SBP_count,
     SBP_percentage
   )
+
+combined_tableBE <- tbl_merge(
+  tbls = list(log_regBE_sample.characteristics.table_unadjusted, log_regBE_sample.characteristics.table),
+  tab_spanner = c("**Unadjusted Model**", "**Adjusted Model**")
+)
+
+print(combined_tableBE)
+

### 2024-11-21 thesis.Rmd
la till event rate

@@
 ```
+This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
+
+The results showed significant lower risk for
+
 Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].

### 2024-11-21 main.R

@@ -220,21 +220,44 @@ sample.characteristics.tableBE <- tbl_summary(reg.sample,
 
 print(sample.characteristics.tableBE)
 
-# Create a table of regression of sample - BE
-log_regBE_sample.characteristics.table <- tbl_regression(
-  log_regBE,
-  exponentiate = TRUE,
-  label = list(
-    pt_age_yrs ~ "Age (Years)",
-    pt_Gender ~ "Gender (M/F)",
-    pt_asa_preinjury ~ "Pre-injury ASA",
-    ed_inr_numeric ~ "INR",
-    ISS ~ "Injury Severity Score",
-    BE_class ~ "Shock classification - BE"
-  )
-) |>
-  add_nevent() |>
-  bold_p()
+# Create a table of of regression - BE
+log_regBE_sample.characteristics.table <- 
+  tbl_regression(
+    log_regBE,
+    exponentiate = TRUE,
+    label = list(
+      pt_age_yrs ~ "Age (Years)",
+      pt_Gender ~ "Gender (M/F)",
+      pt_asa_preinjury ~ "Pre-injury ASA",
+      ed_inr_numeric ~ "INR",
+      ISS ~ "Injury Severity Score",
+      BE_class ~ "Shock classification - BE"
+    )
+  ) |>
+  add_nevent(location = "level") |> # Adds event counts to the table
+  add_n(location = "level") |> # Adds total counts to the table
+  # Add event rate calculation
+  modify_table_body(
+    ~ .x |> 
+      dplyr::mutate(
+        stat_nevent_rate = 
+          ifelse(
+            !is.na(stat_nevent),
+            paste0(style_sigfig(stat_nevent / stat_n, scale = 100), "%"),
+            NA
+          ), 
+        .after = stat_nevent
+      )
+  ) |> 
+  # Combine columns for a cleaner display
+  modify_column_merge(
+    pattern = "{stat_nevent} / {stat_n} ({stat_nevent_rate})",
+    rows = !is.na(stat_nevent)
+  ) |>
+  # Update the header for event rate
+  modify_header(stat_nevent = "**Event Rate**") |>
+  bold_p() # Optional: Bold significant p-values
+
 
 print(log_regBE_sample.characteristics.table)
 

### 2024-11-22 main.R
ändrat ordningen av tabellerna

@@ -147,13 +147,13 @@ log_reg_count <- nrow(reg.sample)
 
 # Binary logistic regression model - BE
 # Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
-log_regBE <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + BE_class,
+log_regBE <- glm(ofinum ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
   family = binomial,
   data = reg.sample
 )
 
 # Binary logistic regression model - SBP
-log_regSBP <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + V4SBP_class,
+log_regSBP <- glm(ofinum ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
                family = binomial,
                data = reg.sample
 )
@@ -226,12 +226,12 @@ log_regBE_sample.characteristics.table <-
     log_regBE,
     exponentiate = TRUE,
     label = list(
+      BE_class ~ "Shock classification - BE",
       pt_age_yrs ~ "Age (Years)",
       pt_Gender ~ "Gender (M/F)",
       pt_asa_preinjury ~ "Pre-injury ASA",
       ed_inr_numeric ~ "INR",
-      ISS ~ "Injury Severity Score",
-      BE_class ~ "Shock classification - BE"
+      ISS ~ "Injury Severity Score"
     )
   ) |>
   add_nevent(location = "level") |> # Adds event counts to the table
@@ -486,10 +486,12 @@ combined_table <- be_data_complete %>%
     SBP_percentage
   )
 
+
+#combined reg analysis
 combined_tableBE <- tbl_merge(
-  tbls = list(log_regBE_sample.characteristics.table_unadjusted, log_regBE_sample.characteristics.table),
-  tab_spanner = c("**Unadjusted Model**", "**Adjusted Model**")
-)
+  tbls = list(log_regBE_sample.characteristics.table, log_regBE_sample.characteristics.table_unadjusted),
+  tab_spanner = c("**Adjusted Model**", "**Unadjusted Model**")
+) 
 
 print(combined_tableBE)
 

### 2024-11-24 thesis.Rmd
la till lite skiss för diskussionen

@@
 
-The results showed significant lower risk for
+The most common type of OFI is clinical judgement error, it is the most common in all BE shock classes and 2/4 classes for SBP classes.
+is around ...% of.
+It generally takes up ...% of all OFI defined as shock (class 2 and below).
+
+When OFI and shock severity are analysed alone, there are statistical significant correlation for these classes: ...
+.
+It shows positive correlation where the more severe the shock is, the higher the odds are for an OFI to occur.
+This in constrast with the adjusted results which showed inverted correlation betwen OFI and shock severity, with 63% lower odds for OFI when comparing the BE class 4 with class 1.
+However there is no correlation for the other classes nor the classes for the SBP group.
+
+When analysing the other adjusted factors, ISS seems to be a major confounder which correlates to both OFI and shock severity.
+This can explain why the results of the adjusted and unadjusted analyses are inverted.
 

### 2024-11-27 manuscript.Rmd
ny plats

--- 2024-11-15 manuscript.Rmd
@@
 ---
-title: "Manuscript title here"
-author: "Your name here"
+title: "Opportunities for improvement in the care of adult trauma patients arriving in shock"
+author: "Carl Skog"
 output:

### 2024-11-27 thesis.Rmd
ny plats

@@
 knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_dir = "out") })
+  rmarkdown::render(inputFile, encoding = encoding, output_file = "/home/carls/out/thesis.html") })
 editor_options: 
@@
 
+Table 3 - Sample characteristics showing missing data.
 ```{r sample-characteristics, echo=FALSE}
@@
 The study demographics median age is `r age` and the most common gender is male, `r male`.
-Clinically, the median ISS is (...) which is class (...), median INR is (...), and the most common Preinjury ASA class is (...), at (...)%.
+Clinically, the median ISS is 5 which is considered minor injuries, median INR is 1,00, and the most common Preinjury ASA class is 1, at 62%.
 
+Table 4 - Sample characteristics used in regression.
 ```{r sample-characteristics BE, echo=FALSE}
-sample.characteristics.tableBE
+sample.characteristics.table_reg
 ```
@@
 ## Outcome data
-Out of the `r sample` patients reviewed, `r ofi` patients had at least one positive OFI.
+Out of the `r log_reg_count` patients reviewed without missing data, `r ofi` patients had at least one positive OFI.
 
-The most common
+The most common broad sub-ofi category was clinical judgement error, at 103 (37%).
+The other three major broad sub-ofis are: inadequate resources 61 (22%), delay in treatment 49 (17%) and missed diagnosis 48 (17%).
+The smallest one, is other errors, at 5 (1,8%).
 
-Table 6 and 7 present sub-ofi broad
+When classified, the most common broad sub-ofi in class 1 (no shock) is clinical judgement error both in the BE and SBP groups, while the other sub-ofi also follow a similar distribution between the groups.
+The largest sub-ofi continuous to be clinical judgment error for BE group in class 2, 3 and 4, at 34,6%; 31,2%; and 42,9% in this order.
+On other hand, the largest sub-ofi in SBP class 2 is delay in treatment at 50%.
+The SBP class 3 is evenly distributed between clinical judgement error and missed diagnosis at 40% each.
+Lastly, class 4 SBP's most common sub-ofi is clinical judgment error 46,7%.
 
-```{r ofitable, echo=FALSE, results='asis'}
-library(kableExtra)
+Overall 3989 (81%) of the patients were classified as having no shock (class 1) according to the BE classification, the sum of the remaining classes accounted for 930 (19%) patients.
+The distribution of the classes were decreasing in numbers with increased severity, with class 2: 644 (13%), class 3: 170 (3,5%), and class 4: 116 (2,4%).
 
-ofi_table %>%
-  kbl(
-    col.names = c("BE Shock Class", "OFI Category", "Count", "Percentage (%)"),
-    caption = "Counts and Percentages of OFI Broad Categories by BE Shock Class"
-  ) %>%
-  kable_classic(full_width = FALSE) %>%
-  row_spec(0, bold = TRUE) %>% # Optional: Makes header bold
-  group_rows(index = table(ofi_table$BE_class)) # Automatically merges BE_class rows
+The SBP classification system sorted 4455 (91%) patients as having no shock (class 1), while the remaining classes with patients in shock accounted for 464 (9%) of all the patients.
+Among the classes defined with shock, class 2 was the biggest at 239 (4,9%), followed by class 4 at 127 (2,6%), and lastly class 3 with 99 (2,0%).
 
+Table 5 - Sub-OFI, classified according to BE
+
+```{r sub-ofi BE, echo=FALSE}
+BEsubofi_tbl
 ```
 
-```{r combined_ofitable, echo=FALSE, results='asis'}
-library(kableExtra)
+Table 6 - Sub-OFI, classified according to SBP
 
-combined_table %>%
-  kbl(
-    col.names = c(
-      "BE Shock Class",
-      "OFI Category",
-      "BE Count",
-      "BE Percentage (%)",
-      "SBP Shock Class",
-      "SBP Count",
-      "SBP Percentage (%)"
-    ),
-    caption = "Counts and Percentages of OFI Broad Categories by BE and SBP Shock Classes"
-  ) %>%
-  kable_classic(full_width = FALSE) %>%
-  row_spec(0, bold = TRUE) # Optional: Makes header bold
-
+```{r sub-ofi SBP, echo=FALSE}
+SBPsubofi_tbl
 ```
 
-```{r BE-subofi, echo=FALSE}
-print(p)
+## Main results
+The adjusted analyses showed only statistical significance when comparing BE class 4 to class 1, which had 63% lower odds for OFI (OR 0,37; 95% CI 0,14-0,83; p-value 0.026).
+
+The un-adjusted analysis showed when comparing to class 1 in BE group, class 2 (OR 1,61; 95% CI 1,17-2,20; p-value 0.003) and 3 (OR 1,91; 95% CI 1,08-3,16; p-value 0.018) were significant in increased risk for OFI.
+For the SBP group, comparison between class 1 and class 4 showed significant increased risk (OR 2,28; 95% CI 1,26-3,85; p-value 0.004).
+
+Table 7 and 8 adjusted log
+```{r Adjusted_log, echo=FALSE}
+combined_tableBE
+combined_tableSBP
 ```
-
-```{r SBP-subofi, echo=FALSE}
-print(pSBP)
-```
-
-Table 8 and 9 present sub-ofi detailed
-
-```{r SBP-subofi-detailed, echo=FALSE}
-print(pd)
-```
-
-```{r BE-subofi-detailed, echo=FALSE}
-print(pSBPd)
-```
-
-Table 10 and 11 adjusted log
-```{r Adjusted_log, echo=FALSE}
-log_regBE_sample.characteristics.table
-log_regSBP_sample.characteristics.table
-```
-
-Table 12 and 13 unadjusted log
-```{r Unadjusted_log, echo=FALSE}
-log_regBE_sample.characteristics.table_unadjusted
-log_regSBP_sample.characteristics.table_unadjusted
-```
-
-You can also embed plots:
-
-```{r plot, echo=FALSE}
-plot(pressure)
-```
-
-You can also mix text and code, so called inline code, like this: `r 2+5`.
 

### 2024-11-27 main.R

@@ -18,7 +18,7 @@
 ## them). They all need to be sourced however when you compile your
 ## manuscript file or run this file as a job, as that happens in a
 ## clean R session.
-noacsr::source_all_functions()
+#noacsr::source_all_functions()
 
 # Load packages
 library(rofi)
@@ -28,12 +28,13 @@ library(gtsummary)
 library(ggplot2)
 library(nnet)
 library(broom.helpers)
+library(kableExtra)
 
 ## Import data
-data <- import_data(test = TRUE)
+data <- import_data()
 
 # Merge data
-merged.data <- merge_data(data, test = TRUE)
+merged.data <- merge_data(data)
 
 # New add ofi categories
 merged.data <- add_ofi_categories(merged.data)
@@ -53,7 +54,6 @@ study.data <- merged.data |>
     ISS,
     ofi,
     ed_inr,
-    ofi.categories.detailed,
     ofi.categories.broad,
     Deceased,
   )
@@ -69,7 +69,7 @@ study.sample <- study.data |>
   filter(!is.na(ofi),
          Deceased == "False",
          pt_age_yrs >= 15
-         )
+  )
 
 # Function for converting to numeric
 convert_number <- function(x) {
@@ -140,7 +140,7 @@ reg.sample <- study.sample %>%
            !is.na(pt_asa_preinjury) & 
            !is.na(BE_class) &
            !is.na(V4SBP_class)
-         )
+  )
 
 # Total rows used in the logistic regression
 log_reg_count <- nrow(reg.sample)
@@ -148,26 +148,58 @@ log_reg_count <- nrow(reg.sample)
 # Binary logistic regression model - BE
 # Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
 log_regBE <- glm(ofinum ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
-  family = binomial,
-  data = reg.sample
-)
-
-# Binary logistic regression model - SBP
-log_regSBP <- glm(ofinum ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
-               family = binomial,
-               data = reg.sample
-)
-
-# Binary logistic regression model unadjusted - BE
-log_regBEun <- glm(ofinum ~ BE_class,
                  family = binomial,
[…]

### 2024-11-27 thesis.Rmd
fixade så att knitten funkade igen

@@
 knit: (function(inputFile, encoding) {
-  rmarkdown::render(inputFile, encoding = encoding, output_file = "/home/carls/out/thesis.html") })
+  rmarkdown::render(inputFile, encoding = encoding, output_file = "thesis") })
 editor_options: 

### 2024-11-27 manuscript.Rmd
fixat main så att den är uppdaterad.
Även lagt in den nya versionen av intro i manuscript.

@@
 ```
+# Introduction
+
+## Epidemiology
 Trauma is a major global public health concern.
-It causes over four milion death and affects hundreds of millions each year, among those younger populations are the most affected [@noauthor_injuries_nodate].
-The initial management of trauma patients once they reach hospital is very time sensitive and error prone [@mackersie_pitfalls_2010; @teixeira_preventable_2007; @ivatury_patient_2008], and shock is the most common potentially preventable death within the first 24h of injury [@berry_shock_2015].
-It is also the second major cause of early mortality in trauma patients [@kauvar_impact_2006], and patients in shock have a higher mortality rate than those not [@vang_shock_2022].
+It causes over four million deaths, among those younger populations are the most affected [@noauthor_injuries_nodate].
+Beyond the high mortality rate, trauma also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs [@noauthor_injuries_nodate].
+Hemorrhagic shock, the second most common cause of death following trauma, is the leading preventable cause of death within the first 24 hours post-injury  [@kauvar_impact_2006;@berry_shock_2015;@teixeira_preventable_2007].
 
-The definition for shock is circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ [@leech_shock_2023].
-There are four types of causes, which are hypovolemic, obstructive, cardiogenic and distributive.
-These causes can occur alone or in combination.
-The most usual one during trauma are hypovolemic in form of a large bleeding [@teixeira_preventable_2007].
-
-To differentiate the severity of trauma shock, we can divide shock into different degrees.
-One of such classification is the ATLS classification for major haemorrhage (See Table \@ref(tab:alts) (ADD CITATION).
-The classification is divided into four classes, based on the patients: estimated bloodloss, heart frequency (HF), bloodpressure (BP), pulse pressure, respiratory rate (RR), mental status and urine output.
-However, this only meant to be used for hypovolemic shocks and not the other types.
-There is other classifications for predicting massive transfusion (TASH, TBSS, ABC), triaging (RETTS) or tracking the clinical deterioration of patients (NEWS2).
-The common denominator for these classifications is they are all roughly based on vital parameters, however none of them is specifically designed to classify trauma shock degree regardless of the type/cause of shock.
-ATLS in this regard, even though having its own shortcomings [@guly_vital_2011], still is one of the most used classifications for this topic (Källa).
+## Shock classifications
+Out of the clinical needs to assess the severity of trauma shock, patients can be categorized by stratifying symptoms and physiological parameters into degrees.
+Many different scoring systems have been developed, such as ATLS classification for hemorrhagic shock (See Table \@ref(tab:alts) [@noauthor_atls_2018], Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion [@el-menyar_review_2019].
+While none of these systems comprehensively classifies shock severity across all types of trauma shock, each aid in assessing severity through core physiological indicators.
+Ultimately, these classifications rely on vital signs and clinical indicators, highlighting the importance of physiological metrics in evaluating shock severity across clinical settings.
 
@@
 
-Table: ATLS classification for haemorrhagic shock, adapted from (ADD CITATION) {#tbl:atls}.
+Table: ATLS classification for haemorrhagic shock, adapted from [@noauthor_atls_2018] {#tbl:atls}.
 
-One way to improve trauma care quality is trough identifying Opportunities for Improvement (OFI).
-This method has an advantage compared to traditional mortality reviews in that it also includes non-fatals.
-There are different processes to finding OFIs, but one of the primary ways is through structured multidisciplinary morbidity and mortality reviews of patient cases [@world_health_organization_guidelines_2009].
-In order for the review content to be as comprehensive as possible for trauma care, the Donabedian quality of care framework can be used[@donabedian_quality_1988].
-It is based on the three factors of structure, process and outcome.
-By following this framework, healthcare providers will be able to systematically find and address OFIs[källa?].
-This will in turn improve the effectiveness of trauma care, directly impacting mortality and morbidity[källa?/8].
+## Trauma quality improvement and opportunities for improvement
+The initial management of trauma patients is highly time-sensitive and prone to error  [@mackersie_pitfalls_2010;@teixeira_preventable_2007;@ivatury_patient_2008].
+Quality Improvement (QI) initiatives address these challenges by systematically evaluating care processes and outcomes, aiming to reduce morbidity and mortality [@world_health_organization_guidelines_2009].
+Through structured quality measures and collaborative reviews, QI has strengthened trauma care and improved outcomes globally  [@juillard_establishing_2009;@hashmi_hospital-based_2013].
+Identifying Opportunities for Improvement (OFI) offers a broader perspective than traditional mortality reviews by also including non-fatal outcomes [@world_health_organization_guidelines_2009].
+Structured multidisciplinary morbidity and mortality (M&M) reviews are commonly used to identify OFIs.
+One systemic and comprehensive way to identify OFI is using the Donabedian framework, which asses the quality of care by the three categories of: structure, process, and outcome  [@donabedian_quality_1988;@world_health_organization_guidelines_2009].
 
+## Shock and OFI
+As mentioned before hemorrhagic shock is the leading cause of preventable death within the first 24 hours of injury [@berry_shock_2015].
+Patients in shock have a higher mortality rate than those not [@vang_shock_2022], with a median time to death of just 2 hours [@tisherman_detailed_2015].
+Despite this, the relationship between shock, its severity, and Opportunities for Improvement (OFI) remains largely unexamined.
+This knowledge would help clinicians better identify and avoid preventable mistakes in shock patients who need urgent care.
+
+## Aims
 This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
@@
 
-Participants
-------------
-
-(a) Report numbers of individuals at each stage of study—eg numbers
-potentially eligible, examined for eligibility, confirmed eligible,
-included in the study, completing follow-up, and analysed
-
-(b) Give reasons for non-participation at each stage
-
-(c) Consider use of a flow diagram
-
+## Participants
 There were a total of `r merged` patients in the trauma registry.
@@
 Out of those, `r sample` patients had been reviewed for the presence of OFI.
-A total of `r sample - log_reg_countBE` patients were excluded for the BE classification group due to missing data, resulting in `r log_reg_countBE` patients for the final analysis.
-For the SBP group, missing data accounted for exclusion of `r sample - log_reg_countSBP` patients, leaving a sum of `r log_reg_countSBP` patients for the final analysis.
+A total of `r sample - log_reg_count` patients were excluded due to missing data, resulting in `r log_reg_count` patients for the final analysis.
 
-Descriptive data
-----------------
+The variable with most missing data is BE which lack the data for 1721 patients.
 
-(a) Give characteristics of study participants (eg demographic,
-clinical, social) and information on exposures and potential
-confounders
-
-(b) Indicate number of participants with missing data for each
-variable of interest
-
-(c) Cohort study — Summarise follow-up time (eg, average and total
-amount)
-
-Sample characteristics are presented in Table \@ref(tab:sample-characteristics).
-A total of `r ofi` patients had at least one opportunity for improvement.
-The median age was `r age`.
-Most patients were male `r male`.
-
-Outcome data
-------------
-
-Cohort study — Report numbers of outcome events or summary measures
-over time
-
-Case-control study — Report numbers in each exposure category, or
-summary measures of exposure
-
-Cross-sectional study — Report numbers of outcome events or summary
-measures
-
-Main results
-------------
-
-(a) Give unadjusted estimates and, if applicable, confounder-adjusted
-estimates and their precision (eg, 95% confidence interval). Make
-clear which confounders were adjusted for and why they were included
-
-(b) Report category boundaries when continuous variables were
-categorized
-
-(c) If relevant, consider translating estimates of relative risk into
-absolute risk for a meaningful time period
-
-Other analyses 
---------------
-
-Report other analyses done—eg analyses of subgroups and interactions,
-and sensitivity analyses
-
-You can include code in this document like this:
-
+Table 3 - Sample characteristics showing missing data.
 ```{r sample-characteristics, echo=FALSE}
@@
 
-You can also embed plots:
+## Descriptive data
+The study demographics median age is `r age` and the most common gender is male, `r male`.
+Clinically, the median ISS is 5 which is considered minor injuries, median INR is 1,00, and the most common Preinjury ASA class is 1, at 62%.
 
-```{r plot, echo=FALSE}
-plot(pressure)
+Table 4 - Sample characteristics used in regression.
+```{r sample-characteristics BE, echo=FALSE}
+sample.characteristics.table_reg
 ```
 
-You can also mix text and code, so called inline code, like this: `r 2+5`.
+## Outcome data
+Out of the `r log_reg_count` patients reviewed without missing data, `r ofi` patients had at least one positive OFI.
+
+The most common broad sub-ofi category was clinical judgement error, at 103 (37%).
+The other three major broad sub-ofis are: inadequate resources 61 (22%), delay in treatment 49 (17%) and missed diagnosis 48 (17%).
+The smallest one, is other errors, at 5 (1,8%).
+
+When classified, the most common broad sub-ofi in class 1 (no shock) is clinical judgement error both in the BE and SBP groups, while the other sub-ofi also follow a similar distribution between the groups.
+The largest sub-ofi continuous to be clinical judgment error for BE group in class 2, 3 and 4, at 34,6%; 31,2%; and 42,9% in this order.
+On other hand, the largest sub-ofi in SBP class 2 is delay in treatment at 50%.
+The SBP class 3 is evenly distributed between clinical judgement error and missed diagnosis at 40% each.
+Lastly, class 4 SBP's most common sub-ofi is clinical judgment error 46,7%.
+
+Overall 3989 (81%) of the patients were classified as having no shock (class 1) according to the BE classification, the sum of the remaining classes accounted for 930 (19%) patients.
+The distribution of the classes were decreasing in numbers with increased severity, with class 2: 644 (13%), class 3: 170 (3,5%), and class 4: 116 (2,4%).
+
+The SBP classification system sorted 4455 (91%) patients as having no shock (class 1), while the remaining classes with patients in shock accounted for 464 (9%) of all the patients.
+Among the classes defined with shock, class 2 was the biggest at 239 (4,9%), followed by class 4 at 127 (2,6%), and lastly class 3 with 99 (2,0%).
+
+Table 5 - Sub-OFI, classified according to BE
+
+```{r sub-ofi BE, echo=FALSE}
+BEsubofi_tbl
+```
+
+Table 6 - Sub-OFI, classified according to SBP
+
+```{r sub-ofi SBP, echo=FALSE}
+SBPsubofi_tbl
+```
+
+## Main results
+The adjusted analyses showed only statistical significance when comparing BE class 4 to class 1, which had 63% lower odds for OFI (OR 0,37; 95% CI 0,14-0,83; p-value 0.026).
+
+The un-adjusted analysis showed when comparing to class 1 in BE group, class 2 (OR 1,61; 95% CI 1,17-2,20; p-value 0.003) and 3 (OR 1,91; 95% CI 1,08-3,16; p-value 0.018) were significant in increased risk for OFI.
+For the SBP group, comparison between class 1 and class 4 showed significant increased risk (OR 2,28; 95% CI 1,26-3,85; p-value 0.004).
+
+Table 7 and 8 adjusted log
+```{r Adjusted_log, echo=FALSE}
+combined_tableBE
+combined_tableSBP
+```
 

### 2024-11-27 main.R

@@ -186,7 +186,7 @@ test <- tbl_regression(
 
 
 # Binary logistic regression model - SBP
-log_regSBP <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS + V4SBP_class,
+log_regSBP <- glm(ofinum ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
   family = binomial,
   data = reg.sample
 )
@@ -257,7 +257,7 @@ sample.characteristics.table <- tbl_summary(study.sample,
   add_p()
 
 # Create a table of sample characteristics - post reg
-sample.characteristics.tableBE <- tbl_summary(reg.sample,
+sample.characteristics.table_reg <- tbl_summary(reg.sample,
   by = ofi
 ) |>
   add_overall() |>
@@ -299,16 +299,17 @@ log_regBE_sample.characteristics.table <-
   bold_p() 
 
 # Create a table of regression of sample - SBP
-log_regSBP_sample.characteristics.table <- tbl_regression(log_regSBP,
-                                                          exponentiate = TRUE,
-                                                          label = list(
-                                                            pt_age_yrs ~ "Age (Years)",
-                                                            pt_Gender ~ "Gender (M/F)",
-                                                            pt_asa_preinjury ~ "Pre-injury ASA",
-                                                            ed_inr_numeric ~ "INR",
-                                                            ISS ~ "Injury Severity Score",
-                                                            V4SBP_class ~ "Shock classification - SBP"
-                                                          )
+log_regSBP_sample.characteristics.table <- 
+  tbl_regression(log_regSBP,
+    exponentiate = TRUE,
+    label = list(
+      pt_age_yrs ~ "Age (Years)",
+      pt_Gender ~ "Gender (M/F)",
+      pt_asa_preinjury ~ "Pre-injury ASA",
+      ed_inr_numeric ~ "INR",
+      ISS ~ "Injury Severity Score",
+      V4SBP_class ~ "Shock classification - SBP"
+  )
 ) |>
   add_nevent(location = "level") |> 
   add_n(location = "level") |> 
@@ -337,7 +338,8 @@ log_regBE_sample.characteristics.table_unadjusted <- tbl_regression(log_regBEun,
   label = list(
[…]

### 2024-11-27 manuscript.Rmd
Fört över ändrignar för metod från manuscript

@@
 Similarly, SBP was divided into four classes: Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and Class IV (below 90 mmHg).
-This classification follows the tenth edition ATLS hemorrhage classification (ADD CITATION), which does not specify exact values for any parameter other than BE.
+This classification follows the tenth edition ATLS hemorrhage classification [@noauthor_atls_2018], which does not specify exact values for any parameter other than BE.
 To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.

### 2024-11-27 thesis.Rmd
Fört över ändrignar för metod från manuscript

@@
 The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the trauma care quality registry.
-An OFI can be various types of errors, which are categorized into: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
-Patients were classified into different shock severity categories in parallel and separately, once with systolic blood pressure (SBP) and once with base excess (BE), irrespective of shock cause.
+An OFI can be various types of preventable events, and can be categorized as: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
+
+The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure (SBP) and once with base excess (BE), regardless of their cause of shock.
 Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications.
-Five other metrics were included to be adjusted for, due to potential for confounding.
-Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous as they were directly from the register.
+
+We included five other metrics to be adjusted for, due to potential for confounding. Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous.
 
@@
 ## Quantitative variables
-
 The BE parameter was divided into four classes with these cutoffs: Class I (above -2), Class II (-2 to -6), Class III (-6 to -10), and Class IV (below -10).
 Similarly, SBP was divided into four classes: Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and Class IV (below 90 mmHg).
-This classification follows the tenth edition ATLS hemorrhage classification, which does not specify exact values for any parameter other than BE.
-To solve this, numerical SBP values were assigned to each class based on findings from Eastridge et al.
+This classification follows the tenth edition ATLS hemorrhage classification [@noauthor_atls_2018], which does not specify exact values for any parameter other than BE.
+To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.
 and Oyetunji et al., who redefined hypotension as 110 mmHg respective 90 mmHg dependent on age[@eastridge_hypotension_2007; @oyetunji_redefining_2011].
-Therefore, 110 mmHg was set as the lower limit for (Class I: no shock) - (ATLS: normal SBP), and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
+Therefore, we used 110 mmHg as the lower limit for (Class I: no shock) - (ATLS: normal SBP) and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
 This also aligns with the classification done by Mutschler et al.[@mutschler_critical_2013].

### 2024-11-28 manuscript.Rmd
försökt åtgärda ref funktionen men inte fått till de.

@@
 ---
-title: "Opportunities for improvement in the care of adult trauma patients arriving in shock"
+title: Opportunities for improvement in the care of adult trauma patients arriving
+  in shock
 author: "Carl Skog"
 output:
+  word_document: default
+  pdf_document: default
   html_document: default
-  word_document: default
 bibliography: bibliography.bib
@@
 Out of the clinical needs to assess the severity of trauma shock, patients can be categorized by stratifying symptoms and physiological parameters into degrees.
-Many different scoring systems have been developed, such as ATLS classification for hemorrhagic shock (See Table \@ref(tab:alts) [@noauthor_atls_2018], Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion [@el-menyar_review_2019].
+Many different scoring systems have been developed, such as ATLS classification for hemorrhagic shock (See Table \@ref(tab:alts) [@noauthor_atls_2018]), Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion [@el-menyar_review_2019].
 While none of these systems comprehensively classifies shock severity across all types of trauma shock, each aid in assessing severity through core physiological indicators.

### 2024-11-28 manuscript.tex
försökt åtgärda ref funktionen men inte fått till de.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{anyfontsize}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Opportunities for improvement in the care of adult trauma patients arriving in shock},
  pdfauthor={Carl Skog},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Opportunities for improvement in the care of adult trauma
patients arriving in shock}
\author{Carl Skog}
\date{}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

\hypertarget{background}{%
\subsection{Background}\label{background}}

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{introduction-1}{%
\section{Introduction}\label{introduction-1}}

\hypertarget{epidemiology}{%
\subsection{Epidemiology}\label{epidemiology}}

Trauma is a major global public health concern. It causes over four
million deaths, among those younger populations are the most affected
(1). Beyond the high mortality rate, trauma also imposes a substantial
socio-economic burden, due to long-term disabilities and rehabilitation
needs (1). Hemorrhagic shock, the second most common cause of death
following trauma, is the leading preventable cause of death within the
first 24 hours post-injury (2--4).

\hypertarget{shock-classifications}{%
\subsection{Shock classifications}\label{shock-classifications}}

Out of the clinical needs to assess the severity of trauma shock,
patients can be categorized by stratifying symptoms and physiological
parameters into degrees. Many different scoring systems have been
developed, such as ATLS classification for hemorrhagic shock (See Table
@ref(tab:alts) (5)), Trauma Associated Severe Hemorrhage (TASH) and
Assessment of Blood Consumption (ABC) scores for predicting massive
transfusion (6). While none of these systems comprehensively classifies
shock severity across all types of trauma shock, each aid in assessing
severity through core physiological indicators. Ultimately, these
classifications rely on vital signs and clinical indicators,
highlighting the importance of physiological metrics in evaluating shock
severity across clinical settings.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2756}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1024}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1654}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2126}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2441}}@{}}
\caption{ATLS classification for haemorrhagic shock, adapted from (5)
\{\#tbl:atls\}.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class I}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class II} (Mild)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class III} (Moderate)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class IV} (Severe)
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class I}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class II} (Mild)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class III} (Moderate)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class IV} (Severe)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Approximate blood loss & \textless15\% & 15--30\% & 31--40\% &
\textgreater40\% \\
Heart rate & ↔ & ↔/↑ & ↑ & ↑/↑↑ \\
Blood pressure & ↔ & ↔ & ↔/↓ & ↓ \\
Pulse pressure & ↔ & ↓ & ↓ & ↓ \\
Respiratory rate & ↔ & ↔ & ↔/↑ & ↑ \\
Urine output & ↔ & ↔ & ↓ & ↓↓ \\
Glasgow Coma Scale score & ↔ & ↔ & ↓ & ↓ \\
Base deficit & 0 to --2 mEq/L & --2 to --6 mEq/L & --6 to --10 mEq/L &
--10 mEq/L or less \\
Need for blood products & Monitor & Possible & Yes & Massive Transfusion
Protocol \\
\end{longtable}

\hypertarget{trauma-quality-improvement-and-opportunities-for-improvement}{%
\subsection{Trauma quality improvement and opportunities for
improvement}\label{trauma-quality-improvement-and-opportunities-for-improvement}}

The initial management of trauma patients is highly time-sensitive and
prone to error (4,7,8). Quality Improvement (QI) initiatives address
these challenges by systematically evaluating care processes and
outcomes, aiming to reduce morbidity and mortality (9). Through
structured quality measures and collaborative reviews, QI has
strengthened trauma care and improved outcomes globally (10,11).
Identifying Opportunities for Improvement (OFI) offers a broader
perspective than traditional mortality reviews by also including
non-fatal outcomes (9). Structured multidisciplinary morbidity and
mortality (M\&M) reviews are commonly used to identify OFIs. One
systemic and comprehensive way to identify OFI is using the Donabedian
framework, which asses the quality of care by the three categories of:
structure, process, and outcome (9,12).

\hypertarget{shock-and-ofi}{%
\subsection{Shock and OFI}\label{shock-and-ofi}}

As mentioned before hemorrhagic shock is the leading cause of
preventable death within the first 24 hours of injury (3). Patients in
shock have a higher mortality rate than those not (13), with a median
time to death of just 2 hours (14). Despite this, the relationship
between shock, its severity, and Opportunities for Improvement (OFI)
remains largely unexamined. This knowledge would help clinicians better
identify and avoid preventable mistakes in shock patients who need
urgent care.

\hypertarget{aims}{%
\subsection{Aims}\label{aims}}

This study aims to describe the types of opportunities for improvement
for adult trauma patients arriving in shock, and to assess how the
degree of shock is associated with opportunities for improvement.

\hypertarget{methods-1}{%
\section{Methods}\label{methods-1}}

\hypertarget{study-design}{%
\subsection{Study design}\label{study-design}}

We conducted a registry based retrospective cohort study, using data
from the trauma registry and trauma care quality database at the
Karolinska University Hospital in Solna.

\hypertarget{setting}{%
\subsection{Setting}\label{setting}}

The trauma registry includes patients treated at Karolinska University
Hospital in Solna, which treats all major trauma in the greater
metropolitan area of Stockholm. We included patients registered between
2014 and 2023. The trauma care quality database is a subset of the
trauma registry and includes patients selected for review.

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

Inclusion in the trauma registry requires either admission through
trauma team activation or presenting with an Injury Severity Score (ISS)
greater than nine after admission to the Karolinska University Hopsital.
Each trauma patient is then included in a morbidity and mortality review
process, which involves both individual case evaluations by specialized
nurses and audit filters. Patients identified with a high potential for
OFIs are discussed at multidisciplinary conferences. The identified OFIs
are then categorized into broad and detailed categories. The presence or
absence of OFIs is determined by consensus among all participants and is
documented in the trauma care quality database.

We included all patients in the trauma registry and Trauma care quality
database. We excluded patients younger than 15 and/or were dead on
arrival.

\hypertarget{variables-and-data-sourcesmeasurements}{%
\subsection{Variables and data
sources/measurements}\label{variables-and-data-sourcesmeasurements}}

The outcome was defined as the presence of at least one OFI, determined
by the multidisciplinary M\&M conference in the trauma care quality
registry. An OFI can be various types of preventable events, and can be
categorized as: clinical judgment error, inadequate resources, delay in
treatment, missed injury, inadequate protocols, preventable death and
other errors.

The patients will be classified to different degrees of shock in
parallel and separately, once with systolic blood pressure (SBP) and
once with base excess (BE), regardless of their cause of shock. Based on
these two parameters, the patients will be classified into four classes
roughly based on the ATLS trauma shock classifications.

We included five other metrics to be adjusted for, due to potential for
confounding. Pre-injury ASA and gender were categorical, meanwhile age,
ISS and INR were kept continuous.

\hypertarget{bias}{%
\subsection{Bias}\label{bias}}

\hypertarget{study-size}{%
\subsection{Study size}\label{study-size}}

All available data in trauma registry and trauma care quality database
will be included.

\hypertarget{quantitative-variables}{%
\subsection{Quantitative variables}\label{quantitative-variables}}

The BE parameter was divided into four classes with these cutoffs: Class
I (above -2), Class II (-2 to -6), Class III (-6 to -10), and Class IV
(below -10). Similarly, SBP was divided into four classes: Class I
(above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and
Class IV (below 90 mmHg). This classification follows the tenth edition
ATLS hemorrhage classification (5), which does not specify exact values
for any parameter other than BE. To solve this, we assigned numerical
SBP values to each class based on findings from Eastridge et al.~and
Oyetunji et al., who redefined hypotension as 110 mmHg respective 90
mmHg dependent on age(15,16). Therefore, we used 110 mmHg as the lower
limit for (Class I: no shock) - (ATLS: normal SBP) and below 90 mmHg for
(Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3
divided equally in between. This also aligns with the classification
done by Mutschler et al.(17).

\textbf{ATLS tenth edition - Systolic blood pressure and Base excess
(BD)}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2734}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1016}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1719}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2109}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2422}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Parameter - original ATLS 10th edition}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class I}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class II} (Mild)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class III} (Moderate)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Class IV} (Severe)
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Base deficit & 0 to --2 mEq/L & --2 to --6 mEq/L & --6 to --10 mEq/L &
--10 mEq/L or less \\
Systolic Blood pressure & ↔ & ↔ & ↔/↓ & ↓ \\
\textbf{Parameter - numerical approximated SBP} & & & & \\
Systolic Blood pressure & \textgreater110 mmhg & 109-100 mmhg & 99-90
mmhg & \textless90 mmhg \\
\end{longtable}

Please note that ATLS classification defines class I for the BE
parameter as (0 to -2) however, we choose to follow its original
source(18), which defines class I as base deficit (Inverted BE) ≤ 2,
with no lower limit (i.e.~an upper limit for BE).

\hypertarget{statistical-methods}{%
\subsection{Statistical methods}\label{statistical-methods}}

The statistical analysis will be performed using R, a programming
language and environment for statistical computing. We will present the
types of OFI as percentage distributions and then visualize them in a
bar chart. Unadjusted logistic regression will be used to determine the
association between the OFI, and degree of shock defined separately by
classifying BE and SBP roughly according to ATLS classification of
hemorrhagic shock. Adjusted logistic regression will then incorporate
other patient factors such as age, sex, preinjury ASA, INR, and ISS. It
will be presented as odds ratios (OR) between the presence of OFI, and
shock classes. The OR will be determined with 95\% confidence intervals,
and a significance level of 5\% will be used. All statistical analysis
will first be done on synthetic data and later implemented on the data
collected from the trauma registry and the trauma care quality database
to ensure objectivity. Missing data will be addressed by listwise
deletion.

\hypertarget{results-1}{%
\section{Results}\label{results-1}}

\hypertarget{participants-1}{%
\subsection{Participants}\label{participants-1}}

There were a total of 14022 patients in the trauma registry. After
excluding the patients younger than 15 and/or were dead on arrival,
there were 12153 patients left. Out of those, 7152 patients had been
reviewed for the presence of OFI. A total of 2233 patients were excluded
due to missing data, resulting in 4919 patients for the final analysis.

The variable with most missing data is BE which lack the data for 1721
patients.

Table 3 - Sample characteristics showing missing data.

\begin{table}[!t]
\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lcccc}
\toprule
\textbf{Characteristic} & \textbf{Overall}  N = 7,152\textsuperscript{\textit{1}} & \textbf{No}  N = 6,716\textsuperscript{\textit{1}} & \textbf{Yes}  N = 436\textsuperscript{\textit{1}} & \textbf{p-value}\textsuperscript{\textit{2}} \\ 
\midrule\addlinespace[2.5pt]
Age (Years) & 40 (26, 56) & 40 (26, 56) & 45 (28, 61) & <0.001 \\ 
Gender (M/F) &  &  &  & 0.14 \\ 
    Female & 2,159 (30\%) & 2,041 (30\%) & 118 (27\%) &  \\ 
    Male & 4,993 (70\%) & 4,675 (70\%) & 318 (73\%) &  \\ 
Pre-injury ASA &  &  &  & 0.031 \\ 
    1 & 4,362 (61\%) & 4,121 (61\%) & 241 (55\%) &  \\ 
    2 & 1,803 (25\%) & 1,684 (25\%) & 119 (27\%) &  \\ 
    3 & 949 (13\%) & 874 (13\%) & 75 (17\%) &  \\ 
    4 & 29 (0.4\%) & 28 (0.4\%) & 1 (0.2\%) &  \\ 
    Unknown & 9 & 9 & 0 &  \\ 
Systolic blood pressure (mmhg) & 135 (121, 150) & 135 (121, 150) & 135 (120, 150) & 0.6 \\ 
    Unknown & 123 & 114 & 9 &  \\ 
Injury Severity Score & 9 (1, 16) & 8 (1, 14) & 17 (10, 24) & <0.001 \\ 
    Unknown & 2 & 2 & 0 &  \\ 
OFI categories broad &  &  &  &  \\ 
    Clinical judgement error & 157 (36\%) & 0 (NA\%) & 157 (36\%) &  \\ 
    Delay in treatment & 67 (15\%) & 0 (NA\%) & 67 (15\%) &  \\ 
    Documentation Issues & 14 (3.2\%) & 0 (NA\%) & 14 (3.2\%) &  \\ 
    Inadequate protocols & 21 (4.8\%) & 0 (NA\%) & 21 (4.8\%) &  \\ 
    Inadequate resources & 104 (24\%) & 0 (NA\%) & 104 (24\%) &  \\ 
    Missed diagnosis & 67 (15\%) & 0 (NA\%) & 67 (15\%) &  \\ 
    Other errors & 5 (1.1\%) & 0 (NA\%) & 5 (1.1\%) &  \\ 
    Unknown & 6,717 & 6,716 & 1 &  \\ 
Base Excess (BE) & 0.8 (-1.3, 2.1) & 0.9 (-1.2, 2.2) & 0.0 (-2.4, 1.5) & <0.001 \\ 
    Unknown & 1,721 & 1,597 & 124 &  \\ 
INR & 1.00 (1.00, 1.10) & 1.00 (1.00, 1.10) & 1.00 (1.00, 1.10) & <0.001 \\ 
    Unknown & 1,620 & 1,517 & 103 &  \\ 
Shock classification - BE &  &  &  & 0.003 \\ 
    Class 1 & 4,391 (81\%) & 4,162 (81\%) & 229 (73\%) &  \\ 
    Class 2 & 719 (13\%) & 664 (13\%) & 55 (18\%) &  \\ 
    Class 3 & 191 (3.5\%) & 172 (3.4\%) & 19 (6.1\%) &  \\ 
    Class 4 & 130 (2.4\%) & 121 (2.4\%) & 9 (2.9\%) &  \\ 
    Unknown & 1,721 & 1,597 & 124 &  \\ 
Shock classification - SBP &  &  &  & 0.006 \\ 
    Class 1 & 6,356 (90\%) & 5,983 (91\%) & 373 (87\%) &  \\ 
    Class 2 & 355 (5.1\%) & 331 (5.0\%) & 24 (5.6\%) &  \\ 
    Class 3 & 147 (2.1\%) & 138 (2.1\%) & 9 (2.1\%) &  \\ 
    Class 4 & 171 (2.4\%) & 150 (2.3\%) & 21 (4.9\%) &  \\ 
    Unknown & 123 & 114 & 9 &  \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Median (Q1, Q3); n (\%)\\
\textsuperscript{\textit{2}}Wilcoxon rank sum test; Pearson's Chi-squared test; Fisher's exact test\\
\end{minipage}
\end{table}

\hypertarget{descriptive-data}{%
\subsection{Descriptive data}\label{descriptive-data}}

The study demographics median age is 39 (25, 56) and the most common
gender is male, 3,506 (71\%). Clinically, the median ISS is 5 which is
considered minor injuries, median INR is 1,00, and the most common
Preinjury ASA class is 1, at 62\%.

Table 4 - Sample characteristics used in regression.

\begin{table}[!t]
\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lcccc}
\toprule
\textbf{Characteristic} & \textbf{Overall}  N = 4,919\textsuperscript{\textit{1}} & \textbf{No}  N = 4,638\textsuperscript{\textit{1}} & \textbf{Yes}  N = 281\textsuperscript{\textit{1}} & \textbf{p-value}\textsuperscript{\textit{2}} \\ 
\midrule\addlinespace[2.5pt]
Age (Years) & 39 (25, 56) & 39 (25, 55) & 44 (26, 60) & 0.016 \\ 
Gender (M/F) &  &  &  & 0.2 \\ 
    Female & 1,413 (29\%) & 1,341 (29\%) & 72 (26\%) &  \\ 
    Male & 3,506 (71\%) & 3,297 (71\%) & 209 (74\%) &  \\ 
Pre-injury ASA &  &  &  & 0.12 \\ 
    1 & 3,032 (62\%) & 2,874 (62\%) & 158 (56\%) &  \\ 
    2 & 1,240 (25\%) & 1,163 (25\%) & 77 (27\%) &  \\ 
    3 & 626 (13\%) & 580 (13\%) & 46 (16\%) &  \\ 
    4 & 21 (0.4\%) & 21 (0.5\%) & 0 (0\%) &  \\ 
Systolic blood pressure (mmhg) & 136 (122, 150) & 136 (122, 150) & 136 (120, 150) & 0.8 \\ 
Injury Severity Score & 5 (1, 14) & 5 (1, 13) & 17 (9, 22) & <0.001 \\ 
OFI categories broad &  &  &  &  \\ 
    Clinical judgement error & 103 (37\%) & 0 (NA\%) & 103 (37\%) &  \\ 
    Delay in treatment & 49 (17\%) & 0 (NA\%) & 49 (17\%) &  \\ 
    Documentation Issues & 9 (3.2\%) & 0 (NA\%) & 9 (3.2\%) &  \\ 
    Inadequate protocols & 6 (2.1\%) & 0 (NA\%) & 6 (2.1\%) &  \\ 
    Inadequate resources & 61 (22\%) & 0 (NA\%) & 61 (22\%) &  \\ 
    Missed diagnosis & 48 (17\%) & 0 (NA\%) & 48 (17\%) &  \\ 
    Other errors & 5 (1.8\%) & 0 (NA\%) & 5 (1.8\%) &  \\ 
    Unknown & 4,638 & 4,638 & 0 &  \\ 
Base Excess (BE) & 0.9 (-1.2, 2.2) & 0.9 (-1.1, 2.3) & 0.0 (-2.4, 1.4) & <0.001 \\ 
INR & 1.00 (1.00, 1.10) & 1.00 (1.00, 1.10) & 1.00 (1.00, 1.10) & <0.001 \\ 
Shock classification - BE &  &  &  & 0.004 \\ 
    Class 1 & 3,989 (81\%) & 3,783 (82\%) & 206 (73\%) &  \\ 
    Class 2 & 644 (13\%) & 592 (13\%) & 52 (19\%) &  \\ 
    Class 3 & 170 (3.5\%) & 154 (3.3\%) & 16 (5.7\%) &  \\ 
    Class 4 & 116 (2.4\%) & 109 (2.4\%) & 7 (2.5\%) &  \\ 
Shock classification - SBP &  &  &  & 0.028 \\ 
    Class 1 & 4,455 (91\%) & 4,208 (91\%) & 247 (88\%) &  \\ 
    Class 2 & 239 (4.9\%) & 225 (4.9\%) & 14 (5.0\%) &  \\ 
    Class 3 & 98 (2.0\%) & 93 (2.0\%) & 5 (1.8\%) &  \\ 
    Class 4 & 127 (2.6\%) & 112 (2.4\%) & 15 (5.3\%) &  \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}Median (Q1, Q3); n (\%)\\
\textsuperscript{\textit{2}}Wilcoxon rank sum test; Pearson's Chi-squared test; Fisher's exact test\\
\end{minipage}
\end{table}

\hypertarget{outcome-data}{%
\subsection{Outcome data}\label{outcome-data}}

Out of the 4919 patients reviewed without missing data, 281 (5.7\%)
patients had at least one positive OFI.

The most common broad sub-ofi category was clinical judgement error, at
103 (37\%). The other three major broad sub-ofis are: inadequate
resources 61 (22\%), delay in treatment 49 (17\%) and missed diagnosis
48 (17\%). The smallest one, is other errors, at 5 (1,8\%).

When classified, the most common broad sub-ofi in class 1 (no shock) is
clinical judgement error both in the BE and SBP groups, while the other
sub-ofi also follow a similar distribution between the groups. The
largest sub-ofi continuous to be clinical judgment error for BE group in
class 2, 3 and 4, at 34,6\%; 31,2\%; and 42,9\% in this order. On other
hand, the largest sub-ofi in SBP class 2 is delay in treatment at 50\%.
The SBP class 3 is evenly distributed between clinical judgement error
and missed diagnosis at 40\% each. Lastly, class 4 SBP's most common
sub-ofi is clinical judgment error 46,7\%.

Overall 3989 (81\%) of the patients were classified as having no shock
(class 1) according to the BE classification, the sum of the remaining
classes accounted for 930 (19\%) patients. The distribution of the
classes were decreasing in numbers with increased severity, with class
2: 644 (13\%), class 3: 170 (3,5\%), and class 4: 116 (2,4\%).

The SBP classification system sorted 4455 (91\%) patients as having no
shock (class 1), while the remaining classes with patients in shock
accounted for 464 (9\%) of all the patients. Among the classes defined
with shock, class 2 was the biggest at 239 (4,9\%), followed by class 4
at 127 (2,6\%), and lastly class 3 with 99 (2,0\%).

Table 5 - Sub-OFI, classified according to BE

\begin{table}[!t]
\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lcccc}
\toprule
\textbf{Characteristic} & \textbf{Class 1}  N = 206\textsuperscript{\textit{1}} & \textbf{Class 2}  N = 52\textsuperscript{\textit{1}} & \textbf{Class 3}  N = 16\textsuperscript{\textit{1}} & \textbf{Class 4}  N = 7\textsuperscript{\textit{1}} \\ 
\midrule\addlinespace[2.5pt]
OFI categories broad &  &  &  &  \\ 
    Clinical judgement error & 77 (37\%) & 18 (35\%) & 5 (31\%) & 3 (43\%) \\ 
    Delay in treatment & 32 (16\%) & 12 (23\%) & 4 (25\%) & 1 (14\%) \\ 
    Documentation Issues & 6 (2.9\%) & 2 (3.8\%) & 1 (6.3\%) & 0 (0\%) \\ 
    Inadequate protocols & 5 (2.4\%) & 1 (1.9\%) & 0 (0\%) & 0 (0\%) \\ 
    Inadequate resources & 46 (22\%) & 9 (17\%) & 4 (25\%) & 2 (29\%) \\ 
    Missed diagnosis & 39 (19\%) & 8 (15\%) & 1 (6.3\%) & 0 (0\%) \\ 
    Other errors & 1 (0.5\%) & 2 (3.8\%) & 1 (6.3\%) & 1 (14\%) \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}n (\%)\\
\end{minipage}
\end{table}

Table 6 - Sub-OFI, classified according to SBP

\begin{table}[!t]
\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lcccc}
\toprule
\textbf{Characteristic} & \textbf{Class 1}  N = 247\textsuperscript{\textit{1}} & \textbf{Class 2}  N = 14\textsuperscript{\textit{1}} & \textbf{Class 3}  N = 5\textsuperscript{\textit{1}} & \textbf{Class 4}  N = 15\textsuperscript{\textit{1}} \\ 
\midrule\addlinespace[2.5pt]
OFI categories broad &  &  &  &  \\ 
    Clinical judgement error & 91 (37\%) & 3 (21\%) & 2 (40\%) & 7 (47\%) \\ 
    Delay in treatment & 39 (16\%) & 7 (50\%) & 1 (20\%) & 2 (13\%) \\ 
    Documentation Issues & 8 (3.2\%) & 1 (7.1\%) & 0 (0\%) & 0 (0\%) \\ 
    Inadequate protocols & 4 (1.6\%) & 1 (7.1\%) & 0 (0\%) & 1 (6.7\%) \\ 
    Inadequate resources & 57 (23\%) & 1 (7.1\%) & 0 (0\%) & 3 (20\%) \\ 
    Missed diagnosis & 44 (18\%) & 1 (7.1\%) & 2 (40\%) & 1 (6.7\%) \\ 
    Other errors & 4 (1.6\%) & 0 (0\%) & 0 (0\%) & 1 (6.7\%) \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}n (\%)\\
\end{minipage}
\end{table}

\hypertarget{main-results}{%
\subsection{Main results}\label{main-results}}

The adjusted analyses showed only statistical significance when
comparing BE class 4 to class 1, which had 63\% lower odds for OFI (OR
0,37; 95\% CI 0,14-0,83; p-value 0.026).

The un-adjusted analysis showed when comparing to class 1 in BE group,
class 2 (OR 1,61; 95\% CI 1,17-2,20; p-value 0.003) and 3 (OR 1,91; 95\%
CI 1,08-3,16; p-value 0.018) were significant in increased risk for OFI.
For the SBP group, comparison between class 1 and class 4 showed
significant increased risk (OR 2,28; 95\% CI 1,26-3,85; p-value 0.004).

Table 7 and 8 adjusted log

\begin{table}[!t]
\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lccccccc}
\toprule
 & \multicolumn{4}{c}{\textbf{Adjusted Model}} & \multicolumn{3}{c}{\textbf{Unadjusted Model}} \\ 
\cmidrule(lr){2-5} \cmidrule(lr){6-8}
\textbf{Characteristic} & \textbf{Event Risk} & \textbf{OR}\textsuperscript{\textit{1}} & \textbf{95\% CI}\textsuperscript{\textit{1}} & \textbf{p-value} & \textbf{OR}\textsuperscript{\textit{1}} & \textbf{95\% CI}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Shock classification - BE &  &  &  &  &  &  &  \\ 
    Class 1 & 206 / 3,989 (5.2\%) & — & — &  & — & — &  \\ 
    Class 2 & 52 / 644 (8.1\%) & 1.12 & 0.79, 1.55 & 0.5 & 1.61 & 1.17, 2.20 & {\bfseries 0.003} \\ 
    Class 3 & 16 / 170 (9.4\%) & 0.93 & 0.50, 1.62 & 0.8 & 1.91 & 1.08, 3.16 & {\bfseries 0.018} \\ 
    Class 4 & 7 / 116 (6.0\%) & 0.37 & 0.14, 0.83 & {\bfseries 0.026} & 1.18 & 0.49, 2.39 & 0.7 \\ 
Age (Years) & 281 / 4,919 (5.7\%) & 1.01 & 1.00, 1.01 & 0.10 &  &  &  \\ 
Gender (M/F) &  &  &  &  &  &  &  \\ 
    Female & 72 / 1,413 (5.1\%) & — & — &  &  &  &  \\ 
    Male & 209 / 3,506 (6.0\%) & 1.10 & 0.84, 1.48 & 0.5 &  &  &  \\ 
Pre-injury ASA &  &  &  &  &  &  &  \\ 
    1 & 158 / 3,032 (5.2\%) & — & — &  &  &  &  \\ 
    2 & 77 / 1,240 (6.2\%) & 1.05 & 0.77, 1.41 & 0.8 &  &  &  \\ 
    3 & 46 / 626 (7.3\%) & 1.19 & 0.79, 1.76 & 0.4 &  &  &  \\ 
    4 & 0 / 21 (0.00\%) & 0.00 &  & >0.9 &  &  &  \\ 
INR & 281 / 4,919 (5.7\%) & 1.15 & 0.78, 1.52 & 0.4 &  &  &  \\ 
Injury Severity Score & 281 / 4,919 (5.7\%) & 1.06 & 1.05, 1.07 & {\bfseries <0.001} &  &  &  \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}OR = Odds Ratio, CI = Confidence Interval\\
\end{minipage}
\end{table}
\begin{table}[!t]
\fontsize{12.0pt}{14.4pt}\selectfont
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lccccccc}
\toprule
 & \multicolumn{4}{c}{\textbf{Adjusted Model}} & \multicolumn{3}{c}{\textbf{Unadjusted Model}} \\ 
\cmidrule(lr){2-5} \cmidrule(lr){6-8}
\textbf{Characteristic} & \textbf{Event Risk} & \textbf{OR}\textsuperscript{\textit{1}} & \textbf{95\% CI}\textsuperscript{\textit{1}} & \textbf{p-value} & \textbf{OR}\textsuperscript{\textit{1}} & \textbf{95\% CI}\textsuperscript{\textit{1}} & \textbf{p-value} \\ 
\midrule\addlinespace[2.5pt]
Shock classification - SBP &  &  &  &  &  &  &  \\ 
    Class 1 & 247 / 4,455 (5.5\%) & — & — &  & — & — &  \\ 
    Class 2 & 14 / 239 (5.9\%) & 0.96 & 0.52, 1.64 & 0.9 & 1.06 & 0.58, 1.78 & 0.8 \\ 
    Class 3 & 5 / 98 (5.1\%) & 0.47 & 0.16, 1.13 & 0.13 & 0.92 & 0.32, 2.05 & 0.8 \\ 
    Class 4 & 15 / 127 (12\%) & 0.73 & 0.37, 1.35 & 0.3 & 2.28 & 1.26, 3.85 & {\bfseries 0.004} \\ 
Age (Years) & 281 / 4,919 (5.7\%) & 1.01 & 1.00, 1.01 & 0.076 &  &  &  \\ 
Gender (M/F) &  &  &  &  &  &  &  \\ 
    Female & 72 / 1,413 (5.1\%) & — & — &  &  &  &  \\ 
    Male & 209 / 3,506 (6.0\%) & 1.13 & 0.85, 1.51 & 0.4 &  &  &  \\ 
Pre-injury ASA &  &  &  &  &  &  &  \\ 
    1 & 158 / 3,032 (5.2\%) & — & — &  &  &  &  \\ 
    2 & 77 / 1,240 (6.2\%) & 1.03 & 0.76, 1.39 & 0.9 &  &  &  \\ 
    3 & 46 / 626 (7.3\%) & 1.17 & 0.77, 1.73 & 0.5 &  &  &  \\ 
    4 & 0 / 21 (0.00\%) & 0.00 &  & >0.9 &  &  &  \\ 
INR & 281 / 4,919 (5.7\%) & 1.14 & 0.76, 1.52 & 0.4 &  &  &  \\ 
Injury Severity Score & 281 / 4,919 (5.7\%) & 1.06 & 1.05, 1.07 & {\bfseries <0.001} &  &  &  \\ 
\bottomrule
\end{tabular*}
\begin{minipage}{\linewidth}
\textsuperscript{\textit{1}}OR = Odds Ratio, CI = Confidence Interval\\
\end{minipage}
\end{table}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusion-1}{%
\section{Conclusion}\label{conclusion-1}}

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-noauthor_injuries_nodate}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Injuries and violence {[}Internet{]}. {[}cited 2024 Sep
10{]}. Available from:
\url{https://www.who.int/news-room/fact-sheets/detail/injuries-and-violence}}

\leavevmode\vadjust pre{\hypertarget{ref-kauvar_impact_2006}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Kauvar DS, Lefering R, Wade CE. Impact of {Hemorrhage}
on {Trauma} {Outcome}: {An} {Overview} of {Epidemiology}, {Clinical}
{Presentations}, and {Therapeutic} {Considerations}. Journal of Trauma
and Acute Care Surgery {[}Internet{]}. 2006 Jun {[}cited 2024 Sep
19{]};60(6):S3. Available from:
\url{https://journals.lww.com/jtrauma/fulltext/2006/06001/impact_of_hemorrhage_on_trauma_outcome__an.2.aspx}}

\leavevmode\vadjust pre{\hypertarget{ref-berry_shock_2015}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Berry S. Shock {Management} in {Trauma}. In: Papadakos
PJ, Gestring ML, editors. Encyclopedia of {Trauma} {Care}
{[}Internet{]}. Berlin, Heidelberg: Springer; 2015 {[}cited 2024 Sep
20{]}. p. 1484--8. Available from:
\url{https://doi.org/10.1007/978-3-642-29613-0_505}}

\leavevmode\vadjust pre{\hypertarget{ref-teixeira_preventable_2007}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Teixeira PGR, Inaba K, Hadjizacharia P, Brown C, Salim
A, Rhee P, et al.
\href{https://doi.org/10.1097/TA.0b013e31815078ae}{Preventable or
potentially preventable mortality at a mature trauma center}. The
Journal of Trauma. 2007 Dec;63(6):1338-1346; discussion 1346-1347. }

\leavevmode\vadjust pre{\hypertarget{ref-noauthor_atls_2018}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{{ATLS} advanced trauma life support {[}Internet{]}. 10th
ed. Vol. 2018. Chicago: American College of Surgeons; 2018 {[}cited 2024
Sep 22{]}. Available from:
\url{https://cirugia.facmed.unam.mx/wp-content/uploads/2018/07/Advanced-Trauma-Life-Support.pdf}}

\leavevmode\vadjust pre{\hypertarget{ref-el-menyar_review_2019}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{El-Menyar A, Mekkodathil A, Abdelrahman H, Latifi R,
Galwankar S, Al-Thani H, et al. Review of {Existing} {Scoring} {Systems}
for {Massive} {Blood} {Transfusion} in {Trauma} {Patients}: {Where} {Do}
{We} {Stand}? Shock {[}Internet{]}. 2019 Sep {[}cited 2024 Oct
24{]};52(3):288. Available from:
\url{https://journals.lww.com/shockjournal/fulltext/2019/09000/review_of_existing_scoring_systems_for_massive.2.aspx}}

\leavevmode\vadjust pre{\hypertarget{ref-mackersie_pitfalls_2010}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Mackersie RC. Pitfalls in the {Evaluation} and
{Resuscitation} of the {Trauma} {Patient}. Emergency Medicine Clinics of
North America {[}Internet{]}. 2010 Feb {[}cited 2024 Sep
10{]};28(1):1--27. Available from:
\url{https://www.sciencedirect.com/science/article/pii/S0733862709001199}}

\leavevmode\vadjust pre{\hypertarget{ref-ivatury_patient_2008}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Ivatury RR, Guilford K, Malhotra AK, Duane T, Aboutanos
M, Martin N. \href{https://doi.org/10.1097/TA.0b013e318163359d}{Patient
safety in trauma: Maximal impact management errors at a level {I} trauma
center}. The Journal of Trauma. 2008 Feb;64(2):265-270; discussion
270-272. }

\leavevmode\vadjust pre{\hypertarget{ref-world_health_organization_guidelines_2009}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{World Health Organization. Guidelines for trauma quality
improvement programmes. 2009 {[}cited 2024 Sep 20{]};104. Available
from: \url{https://iris.who.int/handle/10665/44061}}

\leavevmode\vadjust pre{\hypertarget{ref-juillard_establishing_2009}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Juillard CJ, Mock C, Goosen J, Joshipura M, Civil I.
\href{https://doi.org/10.1007/s00268-009-9959-8}{Establishing the
evidence base for trauma quality improvement: A collaborative
{WHO}-{IATSIC} review}. World Journal of Surgery. 2009
May;33(5):1075--86. }

\leavevmode\vadjust pre{\hypertarget{ref-hashmi_hospital-based_2013}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Hashmi ZG, Haider AH, Zafar SN, Kisat M, Moosa A,
Siddiqui F, et al.
\href{https://doi.org/10.1097/TA.0b013e31829880a0}{Hospital-based trauma
quality improvement initiatives: First step toward improving trauma
outcomes in the developing world}. The Journal of Trauma and Acute Care
Surgery. 2013 Jul;75(1):60--68; discussion 68. }

\leavevmode\vadjust pre{\hypertarget{ref-donabedian_quality_1988}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Donabedian A. The {Quality} of {Care}: {How} {Can} {It}
{Be} {Assessed}? JAMA {[}Internet{]}. 1988 Sep {[}cited 2024 Sep
19{]};260(12):1743--8. Available from:
\url{https://doi.org/10.1001/jama.1988.03410120089033}}

\leavevmode\vadjust pre{\hypertarget{ref-vang_shock_2022}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Vang M, Østberg M, Steinmetz J, Rasmussen LS.
\href{https://doi.org/10.1007/s00068-022-01932-z}{Shock index as a
predictor for mortality in trauma patients: A systematic review and
meta-analysis}. European Journal of Trauma and Emergency Surgery:
Official Publication of the European Trauma Society. 2022
Aug;48(4):2559--66. }

\leavevmode\vadjust pre{\hypertarget{ref-tisherman_detailed_2015}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Tisherman SA, Schmicker RH, Brasel KJ, Bulger EM, Kerby
JD, Minei JP, et al.
\href{https://doi.org/10.1097/SLA.0000000000000837}{Detailed description
of all deaths in both the shock and traumatic brain injury hypertonic
saline trials of the {Resuscitation} {Outcomes} {Consortium}}. Annals of
Surgery. 2015 Mar;261(3):586--90. }

\leavevmode\vadjust pre{\hypertarget{ref-eastridge_hypotension_2007}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Eastridge BJ, Salinas J, McManus JG, Blackburn L, Bugler
EM, Cooke WH, et al.
\href{https://doi.org/10.1097/TA.0b013e31809ed924}{Hypotension begins at
110 mm {Hg}: Redefining "hypotension" with data}. The Journal of Trauma.
2007 Aug;63(2):291-297; discussion 297-299. }

\leavevmode\vadjust pre{\hypertarget{ref-oyetunji_redefining_2011}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Oyetunji TA, Chang DC, Crompton JG, Greene WR, Efron DT,
Haut ER, et al.
\href{https://doi.org/10.1001/archsurg.2011.154}{Redefining hypotension
in the elderly: Normotension is not reassuring}. Archives of Surgery
(Chicago, Ill: 1960). 2011 Jul;146(7):865--9. }

\leavevmode\vadjust pre{\hypertarget{ref-mutschler_critical_2013}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Mutschler M, Nienaber U, Brockamp T, Wafaisade A, Wyen
H, Peiniger S, et al. A critical reappraisal of the {ATLS}
classification of hypovolaemic shock: {Does} it really reflect clinical
reality? Resuscitation {[}Internet{]}. 2013 Mar {[}cited 2024 Oct
24{]};84(3):309--13. Available from:
\url{https://www.sciencedirect.com/science/article/pii/S0300957212003693}}

\leavevmode\vadjust pre{\hypertarget{ref-mutschler_renaissance_2013}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Mutschler M, Nienaber U, Brockamp T, Wafaisade A, Fabian
T, Paffrath T, et al. Renaissance of base deficit for the initial
assessment of trauma patients: A base deficit-based classification for
hypovolemic shock developed on data from 16,305 patients derived from
the {TraumaRegister} {DGU}®. Critical Care {[}Internet{]}. 2013 Mar
{[}cited 2024 Oct 24{]};17(2):R42. Available from:
\url{https://pmc.ncbi.nlm.nih.gov/articles/PMC3672480/}}

\end{CSLReferences}

\end{document}

### 2024-11-28 thesis.Rmd
lite ändringar i metod och lagt till en mer korrekt version av första meningen för diskussion.

@@
 ## Main results
-The adjusted analyses showed only statistical significance when comparing BE class 4 to class 1, which had 63% lower odds for OFI (OR 0,37; 95% CI 0,14-0,83; p-value 0.026).
+The adjusted analyses showed only statistical significance when comparing BE class 4 to class 1, which had 63% reduced odds for OFI (OR 0,37; 95% CI 0,14-0,83; p-value 0.026).
 
-The un-adjusted analysis showed when comparing to class 1 in BE group, class 2 (OR 1,61; 95% CI 1,17-2,20; p-value 0.003) and 3 (OR 1,91; 95% CI 1,08-3,16; p-value 0.018) were significant in increased risk for OFI.
-For the SBP group, comparison between class 1 and class 4 showed significant increased risk (OR 2,28; 95% CI 1,26-3,85; p-value 0.004).
+The un-adjusted analysis showed when comparing to class 1 in BE group, class 2 and 3 increased odds for OFI by 61% (OR 1,61; 95% CI 1,17-2,20; p-value 0.003) respective 91% (OR 1,91; 95% CI 1,08-3,16; p-value 0.018).
+The SBP group showed, when comparing class 4 to class 1, there were increased odds of 128% (OR 2,28; 95% CI 1,26-3,85; p-value 0.004).
 
@@
 ```
+## Key results
+We found that the most common type of OFI is Clinical judgement error across almost all shock severities and that there are minor if any association between shock severity and OFI.
+
+The most common type of OFI is clinical judgement error, it is the most common in all BE shock classes and 2/4 classes for SBP classes.
+is around ...% of.
+It generally takes up ...% of all OFI defined as shock (class 2 and below).
+
+When OFI and shock severity are analysed alone, there are statistical significant correlation for these classes: ...
+.
+It shows positive correlation where the more severe the shock is, the higher the odds are for an OFI to occur.
+This in constrast with the adjusted results which showed inverted correlation betwen OFI and shock severity, with 63% lower odds for OFI when comparing the BE class 4 with class 1.
+However there is no correlation for the other classes nor the classes for the SBP group.
+
+When analysing the other adjusted factors, ISS seems to be a major confounder which correlates to both OFI and shock severity.
+This can explain why the results of the adjusted and unadjusted analyses are inverted.
+
+## Limitations
+
+## Clinical applications
+
+## Future studies
 This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.

### 2024-11-29 manuscript.Rmd
Modify manuscript to create PDF and fix table captions

--- 2024-11-27 manuscript.Rmd
@@
 ---
-title: "Opportunities for improvement in the care of adult trauma patients arriving in shock"
+title: Opportunities for improvement in the care of adult trauma patients arriving
+  in shock
 author: "Carl Skog"
 output:
-  html_document: default
-  word_document: default
+  # word_document: default
+  bookdown::pdf_document2:
+    latex_engine: xelatex
+  # html_document: default
 bibliography: bibliography.bib
@@
 Out of the clinical needs to assess the severity of trauma shock, patients can be categorized by stratifying symptoms and physiological parameters into degrees.
-Many different scoring systems have been developed, such as ATLS classification for hemorrhagic shock (See Table \@ref(tab:alts) [@noauthor_atls_2018], Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion [@el-menyar_review_2019].
+Many different scoring systems have been developed, such as ATLS classification for hemorrhagic shock (See Table \@ref(tab:atls) [@noauthor_atls_2018]), Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion [@el-menyar_review_2019].
 While none of these systems comprehensively classifies shock severity across all types of trauma shock, each aid in assessing severity through core physiological indicators.
 Ultimately, these classifications rely on vital signs and clinical indicators, highlighting the importance of physiological metrics in evaluating shock severity across clinical settings.
+
+Table: (\#tab:atls) ATLS classification for haemorrhagic shock, adapted from [@noauthor_atls_2018].
 
@@
 | Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
-
-Table: ATLS classification for haemorrhagic shock, adapted from [@noauthor_atls_2018] {#tbl:atls}.
 
@@
 
+Table \@ref(tab:mod-atls) shows how we defined shock according to BE and SBP.
 The BE parameter was divided into four classes with these cutoffs: Class I (above -2), Class II (-2 to -6), Class III (-6 to -10), and Class IV (below -10).
@@
 
-**ATLS tenth edition - Systolic blood pressure and Base excess (BD)**
+Table: (\#tab:mod-atls) ATLS tenth edition - Systolic blood pressure and Base excess (BD).
 
@@
 ## Descriptive data
+Table \@ref(tab:sample-characteristics-reg) describes the characteristics of the study population.
 The study demographics median age is `r age` and the most common gender is male, `r male`.
@@
 
-Table 4 - Sample characteristics used in regression.
-```{r sample-characteristics BE, echo=FALSE}
-sample.characteristics.table_reg
+```{r sample-characteristics-reg, echo=FALSE}
+sample.characteristics.table_reg |>
+  modify_caption("Sample characteristics used in regression.")
 ```

### 2024-11-29 manuscript.Rmd
Martins reviderade intro med oreilley

@@
 # Introduction
+Trauma is a major global public health concern.
+It causes over four million deaths, among those younger populations are the most affected[@noauthor_injuries_nodate].
+Beyond the high mortality rate, trauma also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs[@noauthor_injuries_nodate].
 
-## Epidemiology
-Trauma is a major global public health concern.
-It causes over four million deaths, among those younger populations are the most affected [@noauthor_injuries_nodate].
-Beyond the high mortality rate, trauma also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs [@noauthor_injuries_nodate].
-Hemorrhagic shock, the second most common cause of death following trauma, is the leading preventable cause of death within the first 24 hours post-injury  [@kauvar_impact_2006;@berry_shock_2015;@teixeira_preventable_2007].
+The initial management of trauma patients is highly time-sensitive and prone to error [@mackersie_pitfalls_2010;@teixeira_preventable_2007;@ivatury_patient_2008].
+Quality Improvement (QI) initiatives address these challenges by systematically evaluating care processes and outcomes, aiming to reduce morbidity and mortality [@world_health_organization_guidelines_2009].
 
-## Shock classifications
-Out of the clinical needs to assess the severity of trauma shock, patients can be categorized by stratifying symptoms and physiological parameters into degrees.
-Many different scoring systems have been developed, such as ATLS classification for hemorrhagic shock (See Table \@ref(tab:atls) [@noauthor_atls_2018]), Trauma Associated Severe Hemorrhage (TASH) and Assessment of Blood Consumption (ABC) scores for predicting massive transfusion [@el-menyar_review_2019].
-While none of these systems comprehensively classifies shock severity across all types of trauma shock, each aid in assessing severity through core physiological indicators.
-Ultimately, these classifications rely on vital signs and clinical indicators, highlighting the importance of physiological metrics in evaluating shock severity across clinical settings.
+Hemorrhagic shock is, the second most common cause of death following trauma and it, is the leading preventable cause of death within the first 24 hours post-injury [@kauvar_impact_2006;@berry_shock_2015;@teixeira_preventable_2007], and opportunities for improvement are common in patients who die because of massive traumatic bleeding [@oreilly_opportunities_2013].
+
+This may indicate that these preventable events potentially leading to suboptimal outcomes are common also in patients who survive massive bleeding.
+This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
+This knowledge would help clinicians avoid such events in bleeding trauma patients.
+
+(See Table \@ref(tab:atls) [@noauthor_atls_2018])
 
@@
 | Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
-
-## Trauma quality improvement and opportunities for improvement
-The initial management of trauma patients is highly time-sensitive and prone to error  [@mackersie_pitfalls_2010;@teixeira_preventable_2007;@ivatury_patient_2008].
-Quality Improvement (QI) initiatives address these challenges by systematically evaluating care processes and outcomes, aiming to reduce morbidity and mortality [@world_health_organization_guidelines_2009].
-Through structured quality measures and collaborative reviews, QI has strengthened trauma care and improved outcomes globally  [@juillard_establishing_2009;@hashmi_hospital-based_2013].
-Identifying Opportunities for Improvement (OFI) offers a broader perspective than traditional mortality reviews by also including non-fatal outcomes [@world_health_organization_guidelines_2009].
-Structured multidisciplinary morbidity and mortality (M&M) reviews are commonly used to identify OFIs.
-One systemic and comprehensive way to identify OFI is using the Donabedian framework, which asses the quality of care by the three categories of: structure, process, and outcome  [@donabedian_quality_1988;@world_health_organization_guidelines_2009].
-
-## Shock and OFI
-As mentioned before hemorrhagic shock is the leading cause of preventable death within the first 24 hours of injury [@berry_shock_2015].
-Patients in shock have a higher mortality rate than those not [@vang_shock_2022], with a median time to death of just 2 hours [@tisherman_detailed_2015].
-Despite this, the relationship between shock, its severity, and Opportunities for Improvement (OFI) remains largely unexamined.
-This knowledge would help clinicians better identify and avoid preventable mistakes in shock patients who need urgent care.
-
-## Aims
-This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
 

### 2024-11-29 manuscript.Rmd
flyttade på ATLS tabellen

@@
 
-(See Table \@ref(tab:atls) [@noauthor_atls_2018])
-
-Table: (\#tab:atls) ATLS classification for haemorrhagic shock, adapted from [@noauthor_atls_2018].
-
-| **Parameter**                     | **Class I** | **Class II** (Mild) | **Class III** (Moderate)  | **Class IV** (Severe)         |
-|-----------------------------------|-------------|---------------------|---------------------------|-------------------------------|
-| Approximate blood loss            | <15%        | 15–30%              | 31–40%                    | >40%                          |
-| Heart rate                        | ↔           | ↔/↑                 | ↑                         | ↑/↑↑                          |
-| Blood pressure                    | ↔           | ↔                   | ↔/↓                       | ↓                             |
-| Pulse pressure                    | ↔           | ↓                   | ↓                         | ↓                             |
-| Respiratory rate                  | ↔           | ↔                   | ↔/↑                       | ↑                             |
-| Urine output                      | ↔           | ↔                   | ↓                         | ↓↓                            |
-| Glasgow Coma Scale score          | ↔           | ↔                   | ↓                         | ↓                             |
-| Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L    | –6 to –10 mEq/L           | –10 mEq/L or less             |
-| Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
-
 # Methods
@@
 The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure (SBP) and once with base excess (BE), regardless of their cause of shock.
-Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications.
+Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications (See Table \@ref(tab:atls) [@noauthor_atls_2018]).
+
+Table: (\#tab:atls) ATLS classification for haemorrhagic shock, adapted from [@noauthor_atls_2018].
+
+| **Parameter**                     | **Class I** | **Class II** (Mild) | **Class III** (Moderate)  | **Class IV** (Severe)         |
+|-----------------------------------|-------------|---------------------|---------------------------|-------------------------------|
+| Approximate blood loss            | <15%        | 15–30%              | 31–40%                    | >40%                          |
+| Heart rate                        | ↔           | ↔/↑                 | ↑                         | ↑/↑↑                          |
+| Blood pressure                    | ↔           | ↔                   | ↔/↓                       | ↓                             |
+| Pulse pressure                    | ↔           | ↓                   | ↓                         | ↓                             |
+| Respiratory rate                  | ↔           | ↔                   | ↔/↑                       | ↑                             |
+| Urine output                      | ↔           | ↔                   | ↓                         | ↓↓                            |
+| Glasgow Coma Scale score          | ↔           | ↔                   | ↓                         | ↓                             |
+| Base deficit                      | 0 to –2 mEq/L | –2 to –6 mEq/L    | –6 to –10 mEq/L           | –10 mEq/L or less             |
+| Need for blood products           | Monitor     | Possible            | Yes                       | Massive Transfusion Protocol  |
 

### 2024-11-29 main.R
La till ofi som factor istället och kortat ner koden lite

@@ -81,8 +81,8 @@ convert_number <- function(x) {
   return(x)
 }
 
-# Make ofi numerical
-study.sample$ofinum <- ifelse(study.sample$ofi == "Yes", 1, 0)
+# OFI as factor
+study.sample$ofi <- as.factor(study.sample$ofi)
 
 # Converting ed_be_art to numeric
 BEnum <- convert_number(study.sample$ed_be_art)
@@ -112,6 +112,15 @@ ed_inr_numeric <- convert_number(study.sample$ed_inr)
 # Re-add INR as numeric to `study.sample`
 study.sample <- study.sample %>%
   mutate(ed_inr_numeric = ed_inr_numeric)
+
+# Remove no longer used variables - study sample
+# I suggest removing them from the lines 40-52 where the study data is created instead
+study.sample <- study.sample |>
+  select(
+    -ed_be_art,
+    -ed_inr,
+    -Deceased
+  )
 
 # BE shock classification
 study.sample <- study.sample %>%
@@ -148,7 +157,7 @@ log_reg_count <- nrow(reg.sample)
 
 # Binary logistic regression model - BE
 # Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
-log_regBE <- glm(ofinum ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
+log_regBE <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
                  family = binomial,
                  data = reg.sample
 )
@@ -162,7 +171,7 @@ log_regBE <- glm(ofinum ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury +
 #ISS ~ "Injury Severity Score",
 
 #test 
-test1 <- glm(ofinum ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric  + ISS + V4SBP_class,
+test1 <- glm(ofi ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric  + ISS + V4SBP_class,
                  family = binomial,
                  data = reg.sample
 )
@@ -186,42 +195,22 @@ test <- tbl_regression(
 
 
 # Binary logistic regression model - SBP
[…]

### 2024-11-29 thesis.Rmd
gjort step analys, fast kortade ner det hela i 2 steg, även noterat fel i reg mellan study.sample och reg.sample

@@
 ## Key results
-We found that the most common type of OFI is Clinical judgement error across almost all shock severities and that there are minor if any association between shock severity and OFI.
+We found that the most common type of OFI is clinical judgement error across almost all shock severities and there is only statistical significant association between severe shock with a BE of < -10 and OFI.
 

### 2024-11-29 main.R

@@ -165,31 +165,71 @@ log_regBE <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed
 
 
 
-#   
-#pt_asa_preinjury ~ "Pre-injury ASA",
-#ed_inr_numeric ~ "INR",
-#ISS ~ "Injury Severity Score",
-
-#test 
-test1 <- glm(ofi ~ pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric  + ISS + V4SBP_class,
+# step analysis - SBP 
+SBPstep1d <- glm(ofi ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric,
                  family = binomial,
                  data = reg.sample
 )
 
-test <- tbl_regression(
-  test1,
+SBPstep1 <- tbl_regression(
+  SBPstep1d,
   exponentiate = TRUE,
   label = list(
     pt_age_yrs ~ "Age (Years)",
     pt_Gender ~ "Gender (M/F)",
-    V4SBP_class ~ "Shock classification"
+    V4SBP_class ~ "Shock classification - SBP"
   )
 ) |>
-  add_nevent() |>
   bold_p()
 
-#print(test)
+SBPstep2d <- glm(ofi ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric  + ISS,
+                 family = binomial,
+                 data = reg.sample
+)
 
+SBPstep2 <- tbl_regression(
+  SBPstep2d,
+  exponentiate = TRUE,
+  label = list(
+    pt_age_yrs ~ "Age (Years)",
+    pt_Gender ~ "Gender (M/F)",
+    V4SBP_class ~ "Shock classification - SBP"
+  )
+) |>
+  bold_p()
+
+# step analysis - BE
+BEstep1d <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric,
+                 family = binomial,
+                 data = study.sample
+)
+
+BEstep1 <- tbl_regression(
+  BEstep1d,
+  exponentiate = TRUE,
+  label = list(
+    pt_age_yrs ~ "Age (Years)",
+    pt_Gender ~ "Gender (M/F)",
+    BE_class ~ "Shock classification - BE"
+  )
+) |>
+  bold_p()
+
[…]

### 2024-11-30 main.R
ändrade till rätt tabeller

@@ -201,7 +201,7 @@ SBPstep2 <- tbl_regression(
 # step analysis - BE
 BEstep1d <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric,
                  family = binomial,
-                 data = study.sample
+                 data = reg.sample
 )
 
 BEstep1 <- tbl_regression(
@@ -217,7 +217,7 @@ BEstep1 <- tbl_regression(
 
 BEstep2d <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric  + ISS,
                  family = binomial,
-                 data = study.sample
+                 data = reg.sample
 )
 
 BEstep2 <- tbl_regression(
@@ -229,7 +229,7 @@ BEstep2 <- tbl_regression(
     BE_class ~ "Shock classification - BE"
   )
 ) |>
-  bold_p()
+  bold_p() 
 
 
 
@@ -422,16 +422,16 @@ combined_tableSBP <- tbl_merge(
 ) 
 
 #combined step reg
-combined_table_stepSBP <- tbl_merge(
-  tbls = list(log_regSBP_sample.characteristics.table_unadjusted, SBPstep1, SBPstep2),
-  tab_spanner = c("**Unadjusted**","**without ISS**", "**with ISS**")
-) 
-
-print(combined_table_stepSBP)
-
 combined_table_stepBE <- tbl_merge(
   tbls = list(log_regBE_sample.characteristics.table_unadjusted, BEstep1, BEstep2),
   tab_spanner = c("**Unadjusted**","**without ISS**", "**with ISS**")
 ) 
 
 print(combined_table_stepBE)
+
+combined_table_stepSBP <- tbl_merge(
+  tbls = list(log_regSBP_sample.characteristics.table_unadjusted, SBPstep1, SBPstep2),
+  tab_spanner = c("**Unadjusted**","**without ISS**", "**with ISS**")
+) 
+
+print(combined_table_stepSBP)

### 2024-11-30 main.R
la till tabellen i metod på rätt plats

@@ -427,11 +427,7 @@ combined_table_stepBE <- tbl_merge(
   tab_spanner = c("**Unadjusted**","**without ISS**", "**with ISS**")
 ) 
 
-print(combined_table_stepBE)
-
 combined_table_stepSBP <- tbl_merge(
   tbls = list(log_regSBP_sample.characteristics.table_unadjusted, SBPstep1, SBPstep2),
   tab_spanner = c("**Unadjusted**","**without ISS**", "**with ISS**")
 ) 
-
-print(combined_table_stepSBP)

### 2024-11-30 thesis.Rmd
la till tabellen i metod på rätt plats
början av diskussionen

--- 2024-11-28 thesis.Rmd
@@
 
+When doing the regression in steps, it can be observed that the biggest changes of OR occurs when adding the ISS parameter into the regression model both for the BE and SBP groups.
+
+```{r Step log, echo=FALSE}
+combined_table_stepBE
+combined_table_stepSBP
+```
+
 # Discussion
@@
 ## Key results
-We found that the most common type of OFI is Clinical judgement error across almost all shock severities and that there are minor if any association between shock severity and OFI.
+We found that the most common type of OFI is clinical judgement error across almost all shock severities.
+Out of all shock severities, there is only statistical significant association between severe shock with a BE of < -10 and OFI, which showes a 63% reduced odds of OFI when comparing against those with no shock.
+
+Previous studies [@teixeira_preventable_2007] which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment (52,9%) while the second most common is clinical judgement error (21,6%).
+In comparison, our results showed the most common OFI is clinical judgement error (36%), while inadequate was second (22%) and delay in treatment comes third (17%).
+For patients defined as in shock, the most common was still clinical judgement error (35%).
 

### 2024-12-02 thesis.Rmd
sparar från förra gången, tror kommit fram till skriva om shock severity och ofi typ

--- 2024-11-28 thesis.Rmd
@@
 ## Key results
-We found that the most common type of OFI is Clinical judgement error across almost all shock severities and that there are minor if any association between shock severity and OFI.
+We found that the most common type of OFI is clinical judgement error across almost all shock severities.
+Out of all shock severities, there is only statistical significant association between severe shock with a BE of < -10 and OFI, which shows a 63% reduced odds of OFI when comparing against those with no shock.
+
+Previous studies [@teixeira_preventable_2007;@saltzherr_preventability_2011] which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment, while the second most common is clinical judgement error [@teixeira_preventable_2007], or errors in management[@saltzherr_preventability_2011].
+In comparison, our results showed the most common OFI is clinical judgement error (36%), while inadequate resources was second (22%) and delay in treatment comes third (17%), for the whole study population.
+These differences could be due to the nature of the studied population, where delay in treatment may be more deadly, therefore it takes up a smaller part in our results, while clinical judgement error takes a larger part.
+(KÄLLA)
+
+For patients defined as in shock (Class 2-4), the most common was still clinical judgement error (BE 34,7%; SBP 35,3%).
+This result is similar when compared to D O'reilly's study [@oreilly_opportunities_2013], in which the most common type of opportunities for performance improvements (OPI) was ones involved with decision making at 63 out of 150 OPIs.
+
+Comparing
 

### 2024-12-04 thesis.Rmd
sista save innan föra över till google doc

@@
 
-The most common type of OFI is clinical judgement error, it is the most common in all BE shock classes and 2/4 classes for SBP classes.
-is around ...% of.
-It generally takes up ...% of all OFI defined as shock (class 2 and below).
+## Strengths and limitations
+One of the limitations is lack of data to design the study around a modern shock/hemorrhage classification which uses combination of base parameters,
 
-When OFI and shock severity are analysed alone, there are statistical significant correlation for these classes: ...
-.
-It shows positive correlation where the more severe the shock is, the higher the odds are for an OFI to occur.
-This in constrast with the adjusted results which showed inverted correlation betwen OFI and shock severity, with 63% lower odds for OFI when comparing the BE class 4 with class 1.
-However there is no correlation for the other classes nor the classes for the SBP group.
-
-When analysing the other adjusted factors, ISS seems to be a major confounder which correlates to both OFI and shock severity.
-This can explain why the results of the adjusted and unadjusted analyses are inverted.
-
-## Limitations
-
-## Clinical applications
+## Clinical/Practical applications
 
@@
 
+------------------------------------------------------------------------------
 Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
-Results from the BE approximation done by Mutschler et al [mutschler_renaissance_2013] which is the same as the tenth edition ATLS shock classification, shows that class I had a mean SBP of 132,6 mmhg and class IV have 94,8 mmhg at the emergency department.
+
+Results from the BE approximation done by Mutschler et al [@mutschler_renaissance_2013] which is the same as the tenth edition ATLS shock classification, shows that class I had a mean SBP of 132,6 mmhg and class IV have 94,8 mmhg at the emergency department.
 As a SBP of 132,6 is at the higher range of normotensive, we will not be able to use it as a upper limit, but the 94,8 mmhg is close to our approximation of 90mmhg for class IV.
@@
 The first one is ATLS tenth edition lack numerical values for vital parameters.
-The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
-The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
+The second one is that base deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
+The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes at the same time and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
 

### 2024-12-05 main.R
Justerat om ordning och tagit bort tomma platser

@@ -162,8 +162,23 @@ log_regBE <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed
                  data = reg.sample
 )
 
+# Binary logistic regression model - SBP
+log_regSBP <- glm(ofi ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
+                  family = binomial,
+                  data = reg.sample
+)
 
+# Binary logistic regression model unadjusted - BE
+log_regBEun <- glm(ofi ~ BE_class,
+                   family = binomial,
+                   data = reg.sample
+)
 
+# Binary logistic regression model unadjusted - SBP
+log_regSBPun <- glm(ofi ~ V4SBP_class,
+                    family = binomial,
+                    data = reg.sample
+)
 
 # step analysis - SBP 
 SBPstep1d <- glm(ofi ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric,
@@ -230,27 +245,6 @@ BEstep2 <- tbl_regression(
   )
 ) |>
   bold_p() 
-
-
-
-
-# Binary logistic regression model - SBP
-log_regSBP <- glm(ofi ~ V4SBP_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
-  family = binomial,
-  data = reg.sample
-)
-
-# Binary logistic regression model unadjusted - BE
-log_regBEun <- glm(ofi ~ BE_class,
-  family = binomial,
-  data = reg.sample
-)
-
-# Binary logistic regression model unadjusted - SBP
-log_regSBPun <- glm(ofi ~ V4SBP_class,
-  family = binomial,
-  data = reg.sample
-)
 
 # Label variables
 var_label(study.sample$pt_age_yrs) <- "Age (Years)"

### 2024-12-05 main.R
gamla kommentarer raderade

@@ -114,7 +114,6 @@ study.sample <- study.sample %>%
   mutate(ed_inr_numeric = ed_inr_numeric)
 
 # Remove no longer used variables - study sample
-# I suggest removing them from the lines 40-52 where the study data is created instead
 study.sample <- study.sample |>
   select(
     -ed_be_art,
@@ -156,7 +155,6 @@ reg.sample <- study.sample %>%
 log_reg_count <- nrow(reg.sample)
 
 # Binary logistic regression model - BE
-# Note that you cannot adjust for both ways to define shock in the same model, because they are just different ways to define the same thing. I suggest you create separate models for each.
 log_regBE <- glm(ofi ~ BE_class + pt_age_yrs + pt_Gender + pt_asa_preinjury + ed_inr_numeric + ISS,
                  family = binomial,
                  data = reg.sample

### 2024-12-06 main.R
La till ISS confounding tabellen

@@ -423,3 +423,10 @@ combined_table_stepSBP <- tbl_merge(
   tbls = list(log_regSBP_sample.characteristics.table_unadjusted, SBPstep1, SBPstep2),
   tab_spanner = c("**Unadjusted**","**without ISS**", "**with ISS**")
 ) 
+
+# Sample characteristics with shock classification.
+sample.characteristics.table_ISS <- tbl_summary(reg.sample,
+                                            by = BE_class
+) |>
+  add_overall() |>
+  add_p()

### 2024-12-10 thesis.Rmd
La till ISS confounding tabellen
la in första version av discussion

--- 2024-11-30 thesis.Rmd
@@
 
+```{r ISS confounding, echo=FALSE}
+sample.characteristics.table_ISS
+```
+
 # Discussion
@@
 ## Key results
-We found that the most common type of OFI is clinical judgement error across almost all shock severities.
-Out of all shock severities, there is only statistical significant association between severe shock with a BE of < -10 and OFI, which showes a 63% reduced odds of OFI when comparing against those with no shock.
+We found that the most common type of OFI is clinical judgment error across almost all shock severities.
+Out of all shock severities, there is only a statistically significant association between severe shock with a BE of < -10 and OFI, which shows a 63% reduced odds of OFI when compared against those with no shock.
 
-Previous studies [@teixeira_preventable_2007] which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment (52,9%) while the second most common is clinical judgement error (21,6%).
-In comparison, our results showed the most common OFI is clinical judgement error (36%), while inadequate was second (22%) and delay in treatment comes third (17%).
-For patients defined as in shock, the most common was still clinical judgement error (35%).
+The unadjusted analyses demonstrated a positive correlation between more severe shock and higher odds of OFI.
+However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI.
+This shift can be explained by the confounding effect of ISS, which is significantly associated with both shock severity and OFI (p < 0.001).
+Adjusting for ISS revealed that the initial positive correlation was likely driven by shared variability with ISS.
 
-The most common type of OFI is clinical judgement error, it is the most common in all BE shock classes and 2/4 classes for SBP classes.
-is around ...% of.
-It generally takes up ...% of all OFI defined as shock (class 2 and below).
+Association between OFI and ISS can also be observed in a study by Husssein et al [@albaaj_patient_2023].
+It is important to note that Hussein's study was also conducted at Karolinska Solna, but during a different time frame (2017-2021) and with other inclusion and exclusion factors.
+Correlation between ISS and BD or SBP was also shown in two studies by Mutscher et al.
+[@mutschler_critical_2013;@mutschler_renaissance_2013].
 
-When OFI and shock severity are analysed alone, there are statistical significant correlation for these classes: ...
-.
-It shows positive correlation where the more severe the shock is, the higher the odds are for an OFI to occur.
-This in constrast with the adjusted results which showed inverted correlation betwen OFI and shock severity, with 63% lower odds for OFI when comparing the BE class 4 with class 1.
-However there is no correlation for the other classes nor the classes for the SBP group.
+### Interpretation and generalisability
+#### Reduced OR of BE class 4
+Patients with BE < -10, indicating severe shock, were found to have lower odds of OFI compared to other groups, and two hypothesis may explain this finding.
 
-When analysing the other adjusted factors, ISS seems to be a major confounder which correlates to both OFI and shock severity.
-This can explain why the results of the adjusted and unadjusted analyses are inverted.
+First, patients in severe shock should be more easily recognized as critically ill upon arrival, triggering heightened clinical attention and prioritization in triage.
+This results in more structured management and often receives priority in resource allocation, which may minimize variability in care and reduce the likelihood of errors.
 
-## Limitations
+Another explanation is the potential influence of survival bias.
+Patients with BE < -10 have high mortality rates [@mutschler_renaissance_2013] , and those who survive may represent cases where care was particularly effective, or where errors were less impactful.
 
-## Clinical applications
+#### Distribution of significance/insignificant results
+When looking at the distribution of the significant and insignificant results from the “step reg” table, we can notice for the BE group, that the significance reverses when adding ISS.
+This may be a sign that the significant unadjusted result is purely because of the ISS which we discussed before.
+The same can also be said about the OR, as we see bigger changes only happen after the inclusion of ISS.
+
+However the same can not be said about the SBP group.
+When adding ISS, the significance of class 4 disappears, but the significance of classes 2 and 3 does not invert.
+The same can be said about the OR, it does change but not nearly as major and clear as in the BE class.
+Especially for class 2, as the OR merely changes from 1,06 to 1,11 and then to 0,96.
+Class 3 is more, from 0,92 to 0,90 to 0,47; but the OR difference is still < 0,5, whereof for the BE group and SBP class 4 the difference is always> 0,5.
+The reason may be due to the cutoffs of the classification system, as the SBP classification does not follow the same range of shock as the BE classification.
+The result of the BE classification originating study [@mutschler_renaissance_2013], shows that BE class I had a mean SBP of 132,6 mmHg and class IV had 94,8 mmHg.
+As an SBP of 132,6 is at the higher range of normotensive, we were therefore not able to use it as an upper limit, but the 94,8 mmHg is close to our approximation of 90mmhg for class IV.
+The SBP classification is therefore only somewhat connected to the BE classification at the lower limit eg class 4.
+Also worth noting is that the divider between our SBP class 2 and 3, are arbitral and made to connect with the ATLS system to make up a four-class system, as the 90mmgh and 110mmhg are from Eastridge et al [@eastridge_hypotension_2007] and Oyetunji et al [@oyetunji_redefining_2011] which only defined 110mmhg and 90mmhg and nothing in between.
+This means that this divider may have made the classes too small and no longer scientific.
+
+#### Insignificant results
+One surprising result is that there is no significance for the SBP group and classes 2 and 3 of the BE group.
+When comparing to Hussein et al.
+[] results, the SBP results conform, as it also showed significance for the unadjusted results below 90mmHg and no significance when adjusted with other factors including ISS.
+
+Due to a lack of research regarding shock/injury severity and OFI/Preventable death as a whole, it is difficult to determine whether the results are reasonable or not.
+It is therefore also difficult to determine exactly why some of them are significant and others not, leaving us with reasoning and hypothetical explanations.
+
+### Comparison to literature
+Previous studies [@teixeira_preventable_2007;@saltzherr_preventability_2011] which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment, while the second most common is clinical judgment error [@teixeira_preventable_2007], or errors in management[@saltzherr_preventability_2011].
+In comparison, our results showed the most common OFI is clinical judgment error (36%), while inadequate resources were second (22%), and delay in treatment comes third (17%), for the whole study population.
+These differences may be due to the nature of the studied population, where delay in treatment may be more deadly, therefore it takes up a smaller part of our results, while clinical judgment error takes up a larger part.
+
+For patients defined as in shock (Class 2-4), the most common was still clinical judgment error (BE 34,7%; SBP 35,3%).
+This result is similar when compared to D O'Reilly's study [@oreilly_opportunities_2013], in which the most common type of opportunities for performance improvements (OPI) was one involved with decision-making at 63 out of 150 OPIs.
+
+We did not find any publication addressing shock severity and OFI.
+We have therefore looked at studies addressing preventable death and assessed severity by other factors.
+In one study [@davoodabadi_predicting_2021], the odds for the most severe class of the external bleeding severity category can be calculated to 2 and when comparing with the “none” class the odds ratio is then 1,91.
+This unadjusted result can be compared to our 2,28 when comparing class 4 to class 1 of the SBP classification (class 4 of BE was not significant).
+While the results are not very close, however, both show a positive association between trauma severity and preventability of negative outcomes.
+The difference may be explained by different classification/grouping requirements.
+In the same study [@davoodabadi_predicting_2021], the odds ratio between patients with SBP ≥ 80 and < 80 for preventable death is calculated to be 0,071.
+This differs a lot from our findings in the unadjusted and may be due to different study outcomes OFI vs death and different SBP classification intervals.
+
+## Strengths and limitations
+One of the limitations is the lack of data to design the study around a modern shock/hemorrhage classification which uses a combination of base parameters.
+Due to a lack of data on heart rate, FAST, detailed types of injury, and clinical chemistry, we were not able to use categorizations such as TASH, TBSS, and ABC with better sensitivity and specificity than the ATLS framework [@yucel_trauma_2006;@ogura_modified_2016;@hanlin_prehospital_2024].
+The ATLS shock classification framework would, in turn, would also be better studied and evidence-based, than the propetrially classification we simplified and built our classification from.
+
+The choice to not include the ATLS hemorrhage classification as it is was due to four reasons.
+The first one is ATLS tenth edition lacks numerical values for vital parameters.
+The second one is that base deficit alone is superior to vital parameters as a predictive parameter for mortality [@dunham_comparison_2017].
+The third one is due to the nature of having multiple parameters, which makes it difficult to categorize a patient with vital parameters in two or more classes at the same time.
+The fourth one is difficulties in separating lower GCS due to traumatic brain injury from shock as a cause.
+
+Missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias, as the most common missing category was BE which may mean that BE as … (källa)
+
+## Clinical/Practical applications
+The results showed only significantly reduced odds for OFI for patients with BE < -10.
+This could mean that the initial management of trauma patients with severe shock is fairly robust.
+Leveraging these insights, protocols for such cases should be studied.
+By simulating these protocols in cases with less severe shock, trauma systems can expand best practices to improve care for patients with less severe shock, minimizing OFI.
 
 ## Future studies
-This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
+The natural next step for future studies is to validate the findings and improve the methods from this study, by using a validated classification for hemorrhagic shock, like ABC, TASH, and so on.
+There are two reasons, one is the findings are more standardized and therefore easy to validate and compare to other studies.
+The second reason is that the results are more clinically usable, as it should be more related to reality and therefore practical when making decisions.
 
-The most common type of OFI is clinical judgement error, it is the most common in all BE shock classes and 2/4 classes for SBP classes.
-is around ...% of.
-It generally takes up ...% of all OFI defined as shock (class 2 and below).
-
-When OFI and shock severity are analysed alone, there are statistical significant correlation for these classes: ...
-.
-It shows positive correlation where the more severe the shock is, the higher the odds are for an OFI to occur.
-This in constrast with the adjusted results which showed inverted correlation betwen OFI and shock severity, with 63% lower odds for OFI when comparing the BE class 4 with class 1.
-However there is no correlation for the other classes nor the classes for the SBP group.
-
-When analysing the other adjusted factors, ISS seems to be a major confounder which correlates to both OFI and shock severity.
-This can explain why the results of the adjusted and unadjusted analyses are inverted.
-
-Due to lack of data on heart rate, FAST, detailed types of injury and clinical chemistry, we will not be able to use categorisations such as TASH, TBSS and ABC with better sensitivity and specificity than ATLS[@yucel_trauma_2006][@ogura_modified_2016][@hanlin_prehospital_2024].
-Results from the BE approximation done by Mutschler et al [mutschler_renaissance_2013] which is the same as the tenth edition ATLS shock classification, shows that class I had a mean SBP of 132,6 mmhg and class IV have 94,8 mmhg at the emergency department.
-As a SBP of 132,6 is at the higher range of normotensive, we will not be able to use it as a upper limit, but the 94,8 mmhg is close to our approximation of 90mmhg for class IV.
-
-We decided to not include Glasgow Coma Scale (GCS) due to difficulties in differentiating if any lowering is due to shock or traumatic brain injury (TBI).
-Urine output and pulse pressure were also neglected, due to lack of data.
-
-There are three reasons.
-The first one is ATLS tenth edition lack numerical values for vital parameters.
-The second one is that bace deficit is superior to vital parameters as a predictive parameter for mortality[@dunham_comparison_2017].
-The third one is due to nature of having multiple parameters, makes it difficult to categorise a patient with vital parameters in two or more classes and difficult to separate lower GCS due to traumatic brain injury from shock as cause.
-
-Hantera missing med listwise deletion.
+As previously mentioned, ISS seems to be correlated to OFI, and we couldn't find any studies on it.
+Therefore it would be useful in future studies validating the association between other injury markers from like ISS and OFI.
+This information can later be used to improve and standardize the audit filters used for QI programs.
 

### 2024-12-10 thesis.Rmd
Första word verion - ska fixa lite i de nya tabellerna

@@
 output:
-  html_document: default
+  #html_document: default
   word_document: default

### 2024-12-16 thesis.Rmd
La till master combined

@@
 ```{r ISS confounding, echo=FALSE}
-sample.characteristics.table_ISS
+sample.characteristics.table_ISS_BE
+sample.characteristics.table_ISS_SBP
+```
+
+DIVIDER
+
+```{r masterBE, echo=FALSE}
+master_combined_table_stepBE
+```
+
+DIVIDER
+
+```{r masterSBP, echo=FALSE}
+master_combined_table_stepSBP
 ```

### 2024-12-16 main.R

@@ -190,7 +190,9 @@ SBPstep1 <- tbl_regression(
   label = list(
     pt_age_yrs ~ "Age (Years)",
     pt_Gender ~ "Gender (M/F)",
-    V4SBP_class ~ "Shock classification - SBP"
+    V4SBP_class ~ "Shock classification - SBP",
+    pt_asa_preinjury ~ "Pre-injury ASA",
+    ed_inr_numeric ~ "INR"
   )
 ) |>
   bold_p()
@@ -206,7 +208,10 @@ SBPstep2 <- tbl_regression(
   label = list(
     pt_age_yrs ~ "Age (Years)",
     pt_Gender ~ "Gender (M/F)",
-    V4SBP_class ~ "Shock classification - SBP"
+    V4SBP_class ~ "Shock classification - SBP",
+    pt_asa_preinjury ~ "Pre-injury ASA",
+    ed_inr_numeric ~ "INR",
+    ISS ~ "Injury Severity Score"
   )
 ) |>
   bold_p()
@@ -223,7 +228,9 @@ BEstep1 <- tbl_regression(
   label = list(
     pt_age_yrs ~ "Age (Years)",
     pt_Gender ~ "Gender (M/F)",
-    BE_class ~ "Shock classification - BE"
+    BE_class ~ "Shock classification - BE",
+    pt_asa_preinjury ~ "Pre-injury ASA",
+    ed_inr_numeric ~ "INR"
   )
 ) |>
   bold_p()
@@ -239,7 +246,10 @@ BEstep2 <- tbl_regression(
   label = list(
     pt_age_yrs ~ "Age (Years)",
     pt_Gender ~ "Gender (M/F)",
-    BE_class ~ "Shock classification - BE"
+    BE_class ~ "Shock classification - BE",
+    pt_asa_preinjury ~ "Pre-injury ASA",
+    ed_inr_numeric ~ "INR",
+    ISS ~ "Injury Severity Score"
   )
 ) |>
   bold_p() 
@@ -252,7 +262,7 @@ var_label(study.sample$BE_class) <- "Shock classification - BE"
 var_label(study.sample$pt_asa_preinjury) <- "Pre-injury ASA"
 var_label(study.sample$ISS) <- "Injury Severity Score"
 var_label(study.sample$pt_Gender) <- "Gender (M/F)"
-var_label(study.sample$ed_sbp_value) <- "Systolic blood pressure (mmhg)"
[…]

### 2025-02-16 manuscript.Rmd
La in de nya delarna, direkt från exjaobb

@@
 ```
+
+Background: Hemorrhagic shock is the leading cause of preventable trauma deaths.
+Opportunities for improvement (OFI) are common in deaths from massive bleeding.
+Aims: This study aimed to describe the types of OFI for adult trauma patients arriving in shock and assess the association between shock severity and OFI.
+Methods: A retrospective cohort study was conducted on adult trauma patients treated at Karolinska University Hospital (2014-2023).
+Shock severity was categorized into four classes using base excess (BE) and systolic blood pressure (SBP).
+OFI identified during morbidity and mortality conference were analyzed using bi- and multivariable logistic regression, with adjustments for age, sex, American Society of Anesthesiologists Classification (ASA), international normalized ratio (INR), and injury severity score (ISS) Results: 4919 patients were included, of which 281 had OFI.
+Clinical judgment error was the most common OFI type for patients in shock.
+Unadjusted analyses showed positive associations between shock severities and OFI.
+However, after adjusting for ISS, patients with severe shock (BE < -10) showed 63% reduced odds of OFI compared to those without shock (p = 0.026).
+ISS was a significant confounder (p < 0.001).
+No significant association was observed between SBP classes and OFI in adjusted analysis.
+Conclusion: Association between shock and OFI seems to be nonlinear, with moderate shock (BE (-2) – (-10)) having highest odds of OFI.
+The lack of significance in SBP classification highlights potential framework limitations compared to BE.
+These findings emphasize the importance of accounting for ISS and using validated shock classifications in future research.
+
 ## Background
@@
 The study demographics median age is `r age` and the most common gender is male, `r male`.
-Clinically, the median ISS is 5 which is considered minor injuries, median INR is 1,00, and the most common Preinjury ASA class is 1, at 62%.
+Clinically, the median ISS is 5 which is considered minor injuries, median INR is 1.00, and the most common Preinjury ASA class is 1, at 62%.
 
@@
 When classified, the most common broad sub-ofi in class 1 (no shock) is clinical judgement error both in the BE and SBP groups, while the other sub-ofi also follow a similar distribution between the groups.
-The largest sub-ofi continuous to be clinical judgment error for BE group in class 2, 3 and 4, at 34,6%; 31,2%; and 42,9% in this order.
+The largest sub-ofi continuous to be clinical judgment error for BE group in class 2, 3 and 4, at 34.6%; 31.2%; and 42,9% in this order.
 On other hand, the largest sub-ofi in SBP class 2 is delay in treatment at 50%.
 The SBP class 3 is evenly distributed between clinical judgement error and missed diagnosis at 40% each.
-Lastly, class 4 SBP's most common sub-ofi is clinical judgment error 46,7%.
+Lastly, class 4 SBP's most common sub-ofi is clinical judgment error 46.7%.
 
@@
 # Discussion
+The purpose of this study was to assess the association between OFI and adult trauma patients arriving in shock, focusing on the types of OFI and severity of shock.
+
+##Key results
+We found that the most common type of OFI is clinical judgment error across almost all shock severities.
+Out of all shock severities, there is only a statistically significant association between severe shock with a BE of < -10 and OFI, which shows 63% reduced odds of OFI when compared against those with no shock.
+The unadjusted analyses demonstrated a general positive association between shock and OFI. However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI. 
+##Interpretation
+Patients with BE < -10, indicating severe shock, were found to have lower odds of OFI compared to other groups, and two hypotheses may explain this finding.
+First, patients in severe shock should be more easily recognized as critically ill upon arrival, triggering heightened clinical attention and prioritization in triage.
+This results in more structured management and often receives priority in resource allocation, which may minimize variability in care and reduce the likelihood of errors.
+Another explanation is the potential influence of “reversed” survival bias.
+Patients with BE < -10 have high mortality rates (25), and for these patients who die, the care process is interrupted prematurely.
+This may induce generally fewer procedures or shorter care time for patients with deadly outcomes, which means less chance for errors, and therefore OFIs.
+This would result in reduced amount of OFIs in groups with higher mortality, which in this case would be class 4.
+Interpretation of adjusted model after inclusion of ISS
+When looking at the distribution of the significant and insignificant results of the BE group (table 6), it can be observed that the significance reverses for the classes when adding ISS.
+One explanation is that the unadjusted positive correlation may be driven by a shared variability with ISS, as we can see significant positive association between ISS and OFI in our results.
+This positive association between OFI and ISS have also been shown in a study by Husssein et al (26).
+It is important to note that Hussein’s study was also conducted at Karolinska Solna, but during a different time frame (2017-2021) and with other inclusion and exclusion criteria.
+Another hypothetical explanation is multicollinearity, as both BE and ISS assess injury severity.
+Theoretically anatomical injury, bleeding and physiological response should be closely interconnected.
+A correlation between ISS and BD (inversed BE) or SBP was also shown in the two studies by Mutscher et al.
+(11,25).
+Therefore, adjusting for each other may explain the inversion of OR.
+However, it may also result in results that’s difficult to interpret or lack any meaning in a clinical context.
+This may also explain why class 2 and 3 are not significant but class 4 are in the adjusted analysis, as seen in study by Hussein et al (26), ISS does not correlate linearly to OFI, but we included ISS linearly in the model.
+However, when adding ISS to the SBP group, the significance of class 4 disappears, but the significance of classes 2 and 3 does not change, like in the case for BE group.
+The same can be said about the OR, it does change but not nearly as major as in the BE class.
+Especially for class 2, as the OR merely changes from 1.06 to 1.11 and then to 0.96.
+Class 3 OR changes more, from 0.92 to 0.90 to 0.47; but the OR difference is still < 0.5, whereof for the BE group and SBP class 4 the difference is always > 0.5.
+The reason may be due to the cutoffs of the classification system, as the SBP classification does not follow the same range of shock as the BE classification.
+The results of the study that BE classification was originating from (25), shows that BE class 1 had a mean SBP of 132.6 mmHg and class IV had 94.8 mmHg.
+As an SBP of 132.6 mmHg is at the higher range of normotensive, we were therefore not able to use it as an upper limit, but the 94.8 mmHg is close to our approximation of 90 mmHg for class 4.
+The SBP classification is therefore only somewhat connected to the BE classification at the lower limit e.g.
+class 4.
+Also worth noting is that the divider between our SBP class 2 and 3, are arbitrary and made to connect with the ATLS system to make up a four-class system, as the 90 mmHg and 110 mmHg are from Eastridge et al (23) and Oyetunji et al (24) which only defined 110 mmHg and 90 mmHg and nothing in between.
+This means that this divider may have made the classes too small and no longer scientific.
+###Comparison to literature
+Previous studies (7,27) which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment, while the second most common is clinical judgment error (7), or errors in management (27).
+In comparison, our results showed the most common OFI is clinical judgment error (36%), while inadequate resources were second (22%), and delay in treatment comes third (17%), for the whole study population.
+These differences may be due to the nature of the studied population, where delay in treatment may be more deadly, therefore it takes up a smaller part of our results, while clinical judgment error takes up a larger part.
+No studies were found to address this hypothesis of survivorship bias for OFI types.
+For patients defined as in shock (Class 2-4), the most common OFI was still clinical judgment error (BE 34.7%; SBP 35.3%).
+This result is similar when compared to D O’Reilly’s study (28), in which the most common type of opportunities for performance improvements (OPI) was one involved with decision-making at 63 out of 150 OPIs.
+We did not find any publication addressing shock severity and OFI.
+We have therefore looked at studies addressing preventable negative care outcomes and assessed severity by other factors but only found one that investigated predicting factors and preventable death.
+This study (29) showed that there was a OR correlation of 14.1 between preventable death and SBP ≥ 80mmHg, when comparing to SBP < 80 mmHg.
+This differs from our SBP findings, as our adjusted results are not statistically significant, and the significant unadjusted results don’t have that extreme of OR and are positively correlated with severe and not mild shock.
+These differences are not strange, but rather understandable, as the studies are conducted at different hospitals, populations, outcomes (surviving vs death) and different classifications (the lowest of our 4 SBP classes is 90mmHg, while they only got 2 classes that separate at 80mmHg).
+However, if reasoned from a shock standpoint where both SBP and BE are part of, and compare this result with our BE results, they show a similar trend.
+Our BE results showed reduced odds of OFI when comparing the most severe and the no shock class.
+This implies that if changing the reference point to the most severe shock class, the result would be increased odds of OFI when compared to the no shock class.
+Therefore, our results show similar trend to their study (29) result, that showed lesser shock (higher SBP) had increased odds ratio of preventable death.
+One surprising result is that there is no significance for the SBP group in our adjusted analysis.
+When comparing to the results from Hussein et al.
+(26), the SBP results conform, as it also showed significance for the unadjusted results below 90 mmHg and no significance when adjusted with other factors including ISS.
+Due to a lack of research regarding shock/injury severity and OFI/preventable death, it is difficult to determine whether the results are reasonable or not.
+It is therefore also difficult to determine exactly why some of them are significant and others not, leaving us with reasoning and hypothetical explanations.
+##Strengths and limitations
+Strength of this study includes the relatively large cohort of 4919 patients spread over 9 continuous years.
+From a methodical standpoint, even though an established shock/hemorrhage classification system was not used, two clinical parameters were used in parallel.
+This includes two perspectives, one physiological and one chemical, to best describe shock and at the same time still using the original parameters the clinicians meet on a day-to-day basis.
+One of the limitations is the lack of data to design the study around a modern shock/hemorrhage classification which uses a combination of base parameters.
+Due to a lack of data on heart rate, FAST, detailed types of injury, and clinical chemistry, we were not able to use categorizations such as TASH, TBSS, and ABC with better sensitivity and specificity than the ATLS framework (30–32).
+The ATLS shock classification framework would, in turn, would also be better studied and evidence-based, than the proprietary classification we simplified and built our classification from.
+The choice to not include the ATLS hemorrhage classification was due to four reasons.
+The first one is ATLS tenth edition lacks numerical values for vital parameters.
+The second one is that base deficit alone is superior to vital parameters as a predictive parameter for mortality (33).
+The third one is due to the nature of having multiple parameters, which makes it difficult to categorize a patient with vital parameters in two or more classes at the same time.
+The fourth one is difficulties in separating lower GCS due to traumatic brain injury from shock as a cause.
+Missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias, as the most common missing category was BE which may mean that BE was not routinely measured and result in inconsistency for certain patient groups, which risks selection bias.
+Also worth noting is that the sample size for most of the shock classes (class 2 to 4) are roughly around 100 to 200 patients, whereof those with positive OFI are as low as 5 patients.
+This may result in over/underestimation of the OR, as there is higher chance of randomness in smaller sample sizes, which results in broader confidence intervals which may result in more “extreme” average OR.
+##Clinical/Practical applications - Generalisability?
+Clinically the unadjusted are more insightful, as the variables are never separated and instead combined into complex cases.
+Therefore, from a BE standpoint when a patient’s BE is between -2 to -10 the odds of OFI are between 61% to 91% higher than those with BE above -2.
+Suggesting patients with a BE between these levels may need extra attention.
+From a SBP perspective, when SBP dips below 90 mmHg odds for OFI increases a massive 128%, compared with those with SBP above 110 mmHg.
+This supports the idea (24) of using 90 mmHg as a clinical limit for defining hypotension.
+###Equity
+It can be observed that the male gender is more prevalent in our results than females (71% vs 29%), which is like other studies (22,27,28).
+However, when comparing the event rate of OFI among the patients reviewed for OFI, the difference between the genders becomes minimal at 5.1% for female, respective 6% for male.
+At the same time, when fully adjusted, gender is not significant (p = 0.4), this implies that even though gender may not be proportionate in trauma cases, there seems to be no major difference between the genders for trauma care quality.
+This may reflect that the standardized protocols for trauma care minimize gender-based differences.
+When comparing to other studies (27,28), it seems like the percentage potentially preventable death and OPI have a higher amount for males, however, to be noted their percentage is based on ratio between genders and not event risk for each gender.
+Socioeconomical factors were not incorporated into this study, nonetheless an important factor in emergency medicine, when resources are scarcer.
+Especially when “inadequate resources” is the third most common OFI type, even for a major hospital in the capital of Sweden.
+This implies the results of this study cannot be generalized globally, especially not places where standardized trauma care protocols are not incorporated and/or lack adequate resources.
+###Future studies
+The next step for future studies is to validate the findings and improve the methods from this study.
+As previously discussed, the association between anatomical classification ISS and physiological parameters BE and SBP, and how they correlate to OFI are complex and not necessarily clinically relevant.
+One way to avoid this problem is using a validated classification for hemorrhagic shock with both anatomical and physiological components, like ABC, TASH, and so on.
+One other problem when interpreting the results is the “reversed survivorship bias”.
+Due to the design of trauma care quality database, when and where (level of care) the OFI occurred during the length of stay is not recorded.
+A prospective study can clarify whether shock severity directly affects OFI or if length of stay is a confounding factor.
+By observing when and where OFIs most commonly occur, it can determine whether the hypothesis of length of stay is correct.
+This also aids in understanding the reasons behind the nonlinear relationship between ISS, BE or SBP and OFI, as there are currently only theories but no studies.
 
@@
 
+The most common type of OFI for shock patients is clinical judgement error.
+Association between shock and OFI doesn’t seem to be linear, but rather the moderate shock has higher odds of OFI compared to the mild and severe shocks.
+The lack of significance in SBP classification highlights potential framework limitations compared to BE.
+These findings emphasize the importance of accounting for ISS and using validated shock classifications in future research.
+
+#Contributions
+Data collection from the trauma registry, ethics application and study designed by Martin Gerdin Wärnberg.
+Carl sorted, programmed and analyzed the data, and then wrote the thesis with support from Martin Gerdin Wärnberg, Johanna Berg, Jonatan Attergrim and Kelvin Szolnoky.
+Technical tools used include Chat-GPT and Google translate.
+Both were used in the same way, as means to translate more adequate vocabulary for scientific writing between Swedish and English.
+The overall formulations, structure and content were written by me (Carl), and which of the translations done by the two tools, that were to be used to ensure the content, were also done by me (Carl).
+
+##Acknowledgements
+I want to here again thank my supervisor and co-supervisors for supporting, replying to questions and reading and giving me constructive feedback.
+I also want to thank my coordinator Jan Jakobsson for a lot of constructive feedback and guiding my understanding of the practicalities and extent for this thesis.
+My fellow medical student Anton Fedchenko also supported me with constructive feedback.
+Lastly, I also want to thank my friend Gustav Nilsson-Gisleskog, a master student in computer science at Lund University for helping me solve conflicting push pull requests on git-hub.
+
 # References

### 2025-02-16 manuscript.Rmd
la till avstånd och testa comitta html

@@
   # word_document: default
-  bookdown::pdf_document2:
-    latex_engine: xelatex
-  # html_document: default
+  #bookdown::pdf_document2:
+    #latex_engine: xelatex
+   html_document: default
 bibliography: bibliography.bib
@@
 The unadjusted analyses demonstrated a general positive association between shock and OFI. However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI.
+
 ##Interpretation
@@
 This means that this divider may have made the classes too small and no longer scientific.
+
 ###Comparison to literature
@@
 It is therefore also difficult to determine exactly why some of them are significant and others not, leaving us with reasoning and hypothetical explanations.
+
 ##Strengths and limitations
@@
 This may result in over/underestimation of the OR, as there is higher chance of randomness in smaller sample sizes, which results in broader confidence intervals which may result in more “extreme” average OR.
+
 ##Clinical/Practical applications - Generalisability?
@@
 This supports the idea (24) of using 90 mmHg as a clinical limit for defining hypotension.
+
 ###Equity
@@
 This implies the results of this study cannot be generalized globally, especially not places where standardized trauma care protocols are not incorporated and/or lack adequate resources.
+
 ###Future studies

### 2025-02-16 manuscript.Rmd
ändring av result, kommit fram till outcome

@@
   # word_document: default
-  #bookdown::pdf_document2:
-    #latex_engine: xelatex
-   html_document: default
+  bookdown::pdf_document2:
+    latex_engine: xelatex
+  #html_document: default
 bibliography: bibliography.bib
@@
 ## Participants
-There were a total of `r merged` patients in the trauma registry.
-After excluding the patients younger than 15 and/or were dead on arrival, there were `r excludednum` patients left.
-Out of those, `r sample` patients had been reviewed for the presence of OFI.
-A total of `r sample - log_reg_count` patients were excluded due to missing data, resulting in `r log_reg_count` patients for the final analysis.
+There was a total of 14022 patients in the trauma registry.
+After excluding the patients younger than 15 and/or were dead on arrival, there were 12153 patients remaining.
+Out of those, 7152 patients had been reviewed for the presence of OFI.
+A total of 2233 patients were excluded due to missing data, resulting in 4919 patients for the final analysis.
+The selection process can also be seen in Figure 1.
+The variable with the most missing data is BE which lacks the data for 1721 patients.
+Detailed data regarding missing data is displayed in a table in appendix 1.
 
-The variable with most missing data is BE which lack the data for 1721 patients.
-
+#ändra till flowchart
 Table 3 - Sample characteristics showing missing data.
@@
 ```
+#slut
 
@@
 Table \@ref(tab:sample-characteristics-reg) describes the characteristics of the study population.
-The study demographics median age is `r age` and the most common gender is male, `r male`.
-Clinically, the median ISS is 5 which is considered minor injuries, median INR is 1.00, and the most common Preinjury ASA class is 1, at 62%.
+The study demographics median age was `r age` and the most common gender was male, `r male`.
+Clinically, the median ISS was 5 which is considered minor injuries, median INR was 1.00, and the most common Preinjury ASA class was 1, at 62%.
 
@@
 ## Outcome data
-Out of the `r log_reg_count` patients reviewed without missing data, `r ofi` patients had at least one positive OFI.
+Out of the 4919 patients reviewed without missing data, 281 (5.7%) patients had at least one positive OFI.
 
-The most common broad sub-ofi category was clinical judgement error, at 103 (37%).
-The other three major broad sub-ofis are: inadequate resources 61 (22%), delay in treatment 49 (17%) and missed diagnosis 48 (17%).
-The smallest one, is other errors, at 5 (1,8%).
+The most common OFI type was clinical judgement error, numbering 103 (37%) patients.
+The other three major OFI types were: inadequate resources for 61 (22%) patients, delay in treatment for 49 (17%) patients and missed diagnosis for 48 (17%) patients.
+The least common OFI was “other errors”, for 5 (1.8%) patients.
 
-When classified, the most common broad sub-ofi in class 1 (no shock) is clinical judgement error both in the BE and SBP groups, while the other sub-ofi also follow a similar distribution between the groups.
-The largest sub-ofi continuous to be clinical judgment error for BE group in class 2, 3 and 4, at 34.6%; 31.2%; and 42,9% in this order.
-On other hand, the largest sub-ofi in SBP class 2 is delay in treatment at 50%.
-The SBP class 3 is evenly distributed between clinical judgement error and missed diagnosis at 40% each.
-Lastly, class 4 SBP's most common sub-ofi is clinical judgment error 46.7%.
+When classified, the most common OFI type in class 1 (no shock) was clinical judgement error both in the BE and SBP groups, while the other OFI types also followed a similar distribution between the groups.
+The largest OFI type continues to be clinical judgment error for BE group in class 2, 3 and 4, at 34.6%; 31.2%; and 42.9% in this order.
+On other hand, the largest OFI type in SBP class 2 was delay in treatment at 50%.
+The SBP class 3 was evenly distributed between clinical judgement error and missed diagnosis at 40% each.
+Lastly, class 4 SBP’s most common OFI type was clinical judgment error 46.7%.
 
-Overall 3989 (81%) of the patients were classified as having no shock (class 1) according to the BE classification, the sum of the remaining classes accounted for 930 (19%) patients.
-The distribution of the classes were decreasing in numbers with increased severity, with class 2: 644 (13%), class 3: 170 (3,5%), and class 4: 116 (2,4%).
+Overall, 3989 (81%) of the patients were classified as having no shock (class 1) according to the BE classification, the sum of the remaining classes accounted for 930 (19%) patients.
+The distribution of the classes was decreasing in numbers with increased severity, with class 2: 644 (13%), class 3: 170 (3.5%), and class 4: 116 (2.4%).
 
-The SBP classification system sorted 4455 (91%) patients as having no shock (class 1), while the remaining classes with patients in shock accounted for 464 (9%) of all the patients.
-Among the classes defined with shock, class 2 was the biggest at 239 (4,9%), followed by class 4 at 127 (2,6%), and lastly class 3 with 99 (2,0%).
+According to SBP, 4455 (91%) patients had no shock (class 1), while the remaining classes with patients in shock accounted for 464 (9%) of all the patients.
+Among the classes defined with shock, class 2 was the biggest at 239 (4.9%), followed by class 4 at 127 (2.6%), and lastly class 3 with 99 (2.0%).
+
+-------
 

### 2025-02-16 skräpsnuttar.R

@@ -210,3 +210,7 @@ dimnames(tabell) <- list(OFI = c("Yes", "No"),
                          Shock_class = c("Class 1", "Class 2", "Class 3", "Class 4"))
 tabell
 chisq.test(tabell)
+
+There were a total of  patients in the trauma registry. After excluding the patients younger than 15 and/or were dead on arrival, there were `r excludednum` patients left. Out of those, `r sample` patients had been reviewed for the presence of OFI. A total of `r sample - log_reg_count` patients were excluded due to missing data, resulting in `r log_reg_count` patients for the final analysis.
+
+The variable with most missing data is BE which lack the data for 1721 patients.

### 2025-02-16 manuscript.Rmd
ändrade plats för mallen

@@
 # Discussion
+```{=html}
+<!--
+
+Key results
+-----------
+Summarise key results with reference to study objectives
+
+Limitations
+-----------
+
+Discuss limitations of the study, taking into account sources of
+potential bias or imprecision.  Discuss both direction and magnitude
+of any potential bias
+
+Interpretation
+--------------
+
+Give a cautious overall interpretation of results considering
+objectives, limitations, multiplicity of analyses, results from
+similar studies, and other relevant evidence
+
+Generalisability
+----------------
+
+Discuss the generalisability (external validity) of the study results
+
+-->
+```
 The purpose of this study was to assess the association between OFI and adult trauma patients arriving in shock, focusing on the types of OFI and severity of shock.
@@
 
-```{=html}
-<!--
-
-Key results
------------
-Summarise key results with reference to study objectives
-
-Limitations
------------
-
-Discuss limitations of the study, taking into account sources of
-potential bias or imprecision.  Discuss both direction and magnitude
-of any potential bias
-
-Interpretation
---------------
-
-Give a cautious overall interpretation of results considering
-objectives, limitations, multiplicity of analyses, results from
-similar studies, and other relevant evidence
-
-Generalisability
-----------------
-
-Discuss the generalisability (external validity) of the study results
-
--->
-```
 # Conclusion
-
 <!-- Give a short conclusion, corresponding to the aim -->

### 2025-02-16 manuscript.Rmd
la till fler rader

@@
 Out of all shock severities, there is only a statistically significant association between severe shock with a BE of < -10 and OFI, which shows 63% reduced odds of OFI when compared against those with no shock.
+
 The unadjusted analyses demonstrated a general positive association between shock and OFI. However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI.
@@
 Patients with BE < -10, indicating severe shock, were found to have lower odds of OFI compared to other groups, and two hypotheses may explain this finding.
+
 First, patients in severe shock should be more easily recognized as critically ill upon arrival, triggering heightened clinical attention and prioritization in triage.
 This results in more structured management and often receives priority in resource allocation, which may minimize variability in care and reduce the likelihood of errors.
+
 Another explanation is the potential influence of “reversed” survival bias.
@@
 This would result in reduced amount of OFIs in groups with higher mortality, which in this case would be class 4.
-Interpretation of adjusted model after inclusion of ISS
+
+###Interpretation of adjusted model after inclusion of ISS
 When looking at the distribution of the significant and insignificant results of the BE group (table 6), it can be observed that the significance reverses for the classes when adding ISS.
@@
 It is important to note that Hussein’s study was also conducted at Karolinska Solna, but during a different time frame (2017-2021) and with other inclusion and exclusion criteria.
+
 Another hypothetical explanation is multicollinearity, as both BE and ISS assess injury severity.
@@
 This may also explain why class 2 and 3 are not significant but class 4 are in the adjusted analysis, as seen in study by Hussein et al (26), ISS does not correlate linearly to OFI, but we included ISS linearly in the model.
+
 However, when adding ISS to the SBP group, the significance of class 4 disappears, but the significance of classes 2 and 3 does not change, like in the case for BE group.
@@
 No studies were found to address this hypothesis of survivorship bias for OFI types.
+
 For patients defined as in shock (Class 2-4), the most common OFI was still clinical judgment error (BE 34.7%; SBP 35.3%).
@@
 Therefore, our results show similar trend to their study (29) result, that showed lesser shock (higher SBP) had increased odds ratio of preventable death.
+
 One surprising result is that there is no significance for the SBP group in our adjusted analysis.
@@
 This includes two perspectives, one physiological and one chemical, to best describe shock and at the same time still using the original parameters the clinicians meet on a day-to-day basis.
+
 One of the limitations is the lack of data to design the study around a modern shock/hemorrhage classification which uses a combination of base parameters.
@@
 The ATLS shock classification framework would, in turn, would also be better studied and evidence-based, than the proprietary classification we simplified and built our classification from.
+
 The choice to not include the ATLS hemorrhage classification was due to four reasons.
@@
 The fourth one is difficulties in separating lower GCS due to traumatic brain injury from shock as a cause.
+
 Missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias, as the most common missing category was BE which may mean that BE was not routinely measured and result in inconsistency for certain patient groups, which risks selection bias.
@@
 One way to avoid this problem is using a validated classification for hemorrhagic shock with both anatomical and physiological components, like ABC, TASH, and so on.
+
 One other problem when interpreting the results is the “reversed survivorship bias”.

### 2025-03-15 manuscript.Rmd
Ändrat till word

@@
 output:
-  # word_document: default
-  bookdown::pdf_document2:
-    latex_engine: xelatex
+  word_document: default
+  #bookdown::pdf_document2:
+    #latex_engine: xelatex
   #html_document: default

### 2025-04-18 manuscript.Rmd
fixat lite diagram versioner

@@
 #ändra till flowchart
-Table 3 - Sample characteristics showing missing data.
-```{r sample-characteristics, echo=FALSE}
-sample.characteristics.table
-```
+
 #slut
@@
 Clinically, the median ISS was 5 which is considered minor injuries, median INR was 1.00, and the most common Preinjury ASA class was 1, at 62%.
+
+Table 3
 
@@
 
-Table 5 - Sub-OFI, classified according to BE
+Table 4 - Sub-OFI, classified according to BE
 
@@
 
-Table 6 - Sub-OFI, classified according to SBP
+Table 5 - Sub-OFI, classified according to SBP
 
@@
 
-Table 7 and 8 adjusted log
-```{r Adjusted_log, echo=FALSE}
-combined_tableBE
-combined_tableSBP
+Table 6 adjusted log BE
+```{r masterBE, echo=FALSE}
+master_combined_table_stepBE
+```
+
+Table 7 adjusted log SBP
+```{r masterSBP, echo=FALSE}
+master_combined_table_stepSBP
 ```

### 2025-04-18 main.R

@@ -457,3 +457,4 @@ master_combined_table_stepSBP <- tbl_merge(
   tbls = list(log_regSBP_sample.characteristics.table, SBPstep1, log_regSBP_sample.characteristics.table_unadjusted),
   tab_spanner = c("**Fully adjusted**","**without ISS**","**Unadjusted**")
 ) 
+

### 2025-05-30 manuscript.Rmd
testar knitta från ny dator

@@
 output:
-  word_document: default
+  #word_document: default
   #bookdown::pdf_document2:
     #latex_engine: xelatex
-  #html_document: default
+  html_document: default
 bibliography: bibliography.bib

### 2025-07-13 manuscript.Rmd
la in förändringarna enligt google docs och källorna i discussion

@@
 csl: vancouver.csl
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 ```
+
 ```{=html}
@@
 ```
+
 # Abstract
@@
 
-Background: Hemorrhagic shock is the leading cause of preventable trauma deaths.
-Opportunities for improvement (OFI) are common in deaths from massive bleeding.
-Aims: This study aimed to describe the types of OFI for adult trauma patients arriving in shock and assess the association between shock severity and OFI.
-Methods: A retrospective cohort study was conducted on adult trauma patients treated at Karolinska University Hospital (2014-2023).
-Shock severity was categorized into four classes using base excess (BE) and systolic blood pressure (SBP).
-OFI identified during morbidity and mortality conference were analyzed using bi- and multivariable logistic regression, with adjustments for age, sex, American Society of Anesthesiologists Classification (ASA), international normalized ratio (INR), and injury severity score (ISS) Results: 4919 patients were included, of which 281 had OFI.
-Clinical judgment error was the most common OFI type for patients in shock.
-Unadjusted analyses showed positive associations between shock severities and OFI.
-However, after adjusting for ISS, patients with severe shock (BE < -10) showed 63% reduced odds of OFI compared to those without shock (p = 0.026).
-ISS was a significant confounder (p < 0.001).
-No significant association was observed between SBP classes and OFI in adjusted analysis.
-Conclusion: Association between shock and OFI seems to be nonlinear, with moderate shock (BE (-2) – (-10)) having highest odds of OFI.
-The lack of significance in SBP classification highlights potential framework limitations compared to BE.
-These findings emphasize the importance of accounting for ISS and using validated shock classifications in future research.
+Background: Hemorrhagic shock is the leading cause of preventable trauma
+deaths. Opportunities for improvement (OFI) are common in deaths from
+massive bleeding. Aims: This study aimed to describe the types of OFI
+for adult trauma patients arriving in shock and assess the association
+between shock severity and OFI. Methods: A retrospective cohort study
+was conducted on adult trauma patients treated at Karolinska University
+Hospital (2014-2023). Shock severity was categorized into four classes
+using base excess (BE) and systolic blood pressure (SBP). OFI identified
+during morbidity and mortality conference were analyzed using bi- and
+multivariable logistic regression, with adjustments for age, sex,
+American Society of Anesthesiologists Classification (ASA),
+international normalized ratio (INR), and injury severity score (ISS)
+Results: 4919 patients were included, of which 281 had OFI. Clinical
+judgment error was the most common OFI type for patients in shock.
+Unadjusted analyses showed positive associations between shock
+severities and OFI. However, after adjusting for ISS, patients with
+severe shock (BE \< -10) showed 63% reduced odds of OFI compared to
+those without shock (p = 0.026). ISS was a significant confounder (p \<
+0.001). No significant association was observed between SBP classes and
+OFI in adjusted analysis. Conclusion: Association between shock and OFI
+seems to be nonlinear, with moderate shock (BE (-2) – (-10)) having
+highest odds of OFI. The lack of significance in SBP classification
+highlights potential framework limitations compared to BE. These
+findings emphasize the importance of accounting for ISS and using
+validated shock classifications in future research.
 
@@
 ```
+
 # Introduction
-Trauma is a major global public health concern.
-It causes over four million deaths, among those younger populations are the most affected[@noauthor_injuries_nodate].
-Beyond the high mortality rate, trauma also imposes a substantial socio-economic burden, due to long-term disabilities and rehabilitation needs[@noauthor_injuries_nodate].
 
-The initial management of trauma patients is highly time-sensitive and prone to error [@mackersie_pitfalls_2010;@teixeira_preventable_2007;@ivatury_patient_2008].
-Quality Improvement (QI) initiatives address these challenges by systematically evaluating care processes and outcomes, aiming to reduce morbidity and mortality [@world_health_organization_guidelines_2009].
+Trauma is a major global public health concern. It causes over four
+million deaths, among those younger populations are the most
+affected[@noauthor_injuries_nodate]. Beyond the high mortality rate,
+trauma also imposes a substantial socio-economic burden, due to
+long-term disabilities and rehabilitation
+needs[@noauthor_injuries_nodate].
 
-Hemorrhagic shock is, the second most common cause of death following trauma and it, is the leading preventable cause of death within the first 24 hours post-injury [@kauvar_impact_2006;@berry_shock_2015;@teixeira_preventable_2007], and opportunities for improvement are common in patients who die because of massive traumatic bleeding [@oreilly_opportunities_2013].
+The initial management of trauma patients is highly time-sensitive and
+prone to error [@mackersie_pitfalls_2010; @teixeira_preventable_2007;
+@ivatury_patient_2008]. Quality Improvement (QI) initiatives address
+these challenges by systematically evaluating care processes and
+outcomes, aiming to reduce morbidity and mortality
+[@world_health_organization_guidelines_2009].
 
-This may indicate that these preventable events potentially leading to suboptimal outcomes are common also in patients who survive massive bleeding.
-This study aims to describe the types of opportunities for improvement for adult trauma patients arriving in shock, and to assess how the degree of shock is associated with opportunities for improvement.
-This knowledge would help clinicians avoid such events in bleeding trauma patients.
+Hemorrhagic shock is, the second most common cause of death following
+trauma and it, is the leading preventable cause of death within the
+first 24 hours post-injury [@kauvar_impact_2006; @berry_shock_2015;
+@teixeira_preventable_2007], and opportunities for improvement are
+common in patients who die because of massive traumatic bleeding
+[@oreilly_opportunities_2013].
+
+This may indicate that these preventable events potentially leading to
+suboptimal outcomes are common also in patients who survive massive
+bleeding. This study aims to describe the types of opportunities for
+improvement for adult trauma patients arriving in shock, and to assess
+how the degree of shock is associated with opportunities for
+improvement. This knowledge would help clinicians avoid such events in
+bleeding trauma patients.
 
@@
 
-We conducted a registry based retrospective cohort study, using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna.
+We conducted a registry based retrospective cohort study, using data
+from the trauma registry and trauma care quality database at the
+Karolinska University Hospital in Solna.
 
@@
 
-The trauma registry includes patients treated at Karolinska University Hospital in Solna, which treats all major trauma in the greater metropolitan area of Stockholm.
-We included patients registered between 2014 and 2023.
-The trauma care quality database is a subset of the trauma registry and includes patients selected for review.
+The trauma registry is comprised of patients treated at the Karolinska
+University Hospital in Solna, which treats all major trauma in the
+greater metropolitan area of Stockholm. We included patients registered
+between 2014 and 2023. The trauma care quality database is a subset of
+the trauma registry and includes patients selected for review.
 
@@
 
-Inclusion in the trauma registry requires either admission through trauma team activation or presenting with an Injury Severity Score (ISS) greater than nine after admission to the Karolinska University Hopsital.
-Each trauma patient is then included in a morbidity and mortality review process, which involves both individual case evaluations by specialized nurses and audit filters.
-Patients identified with a high potential for OFIs are discussed at multidisciplinary conferences.
-The identified OFIs are then categorized into broad and detailed categories.
-The presence or absence of OFIs is determined by consensus among all participants and is documented in the trauma care quality database.
+Inclusion in the trauma registry requires either admission through
+trauma team activation or presenting with an Injury Severity Score (ISS)
+greater than nine after admission to the Karolinska University Hospital.
+Each trauma patient is then included in a morbidity and mortality review
+process, which involves both individual case evaluations by specialized
+nurses and audit filters. Patients identified with a high potential for
+OFIs are discussed at multidisciplinary conferences. An OFI can be
+various types of preventable events, and can be categorized as: clinical
+judgment error, inadequate resources, delay in treatment, missed injury,
+inadequate protocols, preventable death and other errors. The presence
+or absence of OFIs is determined by consensus among all participants and
+is documented in the trauma care quality database.
 
-We included all patients in the trauma registry and Trauma care quality database. We excluded patients younger than 15 and/or were dead on arrival.
+All patients in the trauma registry and trauma care quality database
+were included. Patients younger than 15 and/or were dead on arrival were
+excluded.
 
@@
 
-The outcome was defined as the presence of at least one OFI, determined by the multidisciplinary M&M conference in the trauma care quality registry.
-An OFI can be various types of preventable events, and can be categorized as: clinical judgment error, inadequate resources, delay in treatment, missed injury, inadequate protocols, preventable death and other errors.
+The outcome was defined as the presence of at least one OFI, determined
+by the multidisciplinary M&M conference in the trauma care quality
+registry. An OFI can be various types of preventable events, and can be
+categorized as: clinical judgment error, inadequate resources, delay in
+treatment, missed injury, inadequate protocols, preventable death and
+other errors.
 
-The patients will be classified to different degrees of shock in parallel and separately, once with systolic blood pressure (SBP) and once with base excess (BE), regardless of their cause of shock.
-Based on these two parameters, the patients will be classified into four classes roughly based on the ATLS trauma shock classifications (See Table \@ref(tab:atls) [@noauthor_atls_2018]).
-
-Table: (\#tab:atls) ATLS classification for haemorrhagic shock, adapted from [@noauthor_atls_2018].
+The patients will be classified to different degrees of shock in
+parallel and separately, once with systolic blood pressure (SBP) and
+once with base excess (BE), regardless of their cause of shock. Based on
+these two parameters, the patients will be classified into four classes
+roughly based on the ATLS trauma shock classifications (See Table
+\@ref(tab:atls) [@noauthor_atls_2018]).
 
 | **Parameter** | **Class I** | **Class II** (Mild) | **Class III** (Moderate) | **Class IV** (Severe) |
-|-----------------------------------|-------------|---------------------|---------------------------|-------------------------------|
-| Approximate blood loss            | <15%        | 15–30%              | 31–40%                    | >40%                          |
+|----------------|--------------|--------------|--------------|--------------|
+| Approximate blood loss | \<15% | 15–30% | 31–40% | \>40% |
 | Heart rate | ↔ | ↔/↑ | ↑ | ↑/↑↑ |
@@
 
-We included five other metrics to be adjusted for, due to potential for confounding. Pre-injury ASA and gender were categorical, meanwhile age, ISS and INR were kept continuous.
+: (#tab:atls) ATLS classification for haemorrhagic shock, adapted from
+[@noauthor_atls_2018].
 
-## Bias
+We included five other metrics to be adjusted for, due to potential for
+confounding. Pre-injury ASA and gender were categorical, meanwhile age,
+ISS and INR were kept continuous.
 
@@
 
-All available data in trauma registry and trauma care quality database will be included.
+All available data in trauma registry and trauma care quality database
+will be included.
 
@@
 
-Table \@ref(tab:mod-atls) shows how we defined shock according to BE and SBP.
-The BE parameter was divided into four classes with these cutoffs: Class I (above -2), Class II (-2 to -6), Class III (-6 to -10), and Class IV (below -10).
-Similarly, SBP was divided into four classes: Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90 mmHg), and Class IV (below 90 mmHg).
-This classification follows the tenth edition ATLS hemorrhage classification [@noauthor_atls_2018], which does not specify exact values for any parameter other than BE.
-To solve this, we assigned numerical SBP values to each class based on findings from Eastridge et al.
-and Oyetunji et al., who redefined hypotension as 110 mmHg respective 90 mmHg dependent on age[@eastridge_hypotension_2007; @oyetunji_redefining_2011].
-Therefore, we used 110 mmHg as the lower limit for (Class I: no shock) - (ATLS: normal SBP) and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear hypotension), class 2 and 3 divided equally in between.
-This also aligns with the classification done by Mutschler et al.[@mutschler_critical_2013].
+Table \@ref(tab:mod-atls) shows how we defined shock according to BE and
+SBP. The BE parameter was divided into four classes with these cutoffs:
+Class I (above -2), Class II (-2 to -6), Class III (-6 to -10), and
+Class IV (below -10). Similarly, SBP was divided into four classes:
+Class I (above 110 mmHg), Class II (109-100 mmHg), Class III (99-90
+mmHg), and Class IV (below 90 mmHg). This classification follows the
+tenth edition ATLS hemorrhage classification [@noauthor_atls_2018],
+which does not specify exact values for any parameter other than BE. To
+solve this, we assigned numerical SBP values to each class based on
+findings from Eastridge et al. and Oyetunji et al., who redefined
+hypotension as 110 mmHg respective 90 mmHg dependent on
+age[@eastridge_hypotension_2007; @oyetunji_redefining_2011]. Therefore,
+we used 110 mmHg as the lower limit for (Class I: no shock) - (ATLS:
+normal SBP) and below 90 mmHg for (Class IV: clear shock) - (ATLS: clear
+hypotension), class 2 and 3 divided equally in between. This also aligns
+with the classification done by Mutschler et
+al.[@mutschler_critical_2013].
 
-Table: (\#tab:mod-atls) ATLS tenth edition - Systolic blood pressure and Base excess (BD).
-
-| **Parameter - original ATLS 10th edition**| **Class I** | **Class II** (Mild)  | **Class III** (Moderate)  | **Class IV** (Severe) |
-|-----------------------------------|-------------|----------------------|---------------------------|-------------------------------|
+| **Parameter - original ATLS 10th edition** | **Class I** | **Class II** (Mild) | **Class III** (Moderate) | **Class IV** (Severe) |
+|----------------|--------------|--------------|--------------|--------------|
 | Base deficit | 0 to –2 mEq/L | –2 to –6 mEq/L | –6 to –10 mEq/L | –10 mEq/L or less |
 | Systolic Blood pressure | ↔ | ↔ | ↔/↓ | ↓ |
-| **Parameter - numerical approximated SBP**                                                                                         |
-| Systolic Blood pressure           | >110 mmhg   | 109-100 mmhg         | 99-90 mmhg                | <90  mmhg                     |
+| **Parameter - numerical approximated SBP** |  |  |  |  |
+| Systolic Blood pressure | \>110 mmhg | 109-100 mmhg | 99-90 mmhg | \<90 mmhg |
 
-Please note that ATLS classification defines class I for the BE parameter as (0 to -2) however, we choose to follow its original source[@mutschler_renaissance_2013], which defines class I as base deficit (Inverted BE) ≤ 2, with no lower limit (i.e.
-an upper limit for BE).
+: (#tab:mod-atls) ATLS tenth edition - Systolic blood pressure and Base
+excess (BD).
+
+Please note that ATLS classification defines class I for the BE
+parameter as (0 to -2) however, we choose to follow its original
+source[@mutschler_renaissance_2013], which defines class I as base
+deficit (Inverted BE) ≤ 2, with no lower limit (i.e. an upper limit for
+BE).
 
@@
 
-The statistical analysis will be performed using R, a programming language and environment for statistical computing.
-We will present the types of OFI as percentage distributions and then visualize them in a bar chart.
-Unadjusted logistic regression will be used to determine the association between the OFI, and degree of shock defined separately by classifying BE and SBP roughly according to ATLS classification of hemorrhagic shock.
-Adjusted logistic regression will then incorporate other patient factors such as age, sex, preinjury ASA, INR, and ISS.
-It will be presented as odds ratios (OR) between the presence of OFI, and shock classes.
-The OR will be determined with 95% confidence intervals, and a significance level of 5% will be used.
-All statistical analysis will first be done on synthetic data and later implemented on the data collected from the trauma registry and the trauma care quality database to ensure objectivity.
-Missing data will be addressed by listwise deletion.
+The statistical analysis was performed using R, a programming language
+and environment for statistical computing. OFI subcategories were
+presented as percentage distributions and visualized in a bar chart.
+Bivariable logistic regression was used to determine the unadjusted
+association between the OFI, and degree of shock defined by BE and SBP
+classifications. Multivariable logistic regression incorporated and
+adjusted for other patient factors such as age, sex, preinjury ASA, INR,
+and ISS. It was presented as odds ratios (OR) between the presence of
+OFI, and shock classes. The OR was determined with 95% confidence
+intervals, and a significance level of 5% was used. All statistical
+analysis was first done on synthetic data and then later implemented on
+the data collected from the trauma registry and the trauma care quality
+database to ensure objectivity. Missing data was addressed by listwise
+deletion.
 
 # Results
+
 ```{r main, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
@@
 ## Participants
-There was a total of 14022 patients in the trauma registry.
-After excluding the patients younger than 15 and/or were dead on arrival, there were 12153 patients remaining.
-Out of those, 7152 patients had been reviewed for the presence of OFI.
-A total of 2233 patients were excluded due to missing data, resulting in 4919 patients for the final analysis.
-The selection process can also be seen in Figure 1.
-The variable with the most missing data is BE which lacks the data for 1721 patients.
-Detailed data regarding missing data is displayed in a table in appendix 1.
 
-#ändra till flowchart
+There was a total of 14022 patients in the trauma registry. After
+excluding the patients younger than 15 and/or were dead on arrival,
+there were 12153 patients remaining. Out of those, 7152 patients had
+been reviewed for the presence of OFI. A total of 2233 patients were
+excluded due to missing data, resulting in 4919 patients for the final
+analysis. The selection process is shown in Figure 1. The variable with
+the most missing data is BE, for which data was missing in 1721
+patients. Detailed data regarding missing data is displayed in the
+supplementary material.
 
-#slut
+Figure 1. Flowchart of selection of study population.
 
 ## Descriptive data
-Table \@ref(tab:sample-characteristics-reg) describes the characteristics of the study population.
-The study demographics median age was `r age` and the most common gender was male, `r male`.
-Clinically, the median ISS was 5 which is considered minor injuries, median INR was 1.00, and the most common Preinjury ASA class was 1, at 62%.
+
+Descriptive data is shown in Table 3. The median age was 39 years (25,
+56), and 3506 (71%) of the individuals were males. Clinically, the
+median ISS was 5, which is considered minor injuries, and the median INR
+was 1.00, and the most common Preinjury ASA class was 1, at 62%.
+
+Overall, 3989 (81%) of the patients were classified as having no shock
+(class 1) according to the BE classification, the sum of the remaining
+classes accounted for 930 (19%) patients. The distribution of the
+classes was decreasing in numbers with increased severity, with class 2:
+644 (13%), class 3: 170 (3.5%), and class 4: 116 (2.4%).
+
+According to SBP, 4455 (91%) patients had no shock (class 1), while the
+remaining classes with patients in shock accounted for 464 (9%) of all
+the patients. Among the classes defined with shock, class 2 was the
+biggest at 239 (4.9%), followed by class 4 at 127 (2.6%), and lastly
+class 3 with 99 (2.0%).
 
@@
 ## Outcome data
-Out of the 4919 patients reviewed without missing data, 281 (5.7%) patients had at least one positive OFI.
 
-The most common OFI type was clinical judgement error, numbering 103 (37%) patients.
-The other three major OFI types were: inadequate resources for 61 (22%) patients, delay in treatment for 49 (17%) patients and missed diagnosis for 48 (17%) patients.
-The least common OFI was “other errors”, for 5 (1.8%) patients.
+Out of the 4919 patients reviewed without missing data, 281 (5.7%)
+patients had at least one OFI. The most common OFI type was clinical
+judgment error, numbering 103 (37%) patients. The other three major OFI
+types were inadequate resources for 61 (22%) patients, delay in
+treatment for 49 (17%) patients, and missed diagnosis for 48 (17%)
+patients. The least common OFI was “other errors”, which was present in
+5 (1.8%) patients.
 
-When classified, the most common OFI type in class 1 (no shock) was clinical judgement error both in the BE and SBP groups, while the other OFI types also followed a similar distribution between the groups.
-The largest OFI type continues to be clinical judgment error for BE group in class 2, 3 and 4, at 34.6%; 31.2%; and 42.9% in this order.
-On other hand, the largest OFI type in SBP class 2 was delay in treatment at 50%.
-The SBP class 3 was evenly distributed between clinical judgement error and missed diagnosis at 40% each.
-Lastly, class 4 SBP’s most common OFI type was clinical judgment error 46.7%.
+The most common OFI in patients without chock (class 1) was clinical
+judgment error both in the BE and SBP groups, while the other OFI types
+also followed a similar distribution between the groups. Clinical
+judgment error was the most common OFI in the BE group class 2, 3 and 4,
+at 34.6%; 31.2%; and 42.9% respectively. On the other hand, the largest
+OFI type in SBP class 2 was delay in treatment at 50%. The frequencies
+of clinical judgement error and missed diagnosis were similar in the SBP
+class 3 group, at 40% each. Lastly, the most common OFI in class 4 of
+the SBP group was clinical judgment error, at 46.7%.
 
-Overall, 3989 (81%) of the patients were classified as having no shock (class 1) according to the BE classification, the sum of the remaining classes accounted for 930 (19%) patients.
-The distribution of the classes was decreasing in numbers with increased severity, with class 2: 644 (13%), class 3: 170 (3.5%), and class 4: 116 (2.4%).
-
-According to SBP, 4455 (91%) patients had no shock (class 1), while the remaining classes with patients in shock accounted for 464 (9%) of all the patients.
-Among the classes defined with shock, class 2 was the biggest at 239 (4.9%), followed by class 4 at 127 (2.6%), and lastly class 3 with 99 (2.0%).
-
--------
+------------------------------------------------------------------------
 
@@
 ## Main results
-The adjusted analyses showed only statistical significance when comparing BE class 4 to class 1, which had 63% lower odds for OFI (OR 0,37; 95% CI 0,14-0,83; p-value 0.026).
 
-The un-adjusted analysis showed when comparing to class 1 in BE group, class 2 (OR 1,61; 95% CI 1,17-2,20; p-value 0.003) and 3 (OR 1,91; 95% CI 1,08-3,16; p-value 0.018) were significant in increased risk for OFI.
-For the SBP group, comparison between class 1 and class 4 showed significant increased risk (OR 2,28; 95% CI 1,26-3,85; p-value 0.004).
+The unadjusted analysis of the association between shock class categorised by BE is shown in Table 6.
+This analysis showed that compared to class 1 (no shock), the odds of OFI in class 2 and 3 were increased by 61% (OR 1.61, 95% CI 1.17-2.20, p-value 0.003) and 91% (OR 1.91, 95% CI 1.08-3.16, p-value 0.018) respectively.
+Table 7 shows the association between shock classes categorised by SBP.
+Compared to class 1 (no shock), the odds of OFI in class 4 were increased by 128% (OR 2.28, 95% CI 1.26-3.85, p-value 0.004).
+
+The adjusted analysis was conducted in two steps.
+In the first step, we adjusted for age, gender, preinjury ASA score, and INR, but not for ISS.
+Compared to the unadjusted analysis, this initial adjustment led to minor changes in ORs.
+In the BE group, the OR for class 2 changed by approximately 1% (from 1.61 to 1.60), and for class 3, it changed by about 2% (from 1.91 to 1.93), as shown in Table 6.
+In the SBP group, the OR for class 4 changed by around 7% (from 2.28 to 2.21), as shown in Table 7.
+
+In the second step, ISS was added to the adjustment model to complete the fully adjusted model.
+For the BE group (Table 6), this led to a shift in statistical significance: instead of class 2 and 3, class 4 became statistically significant.
+In this fully adjusted model, class 4 was associated with 63% lower odds of an opportunity for improvement (OR 0.37, 95% CI 0.14–0.83, p = 0.026).
+In contrast, for the SBP group (Table 7), no class remained statistically significant in the fully adjusted model.
 
 Table 6 adjusted log BE
+
 ```{r masterBE, echo=FALSE}
@@
 Table 7 adjusted log SBP
+
 ```{r masterSBP, echo=FALSE}
@@
 # Discussion
+
 ```{=html}
@@
 ```
-The purpose of this study was to assess the association between OFI and adult trauma patients arriving in shock, focusing on the types of OFI and severity of shock.
 
-##Key results
+## Key results
+
 We found that the most common type of OFI is clinical judgment error across almost all shock severities.
-Out of all shock severities, there is only a statistically significant association between severe shock with a BE of < -10 and OFI, which shows 63% reduced odds of OFI when compared against those with no shock.
+The unadjusted models suggested a positive association between moderate shock and OFI; adjustment reversed this, with severe shock showing 63% reduced odds.
 
-The unadjusted analyses demonstrated a general positive association between shock and OFI. However, this association inverted in the adjusted models, with BE < -10 showing reduced odds for OFI.
+## Limitations
 
-##Interpretation
-Patients with BE < -10, indicating severe shock, were found to have lower odds of OFI compared to other groups, and two hypotheses may explain this finding.
+The study design is limited by the lack of data on heart rate, FAST, detailed types of injury, and clinical chemistry.
+This resulted in choosing a proprietary simplified ATLS framework instead of categorizations such as TASH, TBSS, and ABC with better sensitivity and specificity than the ATLS framework [@yucel_trauma_2006;@ogura_modified_2016;@hanlin_prehospital_2024].
+Another limitation is that missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias due to missing BE values.
+Also worth noting is that the sample size for most of the shock classes (class 2 to 4) are roughly around 100 to 200 patients, whereas those with positive OFI are as low as five patients.
+This may have resulted in higher randomness and inflated effect sizes.
 
-First, patients in severe shock should be more easily recognized as critically ill upon arrival, triggering heightened clinical attention and prioritization in triage.
-This results in more structured management and often receives priority in resource allocation, which may minimize variability in care and reduce the likelihood of errors.
+## Interpretation
 
-Another explanation is the potential influence of “reversed” survival bias.
-Patients with BE < -10 have high mortality rates (25), and for these patients who die, the care process is interrupted prematurely.
-This may induce generally fewer procedures or shorter care time for patients with deadly outcomes, which means less chance for errors, and therefore OFIs.
-This would result in reduced amount of OFIs in groups with higher mortality, which in this case would be class 4.
+The lower odds of OFI for patients with BE < -10 may be explained by two theories.
+One is that more severe injuries may receive heightened attention and more protocol-driven responses.
+This would in turn, result in reduced risk of deviation from optimal care (and thus OFIs).
+This theory is also partly supported by the observation that the OFIs “inadequate protocols” and “missed diagnosis” generally occur with lower frequencies with higher shock degrees, regardless of whether classed with BE or SBP.
+Do note that the low frequencies of OFI in the severe shock classes, limit this theory with higher chance of randomness.
 
-###Interpretation of adjusted model after inclusion of ISS
-When looking at the distribution of the significant and insignificant results of the BE group (table 6), it can be observed that the significance reverses for the classes when adding ISS.
-One explanation is that the unadjusted positive correlation may be driven by a shared variability with ISS, as we can see significant positive association between ISS and OFI in our results.
-This positive association between OFI and ISS have also been shown in a study by Husssein et al (26).
-It is important to note that Hussein’s study was also conducted at Karolinska Solna, but during a different time frame (2017-2021) and with other inclusion and exclusion criteria.
+The second theory is survival bias, where the reduced odds of OFI may be explained by that this study only incorporated the surviving patients.
+This may result in only the most optimally cared patients surviving.
+Patients with BE < -10 also have high mortality rates [@mutschler_renaissance_2013].
+Hypothetically, this would result in the severe classes of shock, with more dying patients with OFIs being eliminated from the data, due to death as the outcome.
 
-Another hypothetical explanation is multicollinearity, as both BE and ISS assess injury severity.
-Theoretically anatomical injury, bleeding and physiological response should be closely interconnected.
-A correlation between ISS and BD (inversed BE) or SBP was also shown in the two studies by Mutscher et al.
-(11,25).
-Therefore, adjusting for each other may explain the inversion of OR.
-However, it may also result in results that’s difficult to interpret or lack any meaning in a clinical context.
-This may also explain why class 2 and 3 are not significant but class 4 are in the adjusted analysis, as seen in study by Hussein et al (26), ISS does not correlate linearly to OFI, but we included ISS linearly in the model.
+The reversal of OR and distribution of statistical significance may be attributed to multicollinearity, as both BE and ISS assess injury severity.
+Theoretically, anatomical injury, bleeding, and physiological response should be closely interconnected.
+A correlation between ISS and BD (inverted BE) or SBP was also shown in the two studies by Mutscher et al.
+[@mutschler_critical_2013;@mutschler_renaissance_2013].
+Therefore, adjusting for each other may generate results that are difficult to interpret.
+This may also explain why classes 2 and 3 are no longer statistically significant, whereas class 4 remains significant.
 
-However, when adding ISS to the SBP group, the significance of class 4 disappears, but the significance of classes 2 and 3 does not change, like in the case for BE group.
-The same can be said about the OR, it does change but not nearly as major as in the BE class.
-Especially for class 2, as the OR merely changes from 1.06 to 1.11 and then to 0.96.
-Class 3 OR changes more, from 0.92 to 0.90 to 0.47; but the OR difference is still < 0.5, whereof for the BE group and SBP class 4 the difference is always > 0.5.
-The reason may be due to the cutoffs of the classification system, as the SBP classification does not follow the same range of shock as the BE classification.
-The results of the study that BE classification was originating from (25), shows that BE class 1 had a mean SBP of 132.6 mmHg and class IV had 94.8 mmHg.
-As an SBP of 132.6 mmHg is at the higher range of normotensive, we were therefore not able to use it as an upper limit, but the 94.8 mmHg is close to our approximation of 90 mmHg for class 4.
-The SBP classification is therefore only somewhat connected to the BE classification at the lower limit e.g.
-class 4.
-Also worth noting is that the divider between our SBP class 2 and 3, are arbitrary and made to connect with the ATLS system to make up a four-class system, as the 90 mmHg and 110 mmHg are from Eastridge et al (23) and Oyetunji et al (24) which only defined 110 mmHg and 90 mmHg and nothing in between.
-This means that this divider may have made the classes too small and no longer scientific.
+One other factor which may have contributed to this pattern of results, especially the distribution of significant classes and not as much the reversal after adjustment with ISS, was that ISS was included as a continuous (linear) variable in the model.
+As shown by Albaaj et al.
+[@albaaj_patient_2023], increases in ISS do not correspond proportionally to increases in the likelihood of OFI.
+Therefore, its relationship with opportunities for improvement (OFI) is likely not linear, which results in different statistical significances between the classes.
 
-###Comparison to literature
-Previous studies (7,27) which studied cases with a deathly outcome, showed that the most common OFI type was delay in treatment, while the second most common is clinical judgment error (7), or errors in management (27).
-In comparison, our results showed the most common OFI is clinical judgment error (36%), while inadequate resources were second (22%), and delay in treatment comes third (17%), for the whole study population.
-These differences may be due to the nature of the studied population, where delay in treatment may be more deadly, therefore it takes up a smaller part of our results, while clinical judgment error takes up a larger part.
-No studies were found to address this hypothesis of survivorship bias for OFI types.
+### Comparison to literature
 
-For patients defined as in shock (Class 2-4), the most common OFI was still clinical judgment error (BE 34.7%; SBP 35.3%).
-This result is similar when compared to D O’Reilly’s study (28), in which the most common type of opportunities for performance improvements (OPI) was one involved with decision-making at 63 out of 150 OPIs.
-We did not find any publication addressing shock severity and OFI.
-We have therefore looked at studies addressing preventable negative care outcomes and assessed severity by other factors but only found one that investigated predicting factors and preventable death.
-This study (29) showed that there was a OR correlation of 14.1 between preventable death and SBP ≥ 80mmHg, when comparing to SBP < 80 mmHg.
-This differs from our SBP findings, as our adjusted results are not statistically significant, and the significant unadjusted results don’t have that extreme of OR and are positively correlated with severe and not mild shock.
-These differences are not strange, but rather understandable, as the studies are conducted at different hospitals, populations, outcomes (surviving vs death) and different classifications (the lowest of our 4 SBP classes is 90mmHg, while they only got 2 classes that separate at 80mmHg).
-However, if reasoned from a shock standpoint where both SBP and BE are part of, and compare this result with our BE results, they show a similar trend.
-Our BE results showed reduced odds of OFI when comparing the most severe and the no shock class.
-This implies that if changing the reference point to the most severe shock class, the result would be increased odds of OFI when compared to the no shock class.
-Therefore, our results show similar trend to their study (29) result, that showed lesser shock (higher SBP) had increased odds ratio of preventable death.
+A previous study by O’Reilly et al [@oreilly_opportunities_2013], which looked into opportunities for performance improvements (OPI) in 7511 cases, whereof 127 patients whose death was affected by hemorrhage, found that the most common OPI was related to decision making (63 out of 150 OPIs).
+This aligns with our findings that patients in shock (Class 2-4) were most affected by the OFI, “clinical judgment error” (BE 34.7%; SBP 35.3%).
 
-One surprising result is that there is no significance for the SBP group in our adjusted analysis.
-When comparing to the results from Hussein et al.
-(26), the SBP results conform, as it also showed significance for the unadjusted results below 90 mmHg and no significance when adjusted with other factors including ISS.
-Due to a lack of research regarding shock/injury severity and OFI/preventable death, it is difficult to determine whether the results are reasonable or not.
-It is therefore also difficult to determine exactly why some of them are significant and others not, leaving us with reasoning and hypothetical explanations.
+No studies were found that specifically examined OFIs in patients in hemorrhagic shock, but studies on OFIs in general trauma populations show results in a similar theme.
+A study by T.P.
+Saltzherr [@saltzherr_preventability_2011] from a Dutch trauma center reported that “delay in treatment” or “error in management” was the most common error type.
+Similarly, Martin et al.
+[@martin_analysis_2009] found that delays to surgery were significant contributors to adverse outcomes, for a military hospital in Baghdad.
+Sanddal et al [@sanddal_analysis_2011] identified the main issues of care for a trauma system in development in Utah as chest injury diagnostic and management, airway management, and fluid resuscitation.
+Teixeira et al.
+[@teixeira_preventable_2007], analyzing preventable deaths in a mature trauma center, also confirmed the impact of delayed treatment and poor clinical decision-making.
 
-##Strengths and limitations
-Strength of this study includes the relatively large cohort of 4919 patients spread over 9 continuous years.
-From a methodical standpoint, even though an established shock/hemorrhage classification system was not used, two clinical parameters were used in parallel.
-This includes two perspectives, one physiological and one chemical, to best describe shock and at the same time still using the original parameters the clinicians meet on a day-to-day basis.
-
-One of the limitations is the lack of data to design the study around a modern shock/hemorrhage classification which uses a combination of base parameters.
-Due to a lack of data on heart rate, FAST, detailed types of injury, and clinical chemistry, we were not able to use categorizations such as TASH, TBSS, and ABC with better sensitivity and specificity than the ATLS framework (30–32).
-The ATLS shock classification framework would, in turn, would also be better studied and evidence-based, than the proprietary classification we simplified and built our classification from.
-
-The choice to not include the ATLS hemorrhage classification was due to four reasons.
-The first one is ATLS tenth edition lacks numerical values for vital parameters.
-The second one is that base deficit alone is superior to vital parameters as a predictive parameter for mortality (33).
-The third one is due to the nature of having multiple parameters, which makes it difficult to categorize a patient with vital parameters in two or more classes at the same time.
-The fourth one is difficulties in separating lower GCS due to traumatic brain injury from shock as a cause.
-
-Missing data were handled by listwise deletion, which reduced the sample size by 2233 and may also introduce bias, as the most common missing category was BE which may mean that BE was not routinely measured and result in inconsistency for certain patient groups, which risks selection bias.
-Also worth noting is that the sample size for most of the shock classes (class 2 to 4) are roughly around 100 to 200 patients, whereof those with positive OFI are as low as 5 patients.
-This may result in over/underestimation of the OR, as there is higher chance of randomness in smaller sample sizes, which results in broader confidence intervals which may result in more “extreme” average OR.
-
-##Clinical/Practical applications - Generalisability?
-Clinically the unadjusted are more insightful, as the variables are never separated and instead combined into complex cases.
-Therefore, from a BE standpoint when a patient’s BE is between -2 to -10 the odds of OFI are between 61% to 91% higher than those with BE above -2.
-Suggesting patients with a BE between these levels may need extra attention.
-From a SBP perspective, when SBP dips below 90 mmHg odds for OFI increases a massive 128%, compared with those with SBP above 110 mmHg.
-This supports the idea (24) of using 90 mmHg as a clinical limit for defining hypotension.
-
-###Equity
-It can be observed that the male gender is more prevalent in our results than females (71% vs 29%), which is like other studies (22,27,28).
-However, when comparing the event rate of OFI among the patients reviewed for OFI, the difference between the genders becomes minimal at 5.1% for female, respective 6% for male.
-At the same time, when fully adjusted, gender is not significant (p = 0.4), this implies that even though gender may not be proportionate in trauma cases, there seems to be no major difference between the genders for trauma care quality.
-This may reflect that the standardized protocols for trauma care minimize gender-based differences.
-When comparing to other studies (27,28), it seems like the percentage potentially preventable death and OPI have a higher amount for males, however, to be noted their percentage is based on ratio between genders and not event risk for each gender.
-Socioeconomical factors were not incorporated into this study, nonetheless an important factor in emergency medicine, when resources are scarcer.
-Especially when “inadequate resources” is the third most common OFI type, even for a major hospital in the capital of Sweden.
-This implies the results of this study cannot be generalized globally, especially not places where standardized trauma care protocols are not incorporated and/or lack adequate resources.
-
-###Future studies
-The next step for future studies is to validate the findings and improve the methods from this study.
-As previously discussed, the association between anatomical classification ISS and physiological parameters BE and SBP, and how they correlate to OFI are complex and not necessarily clinically relevant.
-One way to avoid this problem is using a validated classification for hemorrhagic shock with both anatomical and physiological components, like ABC, TASH, and so on.
-
-One other problem when interpreting the results is the “reversed survivorship bias”.
-Due to the design of trauma care quality database, when and where (level of care) the OFI occurred during the length of stay is not recorded.
-A prospective study can clarify whether shock severity directly affects OFI or if length of stay is a confounding factor.
-By observing when and where OFIs most commonly occur, it can determine whether the hypothesis of length of stay is correct.
-This also aids in understanding the reasons behind the nonlinear relationship between ISS, BE or SBP and OFI, as there are currently only theories but no studies.
+In comparison, our study found that “clinical judgment error” was the most common OFI for all patients and patients in shock.
+While the relative ranking may vary, whether delay in treatment or clinical judgment error is most common, these two categories consistently emerged as the most frequent across studies.
+This also suggests that the types of OFIs that were common in the general trauma population were similarly frequent in patients in shock.
+Moreover, this pattern appears consistent regardless of the studied population, whether the focus was on preventable deaths or on patients who survived their injuries.
 
 # Conclusion
-<!-- Give a short conclusion, corresponding to the aim -->
 
-The most common type of OFI for shock patients is clinical judgement error.
-Association between shock and OFI doesn’t seem to be linear, but rather the moderate shock has higher odds of OFI compared to the mild and severe shocks.
-The lack of significance in SBP classification highlights potential framework limitations compared to BE.
-These findings emphasize the importance of accounting for ISS and using validated shock classifications in future research.
+Clinical judgment error was the most common OFI in shock patients.
+The association between shock and OFI does not seem to be linear, but rather the moderate shock has higher odds of OFI compared to the mild and severe shocks.
+These findings emphasize the importance of improved protocols for initial management of non-critical patients to minimise decision error.
 
-#Contributions
+# Contributions
 Data collection from the trauma registry, ethics application and study designed by Martin Gerdin Wärnberg.
 Carl sorted, programmed and analyzed the data, and then wrote the thesis with support from Martin Gerdin Wärnberg, Johanna Berg, Jonatan Attergrim and Kelvin Szolnoky.
-Technical tools used include Chat-GPT and Google translate.
-Both were used in the same way, as means to translate more adequate vocabulary for scientific writing between Swedish and English.
-The overall formulations, structure and content were written by me (Carl), and which of the translations done by the two tools, that were to be used to ensure the content, were also done by me (Carl).
-
-##Acknowledgements
-I want to here again thank my supervisor and co-supervisors for supporting, replying to questions and reading and giving me constructive feedback.
-I also want to thank my coordinator Jan Jakobsson for a lot of constructive feedback and guiding my understanding of the practicalities and extent for this thesis.
-My fellow medical student Anton Fedchenko also supported me with constructive feedback.
-Lastly, I also want to thank my friend Gustav Nilsson-Gisleskog, a master student in computer science at Lund University for helping me solve conflicting push pull requests on git-hub.
 

### 2025-12-26 manuscript.Rmd
ändrade till bara kön utan (M/F)

@@
 output:
-  #word_document: default
+  word_document: default
   #bookdown::pdf_document2:
     #latex_engine: xelatex
-  html_document: default
+  #html_document: default
 bibliography: bibliography.bib

### 2025-12-26 main.R

@@ -274,7 +274,7 @@ var_label(reg.sample$ed_be_art_numeric) <- "Base Excess (BE)"
 var_label(reg.sample$BE_class) <- "Shock classification - BE"
 var_label(reg.sample$pt_asa_preinjury) <- "Pre-injury ASA"
 var_label(reg.sample$ISS) <- "Injury Severity Score"
-var_label(reg.sample$pt_Gender) <- "Gender (M/F)"
+var_label(reg.sample$pt_Gender) <- "Gender"
 var_label(reg.sample$ed_sbp_value) <- "Systolic blood pressure (mmHg)"
 var_label(reg.sample$ed_inr_numeric) <- "INR"
 var_label(reg.sample$V4SBP_class) <- "Shock classification - SBP"

### 2025-12-26 main.R
Ändrat ordning på tabellerna

@@ -449,12 +449,12 @@ sample.characteristics.table_ISS_SBP <- tbl_summary(reg.sample,
 
 # Step reg + event
 master_combined_table_stepBE <- tbl_merge(
-  tbls = list(log_regBE_sample.characteristics.table, BEstep1, log_regBE_sample.characteristics.table_unadjusted),
-  tab_spanner = c("**Fully adjusted**","**without ISS**","**Unadjusted**")
+  tbls = list(log_regBE_sample.characteristics.table_unadjusted, BEstep1, log_regBE_sample.characteristics.table),
+  tab_spanner = c("**Unadjusted**","**without ISS**","**Fully adjusted**")
 ) 
 
 master_combined_table_stepSBP <- tbl_merge(
-  tbls = list(log_regSBP_sample.characteristics.table, SBPstep1, log_regSBP_sample.characteristics.table_unadjusted),
-  tab_spanner = c("**Fully adjusted**","**without ISS**","**Unadjusted**")
+  tbls = list(log_regSBP_sample.characteristics.table_unadjusted, SBPstep1, log_regSBP_sample.characteristics.table),
+  tab_spanner = c("**Unadjusted**","**without ISS**","**Fully adjusted**")
 ) 
 

### 2025-12-26 main.R
Tar bort (M/F)

@@ -302,7 +302,7 @@ log_regBE_sample.characteristics.table <-
     label = list(
       BE_class ~ "Shock classification - BE",
       pt_age_yrs ~ "Age (Years)",
-      pt_Gender ~ "Gender (M/F)",
+      pt_Gender ~ "Gender",
       pt_asa_preinjury ~ "Pre-injury ASA",
       ed_inr_numeric ~ "INR",
       ISS ~ "Injury Severity Score"
@@ -335,7 +335,7 @@ log_regSBP_sample.characteristics.table <-
     exponentiate = TRUE,
     label = list(
       pt_age_yrs ~ "Age (Years)",
-      pt_Gender ~ "Gender (M/F)",
+      pt_Gender ~ "Gender",
       pt_asa_preinjury ~ "Pre-injury ASA",
       ed_inr_numeric ~ "INR",
       ISS ~ "Injury Severity Score",

### 2025-12-26 main.R
tillbaka för att jag tog bort innan, men det ändrade ingenting i knitt versionen så vill inte råka förstöra något aannat

@@ -302,7 +302,7 @@ log_regBE_sample.characteristics.table <-
     label = list(
       BE_class ~ "Shock classification - BE",
       pt_age_yrs ~ "Age (Years)",
-      pt_Gender ~ "Gender",
+      pt_Gender ~ "Gender (M/F)",
       pt_asa_preinjury ~ "Pre-injury ASA",
       ed_inr_numeric ~ "INR",
       ISS ~ "Injury Severity Score"
@@ -335,7 +335,7 @@ log_regSBP_sample.characteristics.table <-
     exponentiate = TRUE,
     label = list(
       pt_age_yrs ~ "Age (Years)",
-      pt_Gender ~ "Gender",
+      pt_Gender ~ "Gender (M/F)",
       pt_asa_preinjury ~ "Pre-injury ASA",
       ed_inr_numeric ~ "INR",
       ISS ~ "Injury Severity Score",
